# Individual Tree Accuracy Evaluation

In this section, we'll mask the point cloud detected trees using the slash pile detections to get our final UAS point cloud-derived forest inventory tree list. Then we'll bring in field-measured data to evaluate our tree detection accuracy. This accuracy evaluation will yield detection accuracy metric including precision, recall (i.e. detection rate), and F-score (overall accuracy or balanced accuracy). Precision is a measure of the proportion of trees that estimated by the UAS method that were actually trees. Recall is a measure of the proportion of field-measured trees that the UAS method correctly identified. The F-score balances precision and recall to create an overall accuracy measure. All of these metrics range between 0 and 1 with values close to one indicating higher tree detection accuracies. In open canopy ponderosa pine forests, tree detection rates generally range from 65-85% (Creasey et al. 2021; Swayze and Tinkham 2023; Tinkham and Woolsey 2024) with higher proportions of larger trees (e.g. 75-95%) being detected compared to smaller tree detection rates (55-75%). As far as I know there has not been a detailed accuracy evaluation of tree detection methods from aerial point cloud data in Pinyon-Juniper forests.

```{r, warning=FALSE, message=FALSE, echo=FALSE, include=FALSE}
remove(list = ls()[grep("_temp",ls())])
remove(treetops_sf)
gc()
```

## Remove Slash Piles from Predictions

read back in the tree point data and attach to treatment units

```{r}
# treetops_fnm
treetops_sf <- sf::st_read(dsn = treetops_fnm, quiet = TRUE)
# attach to treatment unit
treetops_sf <- treetops_sf %>% 
  dplyr::inner_join(
    treetops_sf %>% 
      sf::st_transform(sf::st_crs(stand_boundary)) %>% 
      sf::st_intersection(
        stand_boundary %>%
          dplyr::select(unit,block,tretmnt,unit_area_ha) %>% 
          dplyr::rename(type = tretmnt) %>% 
          dplyr::mutate(
            type =
              tolower(type) %>% 
              stringr::str_squish()
          ) %>% 
          dplyr::rename_with(.fn = ~paste0("trtmnt_", .x, recycle0 = T))
      ) %>% 
      sf::st_drop_geometry() %>% 
      dplyr::select(treeID, tidyselect::starts_with("trtmnt_")) %>% 
      dplyr::group_by(treeID) %>% 
      dplyr::slice(1) %>% 
      dplyr::ungroup()
    , by = "treeID"
  ) %>% 
  dplyr::mutate(
    is_in_trtmnt_unit = ifelse(is.na(trtmnt_type) | trtmnt_type=="control",F,T)
    , dplyr::across(c(trtmnt_type,trtmnt_block,trtmnt_unit),ordered)
  )
# treetops_sf %>% dplyr::glimpse()
```

use our slash pile mask to remove detected "trees" that are likely not trees. if a tree top point falls within the boundary of a detected pile footprint, it will be removed

```{r}
# save nrow
nrow_temp <- nrow(treetops_sf)
treetops_sf <- 
  sf::st_filter(
    treetops_sf
    , piles_workflow_ans %>% 
      dplyr::filter(is_in_trtmnt_unit, inrange_th_votes>=5) %>% 
      sf::st_union()
    , .predicate = sf::st_disjoint
  )
```

what proportion of "trees" were removed?

```{r}
# nrow_temp
# nrow(treetops_sf)
scales::percent((nrow_temp-nrow(treetops_sf))/nrow_temp, accuracy = 0.1)
```

nice. now we'll keep only trees that are at least 2 m in height since we included trees as small as 1.37 m in our [tree detection](#trees)

```{r}
min_tree_ht_m <- 2
treetops_sf <- treetops_sf %>% dplyr::filter(tree_height_m>=min_tree_ht_m)
```

```{r, warning=FALSE, message=FALSE, echo=FALSE, include=FALSE}
remove(list = ls()[grep("_temp",ls())])
gc()
```

## Reference Field Data

Field-measured data is frequently cited as the gold standard for validating remote sensing methods, but it may be more appropriate to frame field data as the status-quo baseline against which new methodologies are compared. Field inventories carry well-documented challenges, including measurement error, subjective assessments, and data transcription errors ([Elzinga et al. 2005]( https://doi.org/10.1093/wjaf/20.2.134); [Westfall and Woodall 2007](https://doi.org/10.1016/j.foreco.2007.07.014); [Luoma et al. 2017](https://doi.org/10.3390/f8020038); [Wang et al. 2019](https://doi.org/10.1016/j.isprsjprs.2018.11.008); [Sparks et al. 2024](https://doi.org/10.1093/forsci/fxae015)). This means that a comparison of the remote sensing forest measurement method against field data evaluates alignment with current operational practices rather than absolute environmental reality. Here are the notes I got along with the field inventory data:

> Attached is a csv of the PJ data from the Dawson Creek site. I added the plot center locations and calculated locations of each measured tree from the original distance and azimuth information. Below is a list of notes (and unfortunately some weird things) about it.

> *	There are 65 rows (out of 1603) with any missing data. The missing parts are usually in a location-related column with several in a tree metric column.
> *	There are 43 DRC (diameter at root collar) observations with two separate values separated by a comma. 
> *	There are 8 circumstances where there are potential duplicate trees - where plot, distance, and azimuth are identical, but tree information is different. 
> *	In the metadata crown length is described as "length from first longest side of the adult tree canopy, in m" and crown width is described as "width from second longest side of the adult tree canopy, in m". However, crown width is frequently greater than the length value and after some looking the length value doesn't seem to equal height-cbh either.
> *	The plot column includes unit, treatment type, and plot ID whereas the tree_id column includes that information as well as a per-plot ID of each tree.

> The multiple DRC values are due to multiple stems splitting at ground level. I think we should treat these as one tree, as they're recorded in the data. To combine the two stem values into a single DRC, we could either take the larger of the two or sum them.

> The location duplication issue is due to two trees (or more likely stems from the same individual that split below ground) growing very close together. We should continue to treat them as separate trees, as shown in the data.

> Height, cbh, length and width are all in meters. DRC is in cm.

Chojnacky & Rogers ([1999](https://doi.org/10.1093/wjaf/14.1.14)) provide information on converting DRC to DBH if needed

let's get to it and load in the field-measured "truth" data

```{r}
validation_trees <-
  readr::read_csv("../data/pj_trees_george.csv", show_col_types = F, progress = F) %>% 
  dplyr::rename_with(tolower) %>% 
  dplyr::rename_with(stringr::str_squish) %>% 
  dplyr::rename_with(make.names) %>% 
  dplyr::rename_with(~stringr::str_replace_all(.x, "\\.{2,}", ".")) %>% 
  dplyr::rename_with(~stringr::str_remove(.x, "\\.$")) %>% 
  dplyr::rename_with(~stringr::str_replace_all(.x, "\\.", "_")) %>% 
  dplyr::mutate(
    dplyr::across(
      -c(plot,tree_id,species,unit,treatment_code,drc)
      , as.numeric
    )
    , dplyr::across(
      c(species,unit,treatment_code)
      , ~ stringr::str_trim(.x) %>% toupper()
    )
  ) %>% 
  # clean the DRC
  dplyr::mutate(
    drc = stringr::str_remove_all(drc, " ")
     # split by comma
    , drc_list = stringr::str_split(drc, ",") %>% 
      purrr::map(as.numeric)
    , drc_count = purrr::map_dbl(drc_list, length)
    , drc_max_cm = purrr::map_dbl(drc_list, max)
    , drc_sum_cm = purrr::map_dbl(drc_list, sum)
  ) %>% 
  # dplyr::filter(stringr::str_detect(drc,",")) %>% dplyr::select(tidyselect::starts_with("drc")) %>% View()
  dplyr::select(-drc_list) %>% 
  # rename
  dplyr::rename_with(
    .cols = c(height,cbh,crown_width,crown_length)
    , .fn = ~paste0(.x,"_m",recycle0 = T)
  ) %>% 
  dplyr::rename_with(
    .cols = c(drc)
    , .fn = ~paste0(.x,"_cm",recycle0 = T)
  ) %>% 
  dplyr::mutate(
    # Chojnacky & Rogers ([1999](https://doi.org/10.1093/wjaf/14.1.14))
    dbh_cm = dplyr::case_when(
      # the correct code for Juniperus scopulorum is JUSC2...
      species %in% c("JUOS","JUSC","JUSC2") ~ -6.8180 + 1.0222*drc_max_cm + 1.8879*ifelse(drc_count==1,1,0)
      , species == "PIED" ~ -6.8180 + 1.0222*drc_max_cm + 1.8879*ifelse(drc_count==1,1,0) + 1.8971 -0.03998*drc_max_cm
      , species == "QUGA" ~ -6.8180 + 1.0222*drc_max_cm + 1.8879*ifelse(drc_count==1,1,0) + 3.1100 -0.0689*drc_max_cm
      , T ~ NA
    )
  )
# huh?  
validation_trees %>% dplyr::glimpse()
# validation_trees$species %>% table()
# validation_trees %>% dplyr::count(plot,tree_id) %>% dplyr::filter(n!=1) %>% nrow()
```

let's quickly look at the field measurements

```{r, echo=T}
validation_trees %>% 
  dplyr::select(height_m, drc_max_cm, dbh_cm, cbh_m) %>% 
  summary() %>% 
  knitr::kable(format = "markdown")
```

it looks like the Chojnacky & Rogers ([1999](https://doi.org/10.1093/wjaf/14.1.14)) DRC-to-DBH conversion worked well. Those crown and CBH measurements are suspect.

we'll also use the horizontal crown length and width to evaluate the UAS prediction of canopy cover on a plot-level basis. this canopy cover evaluation will:

* average the field measured horizontal crown length and width to approximate a circular diameter
* use the diameter to calculate a 2D circular crown area for each field tree
* calculate the proportion of each plot covered by these assumed circular crowns
* for each plot, compare this proportion to the proportion of the plot predicted to be covered by crowns of UAS trees whose tree top point falls within the plot. requiring the tree top point to fall within the plot prior to calculating crown cover mimics the field technique of only measuring stems within a plot.

the caveats attached to this approach are that it is notoriously difficult to measure crown dimensions using field techniques (source 2011) and the simple circular shape assumption does not match real-world conditions of irregular crown morphologies...

let's check out the horizontal crown length and width dimensions measured in the field

```{r}
validation_trees %>% 
  dplyr::select(crown_length_m, crown_width_m) %>% 
  summary() %>% 
  knitr::kable(format = "markdown")
```

yeah, some of those field-measured crown dimensions will need to be fixed...

ok, let's verify those notes we got with the data and look for missing coordinates

```{r}
validation_trees %>% 
  dplyr::filter(
    is.na(tree_x) | is.na(tree_y)
    | as.character(tree_x)=="" | as.character(tree_y)==""
  ) %>%
  nrow() %>% 
  stringr::str_c(" trees are missing location data")
```

we'll exclude these trees from the accuracy evaluation since we need location data to perform a spatial instance segmentation analysis

let's check out how many trees are missing height measurement data which we will use for matching with our UAS predicted trees

```{r}
validation_trees %>% 
  dplyr::filter(
    is.na(height_m)
  ) %>%
  nrow() %>% 
  stringr::str_c(" trees are missing height data")
```

we'll exclude these as well and implement the exclusions using a spatial join with the treatment unit boundaries. we'll also set a minimum tree height threshold of `r min_tree_ht_m` m to define what is considered a "tree". a minimum tree height of `r min_tree_ht_m` m will be used to filter the reference field trees which we also applied to our UAS-detected trees.

```{r}
# drop missings and spatial it
validation_trees <- 
  validation_trees %>% 
  dplyr::filter(
    !is.na(plot)
    , stringr::str_trim(plot)!=""
    , !is.na(plot_center_x)
    , !is.na(plot_center_y)
    , !is.na(tree_x)
    , !is.na(tree_y)
    , !is.na(height_m)
    , height_m>=min_tree_ht_m
  ) %>%  
  dplyr::mutate(
    dplyr::across(
      c(crown_length_m, crown_width_m)
      , ~ifelse(.x>100,.x/100,.x)
    )
    , crown_diam_m = dplyr::case_when(
      !is.na(crown_length_m) & !is.na(crown_width_m) ~ (crown_length_m+crown_width_m)/2
      , !is.na(crown_width_m) ~ crown_width_m
      , !is.na(crown_length_m) ~ crown_length_m
      , T ~ as.numeric(NA)
    )
  ) %>% 
  # guessing this is WGS 84
  sf::st_as_sf(coords = c("tree_x", "tree_y"), crs = 4326, remove = F) %>% 
  sf::st_transform(sf::st_crs(stand_boundary))
# attach to treatment unit
validation_trees <-
  validation_trees %>% 
  dplyr::inner_join(
     validation_trees %>% 
      sf::st_intersection(
        stand_boundary %>%
          dplyr::select(unit,block,tretmnt,unit_area_ha) %>% 
          dplyr::rename(type = tretmnt) %>% 
          dplyr::mutate(
            type =
              tolower(type) %>% 
              stringr::str_squish()
          ) %>% 
          dplyr::rename_with(.fn = ~paste0("trtmnt_", .x, recycle0 = T))
      ) %>% 
      sf::st_drop_geometry() %>% 
      dplyr::select(plot,tree_id, tidyselect::starts_with("trtmnt_")) %>% 
      dplyr::group_by(plot,tree_id) %>% 
      dplyr::slice(1) %>% 
      dplyr::ungroup()
    , by = dplyr::join_by(plot,tree_id)
  ) %>% 
  dplyr::mutate(
    is_in_trtmnt_unit = ifelse(is.na(trtmnt_type) | trtmnt_type=="control",F,T)
    # , dplyr::across(c(trtmnt_type,trtmnt_block,trtmnt_unit),ordered)
  )
```

what is the distribution of the field-measured tree size values

```{r}
validation_trees %>% 
  sf::st_drop_geometry() %>% 
  dplyr::select(height_m,dbh_cm,crown_diam_m) %>% 
  tidyr::pivot_longer(
    cols = dplyr::everything()
    , names_to = "metric"
    , values_to = "value"
    , values_drop_na = T
  ) %>% 
  dplyr::mutate(
    tree_metric = metric %>% 
      stringr::str_extract("(height|dbh|diam)") %>% 
      factor(
        ordered = T
        , levels = c(
          "height"
          , "dbh"
          , "diam"
        )
        , labels = c(
          "Height (m)"
          , "DBH (cm)"
          , "Crown Dia. (m)"
        )
      )
  ) %>% 
  # dplyr::glimpse()
# plot dist
  ggplot2::ggplot(mapping = ggplot2::aes(x = value)) +
  ggplot2::geom_density(
    mapping = ggplot2::aes(y=ggplot2::after_stat(scaled))
    , alpha = 0.8
    , fill = "brown", color = "brown"
  ) +
  ggplot2::facet_grid(
    cols = dplyr::vars(tree_metric)
    , scales = "free_x", axes = "all_x"
  ) +
  ggplot2::scale_y_continuous(NULL,breaks=NULL) +
  ggplot2::scale_x_continuous(breaks=scales::breaks_extended(n=8)) +
  ggplot2::labs(
    x=""
  ) +
  ggplot2::theme_light() +
  ggplot2::theme(
    legend.position = "top"
    , strip.text.x = ggplot2::element_text(color = "black", face = "bold")
    , panel.grid.major.y = ggplot2::element_blank()
    , panel.grid.minor.y = ggplot2::element_blank()
  )
```

what is the distribution of tree species?

```{r}
# wut
validation_trees %>% 
  sf::st_drop_geometry() %>% 
  dplyr::count(species) %>% 
  dplyr::ungroup() %>% 
  dplyr::mutate(
    pct = scales::percent(n/sum(n), accuracy = 0.1)
    , n = scales::comma(n)
  ) %>% 
  kableExtra::kbl(
    caption = "field-measured species distribution"
    , col.names = c("species symbol","n", "%")
  ) %>% 
  kableExtra::kable_styling()
```

Utah juniper (*Juniperus osteosperma*, JUOS) and pinyon pine (*Pinus edulis*, PIED)

let's plot the field trees on the study unit boundaries

```{r}
mapview::mapview(
  stand_boundary %>% dplyr::select(type,block,unit)
  # , color = "blue"
  , lwd = 1
  , zcol = "type"
  , col.regions = RColorBrewer::brewer.pal(n=length(unique(stand_boundary$type)),name="Dark2")
  , alpha.regions = 0.5
  , label = F
  , legend = T
  , popup = T
  , layer.name = "study boundary"
  , leaflet.options = list(maxZoom = 16)
) +
mapview::mapview(
  validation_trees %>% dplyr::select(tree_x,tree_y,height_m,dbh_cm)
  , zcol = "height_m"
  , col.regions = viridis::plasma(n=100)
  , cex = 3
  , layer.name = "field tree ht. (m)"
)
```

neat. 

we'll have to get plot center locations as a data set so we can intersect our UAS-detected trees with the plot boundaries. a 10 m radius was used to define the circular plots (`r scales::comma(pi*10^2,accuracy = 0.01,suffix=" square meters")`)

```{r}
# validation_trees$distance %>% summary()
validation_plots <-
  validation_trees %>%
  sf::st_drop_geometry() %>% 
  dplyr::count(plot,plot_center_x,plot_center_y,trtmnt_type,trtmnt_block,trtmnt_unit) %>% 
  dplyr::ungroup() %>% 
  dplyr::rename(n_trees = n) %>% 
  dplyr::mutate(
    plot_radius_m = 10
    , plot_area_m2 = pi*plot_radius_m^2
    , trees_per_ha = n_trees/(plot_area_m2/10000)
  ) %>% 
  # dplyr::group_by(plot) %>% dplyr::filter(dplyr::n()>1)
  # guessing this is WGS 84
  sf::st_as_sf(coords = c("plot_center_x", "plot_center_y"), crs = 4326, remove = F) %>% 
  sf::st_transform(sf::st_crs(stand_boundary))
# validation_plots %>% dplyr::glimpse()
```

we aggregated the field-measured plot data. let's look at the TPH to make sure it aligns with expectations

```{r, fig.height=9}
validation_plots %>% 
  dplyr::ungroup() %>% 
  sf::st_drop_geometry() %>% 
  dplyr::mutate(
    trtmnt_type = ordered(trtmnt_type)
    , my_fct = forcats::fct_cross(trtmnt_type,trtmnt_block,trtmnt_unit)
  ) %>% 
  dplyr::arrange(desc(trtmnt_type),trees_per_ha) %>% 
  dplyr::mutate(my_fct = forcats::fct_reorder(my_fct, trees_per_ha)) %>% 
  ggplot2::ggplot(mapping = ggplot2::aes(x=trees_per_ha,y=my_fct,fill=trtmnt_type)) +
  ggplot2::geom_point(mapping = ggplot2::aes(color=trtmnt_type), color = "gray") +
  ggplot2::geom_boxplot(width = 0.5, alpha = 0.8, outliers = F) +
  ggplot2::facet_grid(
    rows = dplyr::vars(trtmnt_type)
    , scales = "free_y"
    , switch = "y"
    , axes = "all_x"
  ) +
  ggplot2::scale_fill_brewer(palette = "Dark2") +
  ggplot2::scale_color_brewer(palette = "Dark2") +
  ggplot2::scale_x_continuous(labels = scales::comma) +
  ggplot2::theme_light() +
  ggplot2::labs(x = latex2exp::TeX("trees $ha^{-1}$"), y = "", subtitle = "field plot TPH measurements") +
  ggplot2::theme(
    legend.position = "none"
    , axis.text.y = ggplot2::element_text(size = 7)
    , axis.text.x = ggplot2::element_text(size = 8)
    , strip.text = ggplot2::element_text(size = 10, color = "black", face = "bold")
    , strip.placement = "outside"
  )
```

appears to be in-line with expectations but note the variability within the units and across the units by treatment type

let's aggregate these by treatment type

```{r}
validation_plots %>% 
  sf::st_drop_geometry() %>% 
  dplyr::group_by(trtmnt_type) %>% 
  dplyr::summarise(
    n_plots = dplyr::n()
    , n_trees = sum(n_trees)
    , plot_area_m2 = sum(plot_area_m2)
  ) %>% 
  dplyr::ungroup() %>% 
  dplyr::mutate(
    area_ha = (plot_area_m2/10000)
    , trees_per_ha = n_trees/area_ha
    , dplyr::across(
      tidyselect::starts_with("n_")
      , ~scales::comma(.x,accuracy=1)
    )
    , dplyr::across(
      dplyr::where(is.numeric)
      , ~scales::comma(.x,accuracy=.1)
    )
  ) %>% 
  dplyr::select(-plot_area_m2) %>% 
  kableExtra::kbl(
    caption = "Field-Measured tree density by treatment type"
    , col.names = c(
      "treatment", "# plots"
      , "# field-measured trees"
      , "field-measured area (ha)"
      , "Trees<br />ha<sup>-1</sup>"
    )
    , escape = F
    # , digits = 2
  ) %>% 
  kableExtra::kable_styling()

```

finally, let's check out an example individual plot (white) and tree locations (blue) on the RGB

```{r}
plt_temp <- validation_plots %>% 
  dplyr::slice_max(n = 22, order_by = n_trees) %>% 
  dplyr::slice_sample(n=1) %>% 
  dplyr::pull(plot)
# rgb
rgb_rast %>% 
   terra::crop(
    validation_plots %>% 
      dplyr::filter(plot==plt_temp) %>% 
      sf::st_buffer(11) %>% 
      sf::st_bbox() %>% 
      sf::st_as_sfc() %>% 
      sf::st_transform(terra::crs(rgb_rast)) %>% 
      terra::vect()
    , mask=T
  ) %>% 
  terra::plotRGB()
# add plot
validation_plots %>% 
  dplyr::filter(plot==plt_temp) %>% 
  sf::st_buffer(.$plot_radius_m) %>% 
  sf::st_transform(terra::crs(rgb_rast)) %>% 
  terra::vect() %>% 
  terra::plot(add = T, col = NA, border = "white", lwd = 1.3)
# add plot
validation_plots %>% 
  dplyr::filter(plot==plt_temp) %>% 
  sf::st_transform(terra::crs(rgb_rast)) %>% 
  terra::vect() %>% 
  terra::plot(add = T, col = "white", pch = 18, cex = 2)
# add trees
validation_trees %>% 
  dplyr::filter(plot==plt_temp) %>% 
  sf::st_transform(terra::crs(rgb_rast)) %>% 
  terra::vect() %>% 
  terra::plot(add = T, col = "cyan", cex = 1.5)
```

interesting. do you have any thoughts?

let's also look at the tree crowns assuming a perfectly circular shape based on the field-measured horizontal crown length and width values

```{r}
rgb_rast %>% 
   terra::crop(
    validation_plots %>% 
      dplyr::filter(plot==plt_temp) %>% 
      sf::st_buffer(11) %>% 
      sf::st_bbox() %>% 
      sf::st_as_sfc() %>% 
      sf::st_transform(terra::crs(rgb_rast)) %>% 
      terra::vect()
    , mask=T
  ) %>% 
  terra::plotRGB()
# add plot
validation_plots %>% 
  dplyr::filter(plot==plt_temp) %>% 
  sf::st_buffer(.$plot_radius_m) %>% 
  sf::st_transform(terra::crs(rgb_rast)) %>% 
  terra::vect() %>% 
  terra::plot(add = T, col = NA, border = "white", lwd = 1.3)
# add plot
validation_plots %>% 
  dplyr::filter(plot==plt_temp) %>% 
  sf::st_transform(terra::crs(rgb_rast)) %>% 
  terra::vect() %>% 
  terra::plot(add = T, col = "white", pch = 18, cex = 2)
# add trees
# add trees
validation_trees %>% 
  dplyr::filter(plot==plt_temp) %>% 
  sf::st_transform(terra::crs(rgb_rast)) %>% 
  terra::vect() %>% 
  terra::plot(add = T, col = "cyan", cex = 1.5)
# add trees
validation_trees %>% 
  dplyr::filter(plot==plt_temp) %>% 
  dplyr::mutate(crown_radius_m=crown_diam_m/2) %>% 
  sf::st_buffer(.$crown_radius_m) %>% 
  sf::st_transform(terra::crs(rgb_rast)) %>% 
  terra::vect() %>% 
  terra::plot(add = T, col = NA, border = "cyan")

```

the percent canopy cover of that specific example is...

```{r}
cc_temp <-
  validation_trees %>% 
  dplyr::filter(plot==plt_temp) %>% 
  dplyr::mutate(crown_radius_m=crown_diam_m/2) %>% 
  sf::st_buffer(.$crown_radius_m) %>%
  sf::st_intersection(
    validation_plots %>% 
      dplyr::filter(plot==plt_temp) %>% 
      sf::st_buffer(.$plot_radius_m) %>% 
      dplyr::select(plot_area_m2)
  ) %>% 
  dplyr::group_by(plot, plot_area_m2) %>% 
  dplyr::summarise(
    geometry = sf::st_union(.)
  ) %>% 
  dplyr::ungroup() %>% 
  dplyr::mutate(
    cc_area_m2 = sf::st_area(.) %>% as.numeric()
    , cc_pct = cc_area_m2/plot_area_m2 
  )
# add plot
validation_plots %>% 
  dplyr::filter(plot==plt_temp) %>% 
  sf::st_buffer(.$plot_radius_m) %>% 
  terra::vect() %>% 
  terra::plot(
    col = "gray", border = "gray", axes = F
    , main = paste0("canopy cover (green) = ", scales::percent(cc_temp$cc_pct, accuracy=0.1))
  )
# add trees
cc_temp %>% 
  terra::vect() %>% 
  terra::plot(add = T, col = "forestgreen", border = "forestgreen")
```

sure.

```{r, warning=FALSE, message=FALSE, echo=FALSE, include=FALSE}
remove(list = ls()[grep("_temp",ls())])
gc()
```

## Detection Accuracy Overview {#detect_metrics_form}

we'll make a function that uses the following input data parameters to match predicted trees with validation/reference trees:

* `pred_data`: predicted data point locations as `sf` data
* `ref_data`: reference/validation/ground truth data point locations as `sf` data
* `max_dist_m`: radius of search for matching
* `max_height_error_m`: maximum allowable difference in height for to be considered correct match

for work in ponderosa pine forests, we generally set the `max_dist_m` to a 3 m radius to account for tree lean and slight horizontal misalignment in GPS data collection with 3 m representing the crown radius of a mature tree. we also typically set the `max_height_error_m` to 2 m for ponderosa pine evaluation but will need to reduce here to account for shorter trees

information used to inform the definition of "typical" crown width can be found in:

* [Gill et al. 2000](https://doi.org/10.1016/S0378-1127(99)00113-9)
* [Bechtold 2003](https://doi.org/10.1093/sjaf/27.4.269)
* [Bechtold 2004](https://doi.org/10.1093/wjaf/19.4.245)
* [Purves et al. 2007](https://doi.org/10.1371/journal.pone.0000870)

Then, we'll aggregate the instance matching results to evaluate omission rate (false negative rate or miss rate), commission rate (false positive rate), precision, recall (detection rate), and the F-score metric. As a reminder, true positive (TP) instances correctly match ground truth instances with a prediction, commission predictions do not match a ground truth instance (false positive; FP), and omissions are ground truth instances for which no predictions match (false negative; FN)

$$\textrm{omission rate} = \frac{FN}{TP+FN}$$

$$\textrm{commission rate} = \frac{FP}{TP+FP}$$

$$\textrm{precision} = \frac{TP}{TP+FP}$$

$$\textrm{recall} = \frac{TP}{TP+FN}$$

$$
\textrm{F-score} = 2 \times \frac{\bigl(precision \times recall \bigr)}{\bigl(precision + recall \bigr)}
$$


```{r}
###############################################################
###############################################################
###############################################################
###############################################################
# check df function
###############################################################
###############################################################
###############################################################
###############################################################
validate_df_fn <- function(data, required_cols = c("tree_height_m"), check_numeric = F) {
  # check data
  if(!inherits(data,"sf") && !inherits(data,"sfc")){stop("all input data requires POINT sf data")}
  if(
    !all( sf::st_is(data, c("POINT")) )
  ){
    stop("all input data requires POINT sf data")
  }
  # names
  # uses base::setdiff and base::names
  missing_cols <- setdiff(required_cols, names(data))
  
  if(length(missing_cols) > 0){
    stop("all input data requires columns: ", paste(missing_cols, collapse = ", "))
  }
  
  # numeric
  if(check_numeric){
    non_numeric <- data %>%
      sf::st_drop_geometry() %>% 
      dplyr::select(dplyr::all_of(required_cols)) %>%
      dplyr::summarise(dplyr::across(dplyr::everything(), ~ !is.numeric(.x))) %>%
      tidyr::pivot_longer(dplyr::everything()) %>%
      dplyr::filter(value == TRUE)

    if (nrow(non_numeric) > 0) {
      stop("columns must be numeric: ", paste(non_numeric$name, collapse = ", "))
    }
  }
  
  # is.na() catches both NA and NaN; is.infinite() catches Inf
  invalid_dta <- data %>%
    sf::st_drop_geometry() %>% 
    dplyr::summarise(
      dplyr::across(
        dplyr::all_of(required_cols)
        , ~ sum(is.na(.x) | is.infinite(.x))
      )
    ) %>%
    tidyr::pivot_longer(
      cols = dplyr::everything()
      , names_to = "column"
      , values_to = "count"
    ) %>%
    dplyr::filter(count > 0)

  if (nrow(invalid_dta) > 0) {
    bad_cols <- paste(invalid_dta$column, collapse = ", ")
    stop("invalid values (NA, NaN, or Inf) found in: ", bad_cols)
  }
  return(TRUE)
}
###############################################################
###############################################################
###############################################################
###############################################################
# match function
###############################################################
###############################################################
###############################################################
###############################################################
reference_prediction_match <- function(
  pred_data
  , ref_data
  , max_dist_m
  , max_height_error_m
) {
  #######################################################################
  # threshold checks
  #######################################################################
  if(
    any(is.na(max_dist_m)) || 
    any(is.null(max_dist_m)) || 
    !inherits(as.numeric(max_dist_m),"numeric")
  ){
    stop("`max_dist_m` must be numeric and not missing")
  }else{max_dist_m <- as.numeric(max_dist_m)[1]}
  if(
    any(is.na(max_height_error_m)) || 
    any(is.null(max_height_error_m)) || 
    !inherits(as.numeric(max_height_error_m),"numeric")
  ){
    stop("`max_height_error_m` must be numeric and not missing")
  }else{max_height_error_m <- as.numeric(max_height_error_m)[1]}
  #######################################################################
  # data checks
  #######################################################################
  if(nrow(pred_data)==0 | nrow(ref_data)==0){
    warning("no data so...?????")
    # return(NULL)
    
    if(nrow(pred_data)!=0){
      validate_df_fn(data = pred_data, required_cols = c("tree_height_m"), check_numeric = T)
      # sort to start with largest
      pred_data <- pred_data %>% 
        dplyr::arrange(desc(tree_height_m)) %>% 
        dplyr::mutate(pred_match_idx = dplyr::row_number()) %>% 
        # add columns
        dplyr::mutate(
          ref_match_idx = as.numeric(NA)
          , pred_match_height_diff_m= as.numeric(NA)
          , pred_match_dist_m = as.numeric(NA)
          , ref_tree_height_m = as.numeric(NA)
        )
    }else{pred_data <- NULL}
    
    if(nrow(ref_data)!=0){
      validate_df_fn(data = ref_data, required_cols = c("tree_height_m"), check_numeric = T)
      # sort to start with largest
      ref_data <- ref_data %>% 
        dplyr::arrange(desc(tree_height_m)) %>% 
        dplyr::mutate(ref_match_idx = dplyr::row_number()) %>% 
        # add columns
        dplyr::mutate(
          pred_match_idx = 0
          , pred_match_height_diff_m = as.numeric(NA)
          , pred_tree_height_m = as.numeric(NA)
          , pred_match_dist_m = as.numeric(NA)
        )
    }else{ref_data <- NULL}
    # return
    return(list(
      ref_data = ref_data
      , pred_data = pred_data
    ))
  }
  
  validate_df_fn(data = pred_data, required_cols = c("tree_height_m"), check_numeric = T)
  validate_df_fn(data = ref_data, required_cols = c("tree_height_m"), check_numeric = T)
  
  # proj
  if(!identical(
    sf::st_crs(pred_data, parameters = T)[["epsg"]]
    , sf::st_crs(ref_data, parameters = T)[["epsg"]]
  )){stop("data must have same projection. see `sf::st_crs()`")}
  # sort to start with largest
  pred_data <- pred_data %>% 
    dplyr::arrange(desc(tree_height_m)) %>% 
    dplyr::mutate(pred_match_idx = dplyr::row_number())
  ref_data <- ref_data %>% 
    dplyr::arrange(desc(tree_height_m)) %>% 
    dplyr::mutate(ref_match_idx = dplyr::row_number())
  #######################################################################
  # build a 2D kd-tree
  #######################################################################
    # Use the X and Y coordinates of the predicted to build the tree.
    pred_coords_temp <- 
      pred_data %>% 
      dplyr::mutate(
        X = sf::st_coordinates(.)[,1] %>% as.numeric()
        , Y = sf::st_coordinates(.)[,2] %>% as.numeric()
      ) %>% 
      dplyr::select(X,Y) %>% 
      sf::st_drop_geometry()
    # Use the X and Y coordinates of the reference to build the tree.
    ref_coords_temp <- 
      ref_data %>% 
      dplyr::mutate(
        X = sf::st_coordinates(.)[,1] %>% as.numeric()
        , Y = sf::st_coordinates(.)[,2] %>% as.numeric()
      ) %>% 
      dplyr::select(X,Y) %>% 
      sf::st_drop_geometry()
    
    # perform a fixed-radius search using RANN::nn2
    nn_ans_temp <- RANN::nn2(
      # the point cloud xy data (builds the kd-tree on this)
      data = pred_coords_temp
      # the seed point xy data (number of columns must be the same as data)
      , query = ref_coords_temp
      # maximum number of neighbors to find per query
      , k = min(nrow(pred_coords_temp),nrow(ref_coords_temp))
      # the search radius (rcyl)
      , radius = max_dist_m
      , searchtype = "radius"
    )
    # huh?
    # nn_ans_temp %>% class()
    # nn_ans_temp %>% names()
    # nn_ans_temp$nn.idx %>% nrow()
    # nrow(ref_coords_temp)
    # nn_ans_temp$nn.idx %>% ncol()
    # nrow(pred_coords_temp)
    # nn_ans_temp$nn.dists %>% nrow()
    # nrow(ref_coords_temp)
    # nn_ans_temp$nn.dists %>% ncol()
    # nrow(pred_coords_temp)
    # nn_ans_temp$nn.dists[1:11,] %>% dplyr::as_tibble() %>% dplyr::glimpse()
  #######################################################################  
  # iterate through the kd-tree to make matches
  #######################################################################
    if(nrow(ref_coords_temp)!=nrow(ref_data)){stop("bad kd-tree")}
    # add pred match tracker
    ref_data$pred_match_idx <- 0
    ref_data$pred_match_height_diff_m <- as.numeric(NA)
    ref_data$pred_tree_height_m <- as.numeric(NA)
    ref_data$pred_match_dist_m <- as.numeric(NA)
    for(i in 1:nrow(ref_coords_temp)){
      # get indices of points in the current cylinder (0 means no point found)
      nn_idx_pts <- nn_ans_temp$nn.idx[i, ]
      cyl_pts <- nn_idx_pts[nn_idx_pts > 0] 
      # remove already matched
      cyl_pts <- cyl_pts[!(cyl_pts %in% ref_data$pred_match_idx)] 
      # if available matches
      if(dplyr::coalesce(length(cyl_pts),0) >= 1) {
        # get heights
        # ref_data %>% dplyr::select(field_tree_id, tree_height_m) %>% dplyr::glimpse()
        # ref_data$tree_height_m[i]
        # pred_data$tree_height_m[cyl_pts]
        ht_diffs <- abs(ref_data$tree_height_m[i]-pred_data$tree_height_m[cyl_pts])
        # get the minimum value among elements below the threshold
        min_val <- min(ht_diffs[ht_diffs <= max_height_error_m], na.rm = T)
        # if not none within threshold
        if(!is.infinite(min_val)){
          # the index of that value in the original full vector
          cyl_idx <- which(ht_diffs == min_val)[1] # if multiple matches, takes the nearest
          match_idx <- cyl_pts[cyl_idx]
          # distance
          nn_dist_idx <- which(nn_idx_pts == match_idx)
          nn_dist <- nn_ans_temp$nn.dists[i,nn_dist_idx]
          # update match
          ref_data$pred_match_height_diff_m[i] <- min_val
          ref_data$pred_tree_height_m[i] <- pred_data$tree_height_m[match_idx]
          ref_data$pred_match_idx[i] <- match_idx
          ref_data$pred_match_dist_m[i] <- nn_dist
          
        }
      }
    }
  #######################################################################
  # add ref id to pred and return
  #######################################################################
  # pred_data %>% dplyr::glimpse()
  # ref_data %>% dplyr::glimpse()
  pred_data <- pred_data %>% 
    # get rid of columns we'll create
    dplyr::select( -dplyr::any_of(c(
      "hey_xxxxxxxxxx"
      , "pred_match_height_diff_m"
      , "pred_match_dist_m"
      , "ref_tree_height_m"
      , "ref_match_idx"
    ))) %>% 
    dplyr::left_join(
      ref_data %>% 
        sf::st_drop_geometry() %>% 
        dplyr::select(ref_match_idx, pred_match_idx, pred_match_height_diff_m, pred_match_dist_m, tree_height_m) %>% 
        dplyr::rename(ref_tree_height_m=tree_height_m)
      , by = "pred_match_idx"
    )
  return(list(
    ref_data = ref_data
    , pred_data = pred_data
  ))
}
###############################################################
###############################################################
###############################################################
###############################################################
# combine match to classify reference/predictions
###############################################################
###############################################################
###############################################################
###############################################################
ground_truth_prediction_match <- function(
  reference_prediction_match_ans
  , comp_cols = c("tree_height_m")
) {
  # check for missing both missing
  if(
    is.null(reference_prediction_match_ans$ref_data)
    && is.null(reference_prediction_match_ans$pred_data)
  ){
    warning("no data so...?????")
    return(NULL)
  }
  
  # if reference but no predictions...they are all omissions (FN)
  if(is.null(reference_prediction_match_ans$pred_data)){
    return_df <-
      reference_prediction_match_ans$ref_data %>%
      dplyr::filter(pred_match_idx==0 | is.na(pred_match_idx)) %>% 
      dplyr::select(dplyr::all_of(c(
        "ref_match_idx"
        , "pred_match_idx"
        , "pred_match_dist_m"
        # , "pred_tree_height_m"
        # , "tree_height_m"
        , comp_cols
      ))) %>% 
      dplyr::rename_with(
        .cols = dplyr::all_of(comp_cols)
        , .fn = ~paste0("ref_", .x, recycle0 = T)
      ) %>% 
      sf::st_set_geometry("ref_geometry") %>% 
      dplyr::mutate(match_grp = "omission")
    # make match_grp factor
    return_df <- return_df %>% 
      dplyr::mutate(
        match_grp = factor(
          match_grp
          , ordered = T
          , levels = c(
            "true positive"
            , "commission"
            , "omission"
          )
        ) %>% forcats::fct_rev()
      )
    return(return_df)
  }
  
  # if predictions but no reference...they are all commissions (FP)
  if(is.null(reference_prediction_match_ans$ref_data)){
    return_df <-
      reference_prediction_match_ans$pred_data %>%
      dplyr::filter(ref_match_idx==0 | is.na(ref_match_idx)) %>% 
      dplyr::select(dplyr::all_of(c(
        "ref_match_idx"
        , "pred_match_idx"
        , "pred_match_dist_m"
        # , "pred_tree_height_m"
        # , "tree_height_m"
        , comp_cols
      ))) %>% 
      dplyr::rename_with(
        .cols = dplyr::all_of(comp_cols)
        , .fn = ~paste0("pred_", .x, recycle0 = T)
      ) %>% 
      dplyr::mutate(pred_geometry = sf::st_geometry(.)) %>%
      sf::st_drop_geometry() %>% 
      dplyr::mutate(match_grp = "commission")
    # make match_grp factor
    return_df <- return_df %>% 
      dplyr::mutate(
        match_grp = factor(
          match_grp
          , ordered = T
          , levels = c(
            "true positive"
            , "commission"
            , "omission"
          )
        ) %>% forcats::fct_rev()
      )
    return(return_df)
  }
  
  ############# otherwise: nrow(ref_data)>0 & nrow(pred_data)>0
  ############# otherwise: nrow(ref_data)>0 & nrow(pred_data)>0
  ############# otherwise: nrow(ref_data)>0 & nrow(pred_data)>0
  ############# otherwise: nrow(ref_data)>0 & nrow(pred_data)>0
  ############# otherwise: nrow(ref_data)>0 & nrow(pred_data)>0
  # check for comparison cols
  comp_cols <- 
    comp_cols %>% 
    # remove whitespace
    stringr::str_trim() %>% 
    # keep only those with a value
    stringr::str_subset(".")
  if(inherits(comp_cols, "character") && length(comp_cols) > 0){
    validate_df_fn(data = reference_prediction_match_ans$ref_data, required_cols = comp_cols, check_numeric = T)
    validate_df_fn(data = reference_prediction_match_ans$pred_data, required_cols = comp_cols, check_numeric = T)
  }
  # tp matches
  return_df <-
    reference_prediction_match_ans$ref_data %>%
    dplyr::filter(pred_match_idx!=0 & !is.na(pred_match_idx)) %>% 
    dplyr::select(dplyr::all_of(c(
      "ref_match_idx"
      , "pred_match_idx"
      , "pred_match_dist_m"
      # , "pred_tree_height_m"
      # , "tree_height_m"
      , comp_cols
    ))) %>% 
    dplyr::rename_with(
      .cols = dplyr::all_of(comp_cols)
      , .fn = ~paste0("ref_", .x, recycle0 = T)
    ) %>% 
    sf::st_set_geometry("ref_geometry") %>% 
    # add on pred geom
    dplyr::inner_join(
      reference_prediction_match_ans$pred_data %>% 
        dplyr::select(dplyr::all_of(c(
          "ref_match_idx"
          , comp_cols
        ))) %>% 
        dplyr::rename_with(
          .cols = dplyr::all_of(comp_cols)
          , .fn = ~paste0("pred_", .x, recycle0 = T)
        ) %>% 
        dplyr::mutate(pred_geometry = sf::st_geometry(.)) %>%
        sf::st_drop_geometry()
      , by = "ref_match_idx"
    ) %>% 
    dplyr::mutate(match_grp = "true positive")
  
  # add omissions
  return_df <-
    return_df %>% 
    dplyr::bind_rows(
      reference_prediction_match_ans$ref_data %>%
        dplyr::filter(pred_match_idx==0 | is.na(pred_match_idx)) %>% 
        dplyr::select(dplyr::all_of(c(
          "ref_match_idx"
          , "pred_match_idx"
          , "pred_match_dist_m"
          # , "pred_tree_height_m"
          # , "tree_height_m"
          , comp_cols
        ))) %>% 
        dplyr::rename_with(
          .cols = dplyr::all_of(comp_cols)
          , .fn = ~paste0("ref_", .x, recycle0 = T)
        ) %>% 
        sf::st_set_geometry("ref_geometry") %>% 
        dplyr::mutate(match_grp = "omission")
    )
    
  # add commissions
  return_df <-
    return_df %>% 
    dplyr::bind_rows(
      reference_prediction_match_ans$pred_data %>%
        dplyr::filter(ref_match_idx==0 | is.na(ref_match_idx)) %>% 
        dplyr::select(dplyr::all_of(c(
          "ref_match_idx"
          , "pred_match_idx"
          , "pred_match_dist_m"
          # , "pred_tree_height_m"
          # , "tree_height_m"
          , comp_cols
        ))) %>% 
        dplyr::rename_with(
          .cols = dplyr::all_of(comp_cols)
          , .fn = ~paste0("pred_", .x, recycle0 = T)
        ) %>% 
        dplyr::mutate(pred_geometry = sf::st_geometry(.)) %>%
        sf::st_drop_geometry() %>% 
        dplyr::mutate(match_grp = "commission")
    )
  
  # make match_grp factor
  return_df <- return_df %>% 
    dplyr::mutate(
      match_grp = factor(
        match_grp
        , ordered = T
        , levels = c(
          "true positive"
          , "commission"
          , "omission"
        )
      ) %>% forcats::fct_rev()
    )
  
  
  if(inherits(comp_cols, "character") && length(comp_cols) > 0){
    # add a tiny value (eps) to the denominator for pct_diff_ 
    # calcs to prevent division by zero...where pct_diff_ is used for MAPE summary
    eps <- 1e-5
    return_df <- 
      return_df %>% 
      dplyr::mutate(
        # 'diff_' columns are calculated as the predicted value minus the actual value
        dplyr::across(
          .cols = paste0(
            "ref_"
            , comp_cols # c("tree_height_m", "dbh_cm")
            , recycle0 = T
          )
          , .fns = ~ base::get(stringr::str_replace(dplyr::cur_column(), "^ref_", "pred_")) - .x
          , .names = "diff_{stringr::str_remove(.col, '^ref_')}"
        )
        # 'pct_diff_' columns are calculated as the actual value minus the predicted value divided by the actual value
        , dplyr::across(
          .cols = paste0(
            "ref_"
            , comp_cols # c("tree_height_m", "dbh_cm")
            , recycle0 = T
          )
          # 
          , .fns = ~ (.x-base::get(stringr::str_replace(dplyr::cur_column(), "^ref_", "pred_")))/(.x + eps)
          , .names = "pct_diff_{stringr::str_remove(.col, '^ref_')}"
        )
      ) 
  }
  # return
  if(nrow(return_df)==0){
    warning("no records found for ground truth to predition matching")
    return(NULL)
  }else{
    return(return_df)
  }
  
}
```

take that thing for a spin for a single plot. 

To account for tree lean and potential slight spatial misalignment between the UAS point clouds and the ground-based field coordinates along the unit boundary, UAS-predicted trees located outside the boundary but within half the horizontal matching threshold were permitted to qualify as true positives, though unmatched exterior trees were excluded from false positive counts. 

```{r, include=F, eval=F}
plt_temp <- validation_plots %>% 
  dplyr::slice(33) %>% 
  dplyr::pull(plot)
# rgb
rgb_rast %>% 
   terra::crop(
    validation_plots %>% 
      dplyr::filter(plot==plt_temp) %>% 
      sf::st_buffer(15) %>% 
      sf::st_bbox() %>% 
      sf::st_as_sfc() %>% 
      sf::st_transform(terra::crs(rgb_rast)) %>% 
      terra::vect()
    , mask=T
  ) %>% 
  terra::plotRGB()
# add plot
validation_plots %>% 
  dplyr::filter(plot==plt_temp) %>% 
  sf::st_buffer(.$plot_radius_m) %>% 
  sf::st_transform(terra::crs(rgb_rast)) %>% 
  terra::vect() %>% 
  terra::plot(add = T, col = NA, border = "white", lwd = 1.3)
# add trees
validation_trees %>% 
  dplyr::filter(plot==plt_temp) %>% 
  sf::st_transform(terra::crs(rgb_rast)) %>% 
  terra::vect() %>% 
  terra::plot(add = T, col = "cyan", cex = 1.5)

treetops_sf %>% 
  sf::st_intersection(
    validation_plots %>% 
      dplyr::slice(33) %>% 
      dplyr::mutate(plot_radius_m = plot_radius_m+1) %>% # 1 = 0.5*max_dist_m where max_dist_m=allowable search distance for TP match
      sf::st_transform(sf::st_crs(treetops_sf)) %>% 
      sf::st_buffer(.$plot_radius_m) %>% 
      dplyr::select(plot_area_m2)
  ) %>% 
  dplyr::left_join(
    treetops_sf %>% 
      # clip to study boundary
      sf::st_intersection(
        validation_plots %>% 
          dplyr::slice(33) %>% 
          sf::st_transform(sf::st_crs(treetops_sf)) %>% 
          sf::st_buffer(.$plot_radius_m) %>% 
          dplyr::mutate(is_in_stand = T)
      ) %>% 
      sf::st_drop_geometry() %>% 
      dplyr::select(treeID, is_in_stand)
    , by = "treeID"
  ) %>% 
  dplyr::mutate(
    is_in_stand = dplyr::coalesce(is_in_stand,F)
  ) %>% 
  dplyr::filter(tree_height_m>=min_tree_ht_m) %>% 
  # dplyr::glimpse()
  sf::st_transform(terra::crs(rgb_rast)) %>% 
  terra::vect() %>% 
  terra::plot(add = T, col = "magenta", cex = 1.5)
```

```{r}
reference_prediction_match_ans_temp <- reference_prediction_match(
  pred_data = 
    treetops_sf %>% 
      dplyr::filter(tree_height_m>=min_tree_ht_m) %>% 
      sf::st_intersection(
        validation_plots %>% 
          dplyr::slice(33) %>% 
          dplyr::mutate(plot_radius_m = plot_radius_m+1) %>% # 1 = 0.5*max_dist_m where max_dist_m=allowable search distance for TP match
          sf::st_transform(sf::st_crs(treetops_sf)) %>% 
          sf::st_buffer(.$plot_radius_m) %>% 
          dplyr::select(plot_area_m2)
      ) %>% 
      dplyr::left_join(
        treetops_sf %>% 
          # clip to study boundary
          sf::st_intersection(
            validation_plots %>% 
              dplyr::slice(33) %>% 
              sf::st_transform(sf::st_crs(treetops_sf)) %>% 
              sf::st_buffer(.$plot_radius_m) %>% 
              dplyr::mutate(is_in_stand = T)
          ) %>% 
          sf::st_drop_geometry() %>% 
          dplyr::select(treeID, is_in_stand)
        , by = "treeID"
      ) %>% 
      dplyr::mutate(
        is_in_stand = dplyr::coalesce(is_in_stand,F)
      )
  , ref_data = 
    validation_trees %>% 
      dplyr::inner_join(
        validation_plots %>% dplyr::slice(33) %>% sf::st_drop_geometry() %>% dplyr::select(plot)
        , by = "plot"
      ) %>% 
      dplyr::rename(tree_height_m=height_m, dbh_cm=dbh_cm, tree_cbh_m=cbh_m) %>% 
      sf::st_transform(sf::st_crs(treetops_sf))
  , max_dist_m = 2 # round(median(validation_trees$crown_width_m)/2)
  , max_height_error_m = 1.5 # round(sd(validation_trees$height_m)*1.1,1)
)
# # wha?
# reference_prediction_match_ans_temp$ref_data %>% dplyr::glimpse()
# reference_prediction_match_ans_temp$pred_data %>% dplyr::glimpse()
```

the `ground_truth_prediction_match()` uses the output from `reference_prediction_match()` and creates a nice, clean data frame with the instances classified as TP, FP, or FN for use in aggregating to calculate detection accuracy metrics. Optionally, users can set the foundation for quantification accuracy assessment via the `comp_cols` argument (see the `diff_*` and `pct_diff_*` columns in the output)

```{r}
gt_temp <- ground_truth_prediction_match(
  reference_prediction_match_ans = reference_prediction_match_ans_temp
  , comp_cols = c("tree_height_m", "dbh_cm", "tree_cbh_m")
)
# huh?
gt_temp %>% dplyr::glimpse()
```

this is neat, we have the geometry of both the reference and prediction trees as well as the difference between the metric columns we gave the function via the `comp_cols` argument

let's use the different geometries to plot the instance matching classification

now, i need one more palette that fits in the theme and has 3 distinct colors that still look on-theme given the palettes currently used. feel free to use gray as one of the colors and the other two should be somewhat distinct and easy to see when plotted over a RGB imagery of a forest. here are the palettes currently in use: `pal1 <- c("#660012FF","#E59950FF")`; 2) `pal2 <- c("brown")`; 3) `pal3 <- c("#f7c5a5ff","#f2704dff","#cb1b4fff")`; 4) `pal4 <- c("#357ba2ff","#414081ff","#26172aff")`; 5) `pal5 <- c("darkslategray4","darkslategray")`. suggest a few different options for me to choose from

```{r}
# custom palette
pal_match_grp = c(
  "omission"= "plum3" # viridis::cividis(3)[1]
  , "commission"= "gray77"
  , "true positive"= "darkgoldenrod2" # viridis::cividis(3)[3]
)
# scales::show_col(pal_match_grp)
# plt
ggplot2::ggplot() +
  ggplot2::geom_sf(
    data = 
      validation_plots %>% 
      dplyr::slice(33) %>% 
      sf::st_buffer(.$plot_radius_m) %>% 
      sf::st_transform(sf::st_crs(treetops_sf))
    , fill = NA, color = "black"
  ) +
  ggplot2::geom_sf(
    data = gt_temp %>% dplyr::filter(match_grp %in% c("true positive","omission"))
    , mapping = ggplot2::aes(geometry = ref_geometry, color = match_grp)
    , size = 3, alpha = 0.8
  ) +
  ggplot2::geom_sf(
    data = gt_temp %>% dplyr::filter(match_grp=="commission")
    , mapping = ggplot2::aes(geometry = pred_geometry, color = match_grp)
    , size = 3, alpha = 0.8
  ) +
  ggplot2::scale_color_manual(values = pal_match_grp) +
  ggplot2::theme_void() +
  ggplot2::theme(legend.position = "top", legend.title = ggplot2::element_blank()) 
```

### Example: Detection Accuracy

let's make a function to calculate the [detection accuracy metrics](#detect_metrics_form) based on aggregated TP, FP, and FN counts

```{r}
###############################################################
###############################################################
###############################################################
###############################################################
# first function takes df with cols tp_n, fp_n, and fn_n to calculate rates
###############################################################
###############################################################
###############################################################
###############################################################
confusion_matrix_scores_fn <- function(df) {
  df %>% 
  dplyr::mutate(
    omission_rate = dplyr::case_when(
      dplyr::coalesce(tp_n,0) == 0 & dplyr::coalesce(fn_n,0) == 0 ~ 0 # if there are no actual trees, there is nothing to miss
      , dplyr::coalesce(tp_n,0) == 0 & dplyr::coalesce(fn_n,0) > 0 ~ 1 # every single actual tree was missed
      , dplyr::coalesce(fn_n,0) == 0 & dplyr::coalesce(tp_n,0) > 0 ~ 0
      , T ~ fn_n/(tp_n+fn_n)
    ) # False Negative Rate or Miss Rate
    , commission_rate = dplyr::case_when(
      dplyr::coalesce(tp_n,0) == 0 & dplyr::coalesce(fp_n,0) == 0 ~ 0 # if no predictions are made, the model could not have made any commission errors
      , dplyr::coalesce(fp_n,0) == 0 & dplyr::coalesce(tp_n,0) > 0 ~ 0
      , T ~ fp_n/(tp_n+fp_n)
    ) # False Positive Rate
    , precision = dplyr::case_when(
      dplyr::coalesce(tp_n,0) == 0 & dplyr::coalesce(fp_n,0) == 0 ~ 1 # if no predictions are made, the model made zero incorrect positive claims
      , dplyr::coalesce(tp_n,0) == 0 & dplyr::coalesce(fp_n,0) > 0 ~ 0
      , T ~ tp_n/(tp_n+fp_n)
    )
    , recall = dplyr::case_when(
      dplyr::coalesce(tp_n,0) == 0 & dplyr::coalesce(fn_n,0) == 0 ~ 1 # if there are no actual trees, there is nothing to miss
      , dplyr::coalesce(tp_n,0) == 0 & dplyr::coalesce(fn_n,0) > 0 ~ 0 # every single actual tree was missed
      , T ~ tp_n/(tp_n+fn_n)
    )
    , f_score = dplyr::case_when(
      dplyr::coalesce(precision,0) == 0 | dplyr::coalesce(recall,0) == 0 ~ 0
      , T ~ 2 * ( (precision*recall)/(precision+recall) )
    )    
  ) 
}
###############################################################
###############################################################
###############################################################
###############################################################
# aggregate results from ground_truth_prediction_match()
###############################################################
###############################################################
###############################################################
###############################################################
agg_ground_truth_match <- function(ground_truth_prediction_match_ans) {
  if(nrow(ground_truth_prediction_match_ans)==0){return(NULL)}
  if( !(names(ground_truth_prediction_match_ans) %>% stringr::str_equal("match_grp") %>% any()) ){stop("ground_truth_prediction_match_ans must contain `match_grp` column")}
  
  # check for difference columns (contains "_diff") and calc rmse for only those to return a single line df with colums for each diff_rmse
    if(
      (ground_truth_prediction_match_ans %>% 
        sf::st_drop_geometry() %>% 
        dplyr::select(tidyselect::starts_with("diff_") | tidyselect::starts_with("pct_diff_")) %>% 
        ncol() )>0
    ){
      # get rmse and mean difference/error for all columns with "_diff" but not "pct_diff"
      # get mape for all columns with "pct_diff" but not "diff_"
      rmse_df <- 
        ground_truth_prediction_match_ans %>% 
        sf::st_drop_geometry() %>% 
        dplyr::ungroup() %>% 
        dplyr::select(
          tidyselect::starts_with("diff_")
          | tidyselect::starts_with("pct_diff_")
        ) %>%
        tidyr::pivot_longer(dplyr::everything(), values_drop_na = T) %>% 
        dplyr::group_by(name) %>% 
        dplyr::summarise(
          sq = sum(value^2, na.rm = T)
          , mean = mean(value, na.rm = T)
          , sumabs = sum(abs(value), na.rm = T)
          , nomiss = sum(!is.na(value))
        ) %>% 
        dplyr::ungroup() %>% 
        dplyr::mutate(
          rmse = dplyr::case_when(
            dplyr::coalesce(nomiss,0)==0 ~ as.numeric(NA)
            , T ~ sqrt(sq/nomiss)
          )
          , mape = dplyr::case_when(
            dplyr::coalesce(nomiss,0)==0 ~ as.numeric(NA)
            , T ~ sumabs/nomiss
          )
        ) %>% 
        # NA nonsense values
        dplyr::mutate(
          mape = dplyr::case_when(
            stringr::str_starts(name, "pct_diff_") ~ mape
            , T ~ as.numeric(NA)
          )
          , rmse = dplyr::case_when(
            stringr::str_starts(name, "pct_diff_") ~ as.numeric(NA)
            , T ~ rmse
          )
          , mean = dplyr::case_when(
            stringr::str_starts(name, "pct_diff_") ~ as.numeric(NA)
            , T ~ mean
          )
        ) %>% 
        dplyr::select(name,rmse,mean,mape) %>% 
        tidyr::pivot_wider(
          names_from = name
          , values_from = c(rmse,mean,mape)
          , names_glue = "{name}_{.value}"
        ) %>% 
        # remove columns with NA in all rows
        dplyr::select( dplyr::where( ~!all(is.na(.x)) ) )
      
      if(
        dplyr::coalesce(nrow(rmse_df),0)==0
        || dplyr::coalesce(ncol(rmse_df),0)==0
      ){
        # empty df
        rmse_df <- dplyr::tibble()
      }
    }else{
      # empty df
      rmse_df <- dplyr::tibble()
    }
  
  # count by match group
    agg <-
      ground_truth_prediction_match_ans %>% 
      sf::st_drop_geometry() %>% 
      dplyr::ungroup() %>% 
      dplyr::count(match_grp) %>% 
      dplyr::mutate(
        match_grp = dplyr::recode_values(
          match_grp
          , "true positive"~"tp"
          , "commission"~"fp"
          , "omission"~"fn"
        )
      )
    # true positive, false positive, false negative rates
    return_df <- dplyr::tibble(match_grp = c("tp","fp","fn")) %>% 
      dplyr::left_join(agg, by = "match_grp") %>% 
      dplyr::mutate(dplyr::across(.cols = c(n), .fn = ~dplyr::coalesce(.x,0))) %>% 
      tidyr::pivot_wider(
        names_from = match_grp
        , values_from = c(n)
        , names_glue = "{match_grp}_{.value}"
      )
    # rates, precision, recall, f-score
    return_df <- confusion_matrix_scores_fn(return_df)
    # add rmse
    if(nrow(rmse_df)>0){
      return_df <- return_df %>% dplyr::bind_cols(rmse_df)
    }
  # return
  return(return_df)
}
```

```{r, include = F, eval = F}
rp_temp <- reference_prediction_match(
  pred_data = 
    treetops_sf %>% 
      dplyr::filter(tree_height_m>=min_tree_ht_m) %>% 
      sf::st_intersection(
        validation_plots %>% 
          dplyr::slice(33) %>% 
          dplyr::mutate(plot_radius_m = plot_radius_m+1) %>% # 1 = 0.5*max_dist_m where max_dist_m=allowable search distance for TP match
          sf::st_transform(sf::st_crs(treetops_sf)) %>% 
          sf::st_buffer(.$plot_radius_m) %>% 
          dplyr::select(plot_area_m2)
      ) %>% 
      dplyr::left_join(
        treetops_sf %>% 
          # clip to study boundary
          sf::st_intersection(
            validation_plots %>% 
              dplyr::slice(33) %>% 
              sf::st_transform(sf::st_crs(treetops_sf)) %>% 
              sf::st_buffer(.$plot_radius_m) %>% 
              dplyr::mutate(is_in_stand = T)
          ) %>% 
          sf::st_drop_geometry() %>% 
          dplyr::select(treeID, is_in_stand)
        , by = "treeID"
      ) %>% 
      dplyr::mutate(
        is_in_stand = dplyr::coalesce(is_in_stand,F)
      )
      # dplyr::slice(0) ###### !!!!!!!!!!!!!!!!!!!!!!!!! NO RECORDS
  , ref_data = 
    validation_trees %>% 
      dplyr::inner_join(
        validation_plots %>% dplyr::slice(33) %>% sf::st_drop_geometry() %>% dplyr::select(plot)
        , by = "plot"
      ) %>% 
      dplyr::rename(tree_height_m=height_m, dbh_cm=dbh_cm, tree_cbh_m=cbh_m) %>% 
      dplyr::slice(0) ###### !!!!!!!!!!!!!!!!!!!!!!!!! NO RECORDS
  , max_dist_m = 2 # round(median(validation_trees$crown_width_m)/2)
  , max_height_error_m = 1.5 # round(sd(validation_trees$height_m)*1.1,1)
)
# wha?
rp_temp$ref_data %>% dplyr::glimpse()
rp_temp$pred_data %>% dplyr::glimpse()
gggttt_temp <- ground_truth_prediction_match(
  reference_prediction_match_ans = rp_temp
  , comp_cols = c("tree_height_m", "dbh_cm", "tree_cbh_m")
)
# huh?
gggttt_temp %>% dplyr::glimpse()
gggttt_temp$match_grp %>% class()
gggttt_temp$match_grp %>% table()

gggttt_temp %>% agg_ground_truth_match()
```

test it out on the example data (but without the `diff_*` and `pct_diff_*` columns which we'll review next)

```{r}
gt_temp %>% 
  # remove FP that were outside boundary since we included them to allow for edge reference trees to be matched
  dplyr::left_join(
    reference_prediction_match_ans_temp$pred_data %>% 
      sf::st_drop_geometry() %>% 
      dplyr::select(pred_match_idx, is_in_stand)
    , by = "pred_match_idx"
  ) %>% 
  dplyr::filter(
    !(!is_in_stand & match_grp=="commission")
  ) %>% 
  # remove diff cols
  dplyr::select(
    -tidyselect::starts_with("diff_")
    , -tidyselect::starts_with("pct_diff_")
  ) %>% 
  agg_ground_truth_match() %>% 
  dplyr::mutate(dplyr::across(dplyr::where(is.numeric),~round(.x,3))) %>% 
  knitr::kable(format = "markdown")
```

## Quantification Accuracy Overview {#quant_metrics_form}

Quantification accuracy metrics such as RMSE, MAPE, and Mean Error of tree form measurements (e.g. height, diameter) are calculated by aggregating the differences between the estimated tree attributes and the field reference values for instances classified as True Positives. These metrics tell us about the method's ability to accurately quantify the form of the trees it successfully identified.

to prepare our results for analysis, we will develop a function that aggregates the tree-level data into a single record for each metric-error measurement combination. this function will calculate detection performance metrics such as F-score, precision, and recall (using the `confusion_matrix_scores_fn()` we defined above), as well as quantification accuracy metrics including Root Mean Squared Error (RMSE), Mean Error (ME), and Mean Absolute Percentage Error (MAPE) to assess the accuracy of our tree form measurements. this could be a valuable function for any future analysis comparing predictions to reference data.

here are the quantification accuracy metric formulas:

$$
\textrm{RMSE} = \sqrt{ \frac{  \sum_{i=1}^{N} (y_{i} - \hat{y_{i}})^{2}}{N}}
$$


$$
\textrm{ME} = \frac{  \sum_{i=1}^{N} (\hat{y_{i}} - y_{i})}{N}
$$


$$
\textrm{MAPE} = \frac{1}{N} \sum_{i=1}^{N} \left| \frac{y_{i} - \hat{y_{i}}}{y_{i}} \right|
$$

Where $N$ is equal to the total number of correctly matched trees, $y_i$ is the ground truth measured value and $\hat{y_i}$ is the predicted value of $i$

There is a lot going on in our `agg_ground_truth_match()` function but it's application is straightforward enough:

* The minimum required input is a data frame of the raw instance matches with a column named `match_grp` which is string/factor with the levels "true positive", "commission", and "omission" as returned by the `ground_truth_prediction_match()` function
* Optionally, if the data contain columns with the prefix "diff_" the mean error (ME) is calculated for those columns with the return having the suffix "_mean" and the RMSE is calculated for those columns with the return having the suffix "_rmse"
  - interpretation of the ME is enhanced if these "diff_" columns are calculated as the predicted value minus the actual value (e.g. `pred_diameter_m - gt_diameter_m`) 
* Optionally, if the data contain columns with the prefix "pct_diff_" the mean absolute percent error (MAPE) is calculated for those columns with the return having the suffix "_mape"
  - these "pct_diff_" columns are calculated as the actual value minus the predicted value divided by the actual value (e.g. `(gt_diameter_m - pred_diameter_m)/gt_diameter_m`) 
### Example: Quantification Accuracy

we already demonstrated minimum use of the `agg_ground_truth_match()` function to get detection accuracy metrics, now we'll demonstrate the quantification accuracy metrics by keeping the `diff_*` and `pct_diff_*` columns calculated with the `ground_truth_prediction_match()` function

```{r,include=F,eval=F}
gt_temp %>% 
  # remove FP that were outside boundary since we included them to allow for edge reference trees to be matched
  dplyr::left_join(
    reference_prediction_match_ans_temp$pred_data %>% 
      sf::st_drop_geometry() %>% 
      dplyr::select(pred_match_idx, is_in_stand)
    , by = "pred_match_idx"
  ) %>% 
  dplyr::filter(
    !(!is_in_stand & match_grp=="commission")
  ) %>% 
  sf::st_drop_geometry() %>% 
  dplyr::ungroup() %>% 
  dplyr::select(
    tidyselect::starts_with("diff_")
    | tidyselect::starts_with("pct_diff_")
    | tidyselect::contains("cbh_m")
  ) %>% View()
```

```{r}
gt_temp %>% 
  # remove FP that were outside boundary since we included them to allow for edge reference trees to be matched
  dplyr::left_join(
    reference_prediction_match_ans_temp$pred_data %>% 
      sf::st_drop_geometry() %>% 
      dplyr::select(pred_match_idx, is_in_stand)
    , by = "pred_match_idx"
  ) %>% 
  dplyr::filter(
    !(!is_in_stand & match_grp=="commission")
  ) %>% 
  agg_ground_truth_match() %>% 
  dplyr::select(tidyselect::ends_with("_rmse"),tidyselect::ends_with("_mean"),tidyselect::ends_with("_mape")) %>% 
  dplyr::select(-tidyselect::contains("cbh_m")) %>% 
  dplyr::mutate(dplyr::across(dplyr::where(is.numeric),~round(.x,3))) %>% 
  knitr::kable("markdown")
```

nice, that could be useful for evaluation

```{r, warning=FALSE, message=FALSE}
remove(list = ls()[grep("_temp",ls())])
gc()
```

```{r, include=F, eval=FALSE}
# test a bunch of different variable window sizes within `lidR::lmf()` 
# that cover a range of possibilities and function shapes including linear, concave up, and concave down
# empty list
vws_functions <- list()
min_ws_temp <- 0.5
max_ws_temp <- 6.0

n_temp <- 40
# end points for the line
linear_targets <- base::seq(from = max_ws_temp, to = max_ws_temp*2*10, length.out = n_temp)

for (i in 1:n_temp) {
  target_height <- linear_targets[i]
  slope <- (max_ws_temp - min_ws_temp) / target_height
  
  # Pad names with a zero (01 to 60) for clean sorting
  func_name <- base::paste0("linear_", base::sprintf("%02d", i))
  
  v_min  <- base::as.numeric(min_ws_temp)
  v_max  <- base::as.numeric(max_ws_temp)
  v_slope <- base::as.numeric(slope)
  
  func_body <- base::substitute(
    dplyr::case_when(
      base::is.na(x) ~ v_min,
      (v_min + v_slope * x) < v_min ~ v_min,
      (v_min + v_slope * x) > v_max ~ v_max,
      TRUE ~ (v_min + v_slope * x)
    ),
    list(v_min = v_min, v_max = v_max, v_slope = v_slope)
  )
  
  vws_functions[[func_name]] <- base::as.function(c(base::alist(x = ), func_body))
}
vws_functions[1]
vws_functions[[1]]
class(vws_functions[[1]])
vws_functions[[1]](1.5)
vws_functions[[1]](0)
vws_functions[[1]](18)

vws_functions[[40]](1.5)
vws_functions[[40]](0)
vws_functions[[40]](18)

# --- Concave Down / Saturating Functions (30 unique variations) ---
# total potential growth (or distance) between the floor and the ceiling is: 6.0-0.5=5.5
# the halfway point of growth is exactly half of that distance: 5.5/2=2.75
# when the curve reaches its exact halfway point, the resulting window size must be the baseline floor plus that half-distance: 0.5+2.75=3.25
# in `half_sat`, setting the constant of 1.375 means the function will hit its halfway mark of 3.25 when the tree is only 1.375 meters tall
  # max slope = 4 (for ever 1 m growth, ws increases by 4 m)
  # 4 = (6.0-0.5)/c >> 4 = 5.5/c >> c = 5.5/4 >> c = 1.375
# how far out to look based on maximum plot/tree height? 45 m tree??
n_temp <- 30
for (i in 1:n_temp) {
  c <- 1.375 # ~dbh height
  max_end <- 45 # biggest tree everrrrr
  half_sat <- c + (i - 1) * ( (max_end-c) / (n_temp-1))
  
  func_name <- base::paste0("concave_down_", base::sprintf("%02d", i))
  
  v_min  <- base::as.numeric(min_ws_temp)
  v_max  <- base::as.numeric(max_ws_temp)
  v_half <- base::as.numeric(half_sat)
  
  func_body <- base::substitute(
    dplyr::case_when(
      base::is.na(x) ~ v_min,
      (v_min + ((v_max - v_min) * x) / (v_half + x)) < v_min ~ v_min,
      (v_min + ((v_max - v_min) * x) / (v_half + x)) > v_max ~ v_max,
      TRUE ~ (v_min + ((v_max - v_min) * x) / (v_half + x))
    ),
    list(v_min = v_min, v_max = v_max, v_half = v_half)
  )
  
  vws_functions[[func_name]] <- base::as.function(c(base::alist(x = ), func_body))
}


# 3. EVALUATE FUNCTIONS OVER A RANGE FOR VISUALIZATION -----------------------

height_sequence <- base::seq(0, 45, by = 0.25)

plot_data <- 
  purrr::map_dfr(
  names(vws_functions)
  , function(name) {
    tibble::tibble(
      height = height_sequence,
      window_size = vws_functions[[name]](height_sequence),
      function_id = name,
      shape_type = dplyr::case_when(
        base::grepl("linear", name) ~ "linear",
        base::grepl("concave_down", name) ~ "concave down"
      )
    )
  }
)


ggplot2::ggplot(
  data = plot_data, 
  mapping = ggplot2::aes(x = height, y = window_size, group = function_id, color = shape_type)
) +
  ggplot2::geom_line(alpha = 0.6, linewidth = 0.7) +
  ggplot2::labs(
    x = "CHM (m)",
    , y = "ws diameter (m)"
    , color = "ws type"
  ) +
  ggplot2::theme_minimal() +
  ggplot2::theme(
    legend.position = "bottom",
    plot.title = ggplot2::element_text(face = "bold", size = 13),
    panel.grid.minor = ggplot2::element_blank()
  )

names(vws_functions)
vws_functions[["concave_down_05"]]
# function(x){ 
#   dplyr::case_when(
#     is.na(x) ~ 0.5
#     , (0.5 + ((6 - 0.5) * x)/(7.39224 + x)) < 0.5 ~ 0.5
#     , (0.5 + ((6 - 0.5) * x)/(7.39224 + x)) > 6 ~ 6
#     , TRUE ~ (0.5 + ((6 - 0.5) * x)/(7.39224 + x))
#   )
# }
vws_functions[["concave_down_06"]]
# function(x){ 
#   dplyr::case_when(
#     is.na(x) ~ 0.5
#     , (0.5 + ((6 - 0.5) * x)/(8.89655 + x)) < 0.5 ~ 0.5
#     , (0.5 + ((6 - 0.5) * x)/(8.89655 + x)) > 6 ~ 6
#     , TRUE ~ (0.5 + ((6 - 0.5) * x)/(8.89655 + x))
#   )
# }

#### set up CHM for raster2trees
agg_chm_rast <- terra::rast( file.path(c2t_process_dir, "chm_0.20m.tif") )
if(!file.exists("../data/chm_rast_test.tif")){
  chm_rast_test <- 
    agg_chm_rast %>% 
    terra::crop(
      validation_plots %>% 
        sf::st_buffer(.$plot_radius_m) %>% 
        sf::st_union() %>% 
        sf::st_buffer(8) %>% 
        sf::st_transform(terra::crs(agg_chm_rast)) %>% 
        terra::vect()
      , mask = T
      , filename = "../data/chm_rast_test.tif"
    )
}else{
  chm_rast_test <- terra::rast("../data/chm_rast_test.tif")
}
chm_rast_test <- chm_rast_test*1
# chm_rast_test %>% terra::plot()

testtest_fn <- function(nm) {
  if (exists("my_temp_dir") && dir.exists(my_temp_dir)) {
    unlink(my_temp_dir, recursive = TRUE, force = TRUE)
  }
  if (exists("my_temp_dir2") && dir.exists(my_temp_dir2)) {
    unlink(my_temp_dir2, recursive = TRUE, force = TRUE)
  }
  warning(paste0("making dir: ",nm,"....",Sys.time()),immediate. = T,call. = F)
  my_temp_dir <- tempfile(pattern = nm)
  dir.create(my_temp_dir, showWarnings = FALSE)
  my_temp_dir2 <- tempfile(pattern = paste0("xx",nm))
  dir.create(my_temp_dir2, showWarnings = FALSE)
  warning(paste0("raster2trees: ",nm,"....",Sys.time()),immediate. = T,call. = F)
  xxx_raster2trees_ans <- cloud2trees::raster2trees(
      chm_rast = chm_rast_test
      , outfolder = my_temp_dir
      , ws = vws_functions[[nm]]
      , min_height = 1.37
      , min_crown_area = 0.5
      , tempdir = my_temp_dir2
    )
  # xxx_raster2trees_ans %>% names()
  xxx_treetops_sf <- xxx_raster2trees_ans %>% 
    sf::st_drop_geometry() %>% 
    sf::st_as_sf(coords = c("tree_x", "tree_y"), crs = sf::st_crs(xxx_raster2trees_ans)) %>% 
    dplyr::mutate(dbh_cm = mean(validation_trees$dbh_cm,na.rm=T))
  # xxx_treetops_sf %>% mapview::mapview(zcol="tree_height_m", layer.name="trees")
  # nrow(xxx_treetops_sf)
  xxx_treetops_sf <- 
    sf::st_filter(
      xxx_treetops_sf
      , piles_workflow_ans %>% 
        dplyr::filter(is_in_trtmnt_unit, inrange_th_votes>=5) %>% 
        sf::st_union()
      , .predicate = sf::st_disjoint
    )
  # nrow(xxx_treetops_sf)
  warning(paste0("reference_prediction_match: ",nm,"....",Sys.time()),immediate. = T,call. = F)
  ################### map it
  xxx_gt_pred_match_map_ans <- validation_trees$plot %>% 
    unique() %>% 
    purrr::set_names() %>% 
    purrr::map(function(plt_id) {
      # reference_prediction_match
      rp_match_ans_temp <- reference_prediction_match(
        pred_data = 
          xxx_treetops_sf %>% 
            dplyr::filter(tree_height_m>=min_tree_ht_m) %>% 
            sf::st_intersection(
              validation_plots %>% 
                dplyr::filter(plot==plt_id) %>%  
                dplyr::mutate(plot_radius_m = plot_radius_m+1) %>% 
                sf::st_transform(sf::st_crs(xxx_treetops_sf)) %>% 
                sf::st_buffer(.$plot_radius_m) %>% 
                dplyr::select(plot_area_m2)
            ) %>% 
            dplyr::left_join(
              xxx_treetops_sf %>% 
                # clip to study boundary
                sf::st_intersection(
                  validation_plots %>% 
                    dplyr::filter(plot==plt_id) %>%  
                    sf::st_transform(sf::st_crs(xxx_treetops_sf)) %>% 
                    sf::st_buffer(.$plot_radius_m) %>% 
                    dplyr::mutate(is_in_stand = T)
                ) %>% 
                sf::st_drop_geometry() %>% 
                dplyr::select(treeID, is_in_stand)
              , by = "treeID"
            ) %>% 
            dplyr::mutate(
              is_in_stand = dplyr::coalesce(is_in_stand,F)
            )
        , ref_data = 
          validation_trees %>% 
            dplyr::filter(plot==plt_id) %>%
            dplyr::rename(tree_height_m=height_m, dbh_cm=dbh_cm, tree_cbh_m=cbh_m) %>% 
            sf::st_transform(sf::st_crs(xxx_treetops_sf))
        , max_dist_m = 2 # round(median(validation_trees$crown_width_m)/2)
        , max_height_error_m = 1.5 # round(sd(validation_trees$height_m)*1.1,1)
      )
      # # wha?
      # rp_match_ans_temp$ref_data %>% dplyr::glimpse()
      # rp_match_ans_temp$pred_data %>% dplyr::glimpse()
      
      gt_temp <- ground_truth_prediction_match(
        reference_prediction_match_ans = rp_match_ans_temp
        , comp_cols = c("tree_height_m", "dbh_cm") #, "tree_cbh_m")
      )
      
      if(!is.null(rp_match_ans_temp$pred_data) && nrow(rp_match_ans_temp$pred_data)>0){
        gt_temp <- gt_temp %>% 
          # remove FP that were outside boundary since we included them to allow for edge reference trees to be matched
          dplyr::left_join(
            rp_match_ans_temp$pred_data %>% 
              sf::st_drop_geometry() %>% 
              dplyr::select(pred_match_idx, is_in_stand)
            , by = "pred_match_idx"
          ) %>% 
          dplyr::filter(
            !(!is_in_stand & match_grp=="commission")
          ) 
      }
      # gt_temp %>% agg_ground_truth_match() %>% dplyr::glimpse()
      return(gt_temp)
  })
  xxx_ground_truth_prediction_match_ans <- 
    xxx_gt_pred_match_map_ans %>%
    purrr::compact() %>% 
    purrr::list_rbind(names_to = "plot")
  ans <- xxx_ground_truth_prediction_match_ans %>% 
    sf::st_drop_geometry() %>% 
    agg_ground_truth_match()
  ans %>% readr::write_csv(file.path("../data/testtest", paste0(nm,".csv")))
  return(ans)
}

# testtest_ans <- 
#   names(vws_functions) %>%
#   # .[2:3] %>% 
#   purrr::set_names() %>% 
#   purrr::map(testtest_fn)
# testtest_ans %>% 
#   purrr::compact() %>% 
#   purrr::list_rbind(names_to = "vws_function_nm") %>% 
#   dplyr::glimpse()

readr::write_csv(file.path("../data/testtest", paste0(nm,".csv")))

p_temp <- 
  testtest_ans %>% 
  purrr::compact() %>% 
  purrr::list_rbind(names_to = "vws_function_nm") %>% 
  ggplot(aes(x=recall,y=precision,color=f_score,label=vws_function_nm)) +
  geom_point() +
  scale_color_viridis_c()
p_temp
plotly::ggplotly(p_temp)
```

```{r, warning=FALSE, message=FALSE, echo=FALSE, include=FALSE}
remove(list = ls()[grep("_temp",ls())])
gc()
```

## Implement Accuracy Evaluation

because we have plot-level reference field data but a complete census of predictions, we need to perform accuracy evaluation for each plot and then aggregate to the entire study site. 

let's perform instance matching over each plot and then bring those results together after

```{r}
# make a temp function to map over each plot
inst_match_fn_temp <- function(plt_id) {
  # reference_prediction_match
  rp_match_ans_temp <- reference_prediction_match(
    pred_data = 
      treetops_sf %>% 
        dplyr::filter(tree_height_m>=min_tree_ht_m) %>% 
        sf::st_intersection(
          validation_plots %>% 
            dplyr::filter(plot==plt_id) %>%  
            dplyr::mutate(plot_radius_m = plot_radius_m+1) %>% # 1 = 0.5*max_dist_m where max_dist_m=allowable search distance for TP match
            sf::st_transform(sf::st_crs(treetops_sf)) %>% 
            sf::st_buffer(.$plot_radius_m) %>% 
            dplyr::select(plot_area_m2)
        ) %>% 
        dplyr::left_join(
          treetops_sf %>% 
            # clip to study boundary
            sf::st_intersection(
              validation_plots %>% 
                dplyr::filter(plot==plt_id) %>%  
                sf::st_transform(sf::st_crs(treetops_sf)) %>% 
                sf::st_buffer(.$plot_radius_m) %>% 
                dplyr::mutate(is_in_stand = T)
            ) %>% 
            sf::st_drop_geometry() %>% 
            dplyr::select(treeID, is_in_stand)
          , by = "treeID"
        ) %>% 
        dplyr::mutate(
          is_in_stand = dplyr::coalesce(is_in_stand,F)
        )
    , ref_data = 
      validation_trees %>% 
        dplyr::filter(plot==plt_id) %>%
        dplyr::rename(tree_height_m=height_m, dbh_cm=dbh_cm, tree_cbh_m=cbh_m) %>% 
        sf::st_transform(sf::st_crs(treetops_sf))
    , max_dist_m = 2 # round(median(validation_trees$crown_width_m)/2)
    , max_height_error_m = 1.5 # round(sd(validation_trees$height_m)*1.1,1)
  )
  # # wha?
  # rp_match_ans_temp$ref_data %>% dplyr::glimpse()
  # rp_match_ans_temp$pred_data %>% dplyr::glimpse()
  
  gt_temp <- ground_truth_prediction_match(
    reference_prediction_match_ans = rp_match_ans_temp
    , comp_cols = c("tree_height_m", "dbh_cm") #, "tree_cbh_m")
  )
  
  if(!is.null(rp_match_ans_temp$pred_data) && nrow(rp_match_ans_temp$pred_data)>0){
    gt_temp <- gt_temp %>% 
      # remove FP that were outside boundary since we included them to allow for edge reference trees to be matched
      dplyr::left_join(
        rp_match_ans_temp$pred_data %>% 
          sf::st_drop_geometry() %>% 
          dplyr::select(pred_match_idx, is_in_stand)
        , by = "pred_match_idx"
      ) %>% 
      dplyr::filter(
        !(!is_in_stand & match_grp=="commission")
      ) 
  }
  # gt_temp %>% agg_ground_truth_match() %>% dplyr::glimpse()
  return(gt_temp)
}


################### map it
################### map it
################### map it
gt_pred_match_map_ans <- validation_trees$plot %>% 
  unique() %>% 
  purrr::set_names() %>% 
  purrr::map(inst_match_fn_temp)

# gt_pred_match_map_ans %>% 
#   purrr::map(~nrow(.x) %>% dplyr::as_tibble() %>% dplyr::rename(nrow=1)) %>% 
#   purrr::list_rbind(names_to = "plot") %>% 
#   dplyr::glimpse()
# 
# gt_pred_match_map_ans %>% 
#   purrr::map(~is.null(.x) %>% dplyr::as_tibble() %>% dplyr::rename(is_null=1)) %>% 
#   purrr::list_rbind(names_to = "plot") %>% 
#   dplyr::count(is_null)

# # what if we get NULL back in the map list????
# # get empty data frame based on the str of the first non-NULL data frame
# df_temp <- 
#   gt_pred_match_map_ans %>%
#   purrr::detect(is.data.frame) %>%
#   slice(0)
# # fill the NULL and bind the rows
# final_ans <- gt_pred_match_map_ans %>%
#   purrr::modify_if(is.null, ~ df_temp) %>%
#   purrr::list_rbind(names_to = "plot")

# combine into one big tracking data
ground_truth_prediction_match_ans <- 
  gt_pred_match_map_ans %>%
  purrr::compact() %>% 
  purrr::list_rbind(names_to = "plot")
```

```{r, warning=FALSE, message=FALSE, echo=FALSE, include=FALSE}
remove(list = ls()[grep("_temp",ls())])
gc()
```

now we have a single data set with all predicted trees and all reference trees within any field plot classified using the instance segmentation matching approach defined above

```{r}
ground_truth_prediction_match_ans %>% dplyr::glimpse()
```

we can use this data to get study-wide detection and quantification accuracy metrics

## Detection Accuracy Results

first, we'll check out the site-wide detection accuracy results

```{r}
agg_ground_truth_match_ans <-
  ground_truth_prediction_match_ans %>% 
  sf::st_drop_geometry() %>% 
  agg_ground_truth_match() %>% 
  dplyr::bind_cols(
    ground_truth_prediction_match_ans %>% 
      sf::st_drop_geometry() %>% 
      dplyr::summarise(plots_n = dplyr::n_distinct(plot))
  ) %>% 
  dplyr::relocate(plots_n)
# table
agg_ground_truth_match_ans %>% 
  dplyr::select(
    tidyselect::ends_with("_n")
    , precision, recall, f_score
  ) %>% 
  dplyr::mutate(
    dplyr::across(
      tidyselect::ends_with("_n")
      , ~scales::comma(.x,accuracy=1)
    )
    , dplyr::across(
      dplyr::where(is.numeric)
      , ~scales::percent(.x,accuracy=1)
    )
  ) %>% 
  dplyr::mutate(data = "site-wide") %>% 
  dplyr::relocate(data) %>% 
  kableExtra::kbl(
    caption = "Site-Wide Detection Accuracy"
    , col.names = c(
      "data", "# plots"
      , "TP predictions", "FP predictions", "FN predictions"
      , "Precision", "Recall", "F-score"
    )
    , escape = F
    # , digits = 2
  ) %>% 
  kableExtra::kable_styling()
```

looks pretty good for a first check 

what if we look at the detection accuracy results by treatment type?

```{r}
agg_ground_truth_match_by_type_ans <- 
  validation_plots$trtmnt_type %>% 
  unique() %>% 
  purrr::map(function(x){
    which_dta <- 
      ground_truth_prediction_match_ans %>% 
      sf::st_drop_geometry() %>% 
      dplyr::inner_join(
        validation_plots %>% 
          sf::st_drop_geometry() %>% 
          dplyr::filter(trtmnt_type==x) %>% 
          dplyr::select(plot,trtmnt_type)
        , by = "plot"
      )
    which_dta %>% 
      dplyr::group_by(trtmnt_type) %>% 
      dplyr::summarise(plots_n = dplyr::n_distinct(plot)) %>% 
      dplyr::ungroup() %>% 
      dplyr::bind_cols(
        agg_ground_truth_match(which_dta)
      )
  }) %>% 
  purrr::list_rbind() %>% 
  dplyr::mutate(trtmnt_type = ordered(trtmnt_type))

# table
agg_ground_truth_match_by_type_ans %>% 
  dplyr::select(
    trtmnt_type
    , tidyselect::ends_with("_n")
    , precision, recall, f_score
  ) %>% 
  dplyr::mutate(
    dplyr::across(
      tidyselect::ends_with("_n")
      , ~scales::comma(.x,accuracy=1)
    )
    , dplyr::across(
      dplyr::where(is.numeric)
      , ~scales::percent(.x,accuracy=0.1)
    )
  ) %>% 
  dplyr::arrange(trtmnt_type) %>% 
  kableExtra::kbl(
    caption = "Detection Accuracy by Treatment Type"
    , col.names = c(
      "treatment", "# plots"
      , "TP predictions", "FP predictions", "FN predictions"
      , "Precision", "Recall", "F-score"
    )
    , escape = F
    # , digits = 2
  ) %>% 
  kableExtra::kable_styling()
```

interesting to see the variation in detection accuracy across the different treatment types. the more aggressive treatments (i.e. treatments that removed more trees and left fewer residual trees) have a smaller margin for error with respect to tree detection accuracy since each tree carries more weight. however, the more open stand conditions created by the more aggressive treatments should make it easier to detect each individual if they are spaced with openings between individuals. at the same time the units with a high proportion of slash piles might result in increased commission rate (i.e. higher number of false positive tree predictions) even though we applied a slash pile mask. an additional data check may be needed for the remaining trees to check whether or not they are "green" while it is not clear if only live or both dead and live trees were sampled during the field inventory. and many, many more questions...

let's see the relationship between tree detection (recall rate) and tree height within each flight. we'll look at true positive and omissions (FN) and use the field-measured tree height to quantify the proportion of trees within a height bin that were detected by the UAS method

```{r}
# make ggplot2::cut_* bins pretty
make_bins_pretty <- function(bin_factor, suffix = "") {
  # convert factor to character and remove brackets and parens
  clean_labels <- bin_factor %>%
    as.character() %>%
    stringr::str_replace_all(c("\\[" = "", "\\(" = "", "\\]" = "", "\\)" = "")) %>%
    stringr::str_replace(",", "-")
  
  # suffix
  if(stringr::str_squish(suffix) != "" && inherits(suffix,"character")){
    clean_labels <- stringr::str_c(clean_labels, " ", suffix) %>% stringr::str_squish()
  }
  
  # reorder based on the original factor levels 
  forcats::fct_reorder(factor(clean_labels), as.numeric(bin_factor))
}

ground_truth_prediction_match_ans %>%
  sf::st_drop_geometry() %>% 
  dplyr::filter(match_grp %in% c("true positive","omission")) %>% 
  # dplyr::glimpse()
  # metric cuts/bins
  dplyr::mutate(
    tree_height_m_bin = ggplot2::cut_width(ref_tree_height_m, width = 3, boundary = 0, closed = "left") %>% 
      make_bins_pretty(suffix = "m") %>% 
      forcats::fct_relabel(~ stringr::str_replace(.x, "^0-", paste0(min_tree_ht_m,"-")) )
    , is_detected = match_grp %in% c("true positive")
    , is_detected = as.numeric(is_detected)
  ) %>% 
  dplyr::count(match_grp, tree_height_m_bin) %>% 
  dplyr::group_by(tree_height_m_bin) %>% 
  dplyr::mutate(pct = n/sum(n,na.rm = T)) %>% 
  dplyr::ungroup() %>% 
  dplyr::mutate(
    lab = ifelse(
      pct>0.095
      , paste0(
        scales::percent(pct,accuracy=1) 
        ,"\n("
        , scales::comma(n,accuracy=1) 
        , ")"
      )
      , ""
    )
  ) %>% 
  dplyr::ungroup() %>% 
  # dplyr::glimpse()
  ggplot2::ggplot(
    mapping = ggplot2::aes(y = pct, x = tree_height_m_bin)
  ) +
  ggplot2::geom_col(
    mapping = ggplot2::aes(fill = match_grp)
    , width = 0.6
    , color = NA, alpha = 0.9
  ) +
  ggplot2::geom_text(
    mapping = ggplot2::aes(label = lab, group = match_grp, fontface = "bold", color = match_grp)
    # , color = "black"
    , size = 2.5
    , position = ggplot2::position_stack(vjust = 0.5)
  ) +
  ggplot2::scale_fill_manual(values = pal_match_grp) +
  ggplot2::scale_color_manual(values = c("white","black"), guide = "none") +
  ggplot2::scale_x_discrete(drop = F) +
  ggplot2::scale_y_continuous(
    breaks = seq(0,1,by=0.2)
    , labels = scales::percent
    , expand = ggplot2::expansion(mult = c(0,0.02))
  ) +
  ggplot2::labs(
    x = "tree height", y = "", fill = ""
    , subtitle = "Recall rate by tree height bin"
  ) +
  ggplot2::theme_light()+
  ggplot2::theme(
    legend.position = "top"
    , strip.text = ggplot2::element_text(size = 11, color = "black", face = "bold")
    , axis.text.x = ggplot2::element_text(size = 8, angle = 45, hjust = 1, vjust = 1)
    , axis.text.y = ggplot2::element_blank()
    , axis.ticks.y = ggplot2::element_blank()
  )
```

the trend of increasing detection rate with tree height demonstrated with this data aligns well with prior work using aerial point cloud data for single-tree forest inventories

let's look at the commission rate (commission error) by tree height using the UAS-detected tree heights to determine if our ITD variable window function could be improved

```{r, fig.height=10}
ground_truth_prediction_match_ans %>%
  sf::st_drop_geometry() %>% 
  dplyr::filter(match_grp %in% c("true positive","commission")) %>% 
  # dplyr::glimpse()
  # metric cuts/bins
  dplyr::mutate(
    tree_height_m_bin = ggplot2::cut_width(pred_tree_height_m, width = 3, boundary = 0, closed = "left") %>% 
      make_bins_pretty(suffix = "m") %>% 
      forcats::fct_relabel(~ stringr::str_replace(.x, "^0-", paste0(min_tree_ht_m,"-")) )
    , is_detected = match_grp %in% c("true positive")
    , is_detected = as.numeric(is_detected)
  ) %>% 
  dplyr::count(match_grp, tree_height_m_bin) %>% 
  dplyr::group_by(tree_height_m_bin) %>% 
  dplyr::mutate(pct = n/sum(n,na.rm = T)) %>% 
  dplyr::ungroup() %>% 
  dplyr::mutate(
    lab = ifelse(
      pct>0.095
      , paste0(
        scales::percent(pct,accuracy=1) 
        ,"\n("
        , scales::comma(n,accuracy=1) 
        , ")"
      )
      , ""
    )
    , match_grp = forcats::fct_rev(match_grp)
  ) %>% 
  dplyr::ungroup() %>% 
  # dplyr::glimpse()
  ggplot2::ggplot(
    mapping = ggplot2::aes(y = pct, x = tree_height_m_bin)
  ) +
  ggplot2::geom_col(
    mapping = ggplot2::aes(fill = match_grp)
    , width = 0.6
    , color = NA, alpha = 0.9
  ) +
  ggplot2::geom_text(
    mapping = ggplot2::aes(label = lab, group = match_grp, fontface = "bold", color = match_grp)
    , color = "black"
    , size = 2.5
    , position = ggplot2::position_stack(vjust = 0.5)
  ) +
  ggplot2::scale_fill_manual(values = pal_match_grp) +
  # ggplot2::scale_color_manual(values = c("black","white"), guide = "none") +
  ggplot2::scale_x_discrete(drop = F) +
  ggplot2::scale_y_continuous(
    breaks = seq(0,1,by=0.2)
    , labels = scales::percent
    , expand = ggplot2::expansion(mult = c(0,0.02))
  ) +
  ggplot2::labs(
    x = "tree height", y = "", fill = ""
    , subtitle = "Commission rate by tree height bin"
  ) +
  ggplot2::theme_light()+
  ggplot2::theme(
    legend.position = "top"
    , strip.text = ggplot2::element_text(size = 11, color = "black", face = "bold")
    , axis.text.x = ggplot2::element_text(size = 8, angle = 45, hjust = 1, vjust = 1)
    , axis.text.y = ggplot2::element_blank()
    , axis.ticks.y = ggplot2::element_blank()
  )
```

the commission error (percentage shown in the gray bar) is highest for the small trees which suggests we could potentially improve our ITD window function to decrease the number of trees detected in this height range. However, going back to the detection rate above, this same height range also had the lowest detection rate which would suggest we increase the number of trees identified in this height range. This conflict suggests that the trees in this height range are likely difficult to detect for the study area using the data and ITD methods used in this study. For example, they may be subordinate to directly adjacent taller trees with the crowns interlocking, so they are often grouped with the taller tree during the CHM-based ITD.

## Quantification Accuracy Results{#quant_acc}

let's table the tree-level aggregated measurement error

```{r}
agg_ground_truth_match_ans %>% 
  dplyr::mutate(
    ref_trees = tp_n+fn_n
    , det_trees = tp_n+fp_n
  ) %>% 
  dplyr::select(
    # detection
    ref_trees
    , det_trees
    , tp_n
    , omission_rate,commission_rate,recall,precision,f_score
    # quantification
    , tidyselect::ends_with("_mean")
    , tidyselect::ends_with("_rmse")
    , tidyselect::ends_with("_mape")
  ) %>% 
  # second select to arrange tree_metric
  dplyr::select(
    # detection
    ref_trees
    , det_trees
    , tp_n
    , omission_rate,commission_rate,recall,precision,f_score
    # quantification
    # , c(tidyselect::contains("volume") & !tidyselect::contains("paraboloid"))
    , tidyselect::contains("area")
    , tidyselect::contains("height")
    , tidyselect::contains("diameter")
    , tidyselect::contains("dbh")
  ) %>% 
  # names()
  dplyr::mutate(
    dplyr::across(
      .cols = c(
        f_score, recall, precision, tidyselect::ends_with("_mape")
        , tidyselect::ends_with("_rate")
      )
      , .fn = ~ scales::percent(.x, accuracy = 1)
    )
    , dplyr::across(
      .cols = c(tidyselect::ends_with("_mean"))
      , .fn = ~ scales::comma(.x, accuracy = 0.01)
    )
    , dplyr::across(
      .cols = c(tidyselect::ends_with("_rmse"))
      , .fn = ~ scales::comma(.x, accuracy = 0.01)
    )
  ) %>% 
  dplyr::select(
    !tidyselect::contains("_area_")
    & !tidyselect::contains("diff_diameter_")
    & !tidyselect::ends_with("_trees")
    # & !tidyselect::ends_with("_n")
    & !tidyselect::ends_with("_rate")
  ) %>% 
  dplyr::select(
    -c(
      recall,precision,f_score
    )
  ) %>% 
  dplyr::mutate(data = "site-wide") %>% 
  dplyr::relocate(data) %>% 
  # ncol()
  # dplyr::glimpse()
  kableExtra::kbl(
    caption = "Quantification Accuracy"
    , col.names = c(
      "data"
      , "TP predictions"
      , rep(c("ME","RMSE","MAPE"), times = 2)
    )
    , escape = F
    # , digits = 2
  ) %>% 
  kableExtra::kable_styling(font_size = 11.5) %>% 
  kableExtra::add_header_above(c(
    " "=2
    # , "Detection" = 3
    # , "Area" = 3
    , "Height (m)" = 3
    , "DBH (cm)" = 3
  )) %>% 
  kableExtra::column_spec(seq(2,8,by=3), border_right = TRUE, include_thead = TRUE) 
```

UAS-measured heights are highly accurate but those DBH measurements do not align very well. perhaps this is due to the fact that DBH wasn't actually measured in the field. instead, DRC (diameter at root crown) was measured and then we converted DRC to DBH using the Chojnacky & Rogers ([1999](https://doi.org/10.1093/wjaf/14.1.14)) model...therefore, it's probably best not to view these DBH metrics as measurements of "error" but as measurements of "comparison"

and plot predicted versus reference values for correct predictions (TP)

```{r}
####################################
# area scatter
####################################
dta_temp <-
  ground_truth_prediction_match_ans %>% 
  dplyr::filter(match_grp=="true positive") %>% 
  sf::st_drop_geometry() %>% 
  dplyr::bind_rows() %>% 
  dplyr::bind_cols(
    agg_ground_truth_match_ans %>% 
      dplyr::select(tidyselect::ends_with("_rmse") , tidyselect::ends_with("_mape"), tidyselect::ends_with("_mean") )
  ) %>% 
  dplyr::glimpse()
# plt fn temp
plt_fn_temp <- function(
    dta
    , which_col = "dbh_cm"
    , title_lab = "DBH"
    , x_lab = "reference DBH (cm)" # latex2exp::TeX("reference area ($m^2$)")
    , y_lab = "predicted DBH (cm)" # latex2exp::TeX("predicted area ($m^2$)")
){
    d <- dta
    # construct the formula: "y_var ~ x_var"
    y_var <- paste0("pred_",which_col)
    x_var <- paste0("ref_",which_col)
    formula_obj <- stats::as.formula(
      paste( 
        y_var
        , "~"
        , x_var 
      )
    )
    lm_temp <- lm(
      formula = formula_obj
      , data = d
    )
    # summary(lm_temp)
    # scales::percent(summary(lm_temp)$r.squared, accuracy = 0.1)
    lmf_temp <- paste0(
      "y = "
      , scales::number(summary(lm_temp)$coefficients[2,1], accuracy = 0.01) #slope
      , "x"
      , ifelse(summary(lm_temp)$coefficients[1,1]<0," - ", " + ")
      , scales::number(abs(summary(lm_temp)$coefficients[1,1]), accuracy = 0.01) #intercept
    )
    # scale limits
    max_val <- max( 
      max(dta_temp[[y_var]],na.rm=T)
      , max(dta_temp[[x_var]],na.rm=T)
    )
    
    # plot
    d %>% 
      ggplot2::ggplot(mapping = ggplot2::aes(y = .data[[y_var]], x = .data[[x_var]])) +
      ggplot2::geom_abline(color = "gray33", lwd = 1) +
      ggplot2::geom_point(color = "brown", alpha = 0.7, size = 0.9) +
      ggplot2::geom_smooth(lwd = 0.8, method = "lm", se=F, color = "gray", linetype = "dashed") +
      ggplot2::annotate(
        "text", 
        x = -Inf, y = Inf # Top Left
        , label = 
          # paste0(
          #   "R^2 == "
          #   , scales::number((summary(lm_temp)$r.squared)*100, accuracy = 0.1)
          #   , "*'%'"
          # )
          paste0(
            "RMSE = "
            , d %>% 
              dplyr::slice(1) %>% 
              dplyr::select(dplyr::all_of(paste0("diff_",which_col, "_rmse"))) %>% 
              dplyr::pull(1) %>% 
              unique() %>% 
              # scales::number(accuracy = 0.1, suffix = " (m²)")
              scales::number(
                accuracy = ifelse(which_col=="dbh_cm",0.1,0.01)
                , suffix = ifelse(which_col=="dbh_cm"," (cm)"," (m)") 
              )
            , "\nMAPE = "
            , d %>% 
              dplyr::select(dplyr::all_of(paste0("pct_diff_",which_col, "_mape"))) %>% 
              dplyr::pull(1) %>% 
              unique() %>% 
              scales::percent(accuracy = 0.1)
            , "\nME = "
            , d %>% 
              dplyr::select(dplyr::all_of(paste0("diff_",which_col, "_mean"))) %>% 
              dplyr::pull(1) %>% 
              unique() %>% 
              scales::number(
                accuracy = ifelse(which_col=="dbh_cm",0.1,0.01)
                , suffix = ifelse(which_col=="dbh_cm"," (cm)"," (m)") 
              )
            , "\n"
            , lmf_temp
            , "\n R² = "
            # , scales::number((summary(lm_temp)$r.squared)*100, accuracy = 0.1, suffix = "%")
            , scales::number((summary(lm_temp)$r.squared), accuracy = 0.01)
          )
        , hjust = -0.1, vjust = 1.1
        , parse = F
        , size = 3
      ) +
      # ggplot2::facet_wrap(
      #   facets = dplyr::vars(dataset_factor), scales = "free", axis.labels = "all"
      #   # , labeller = ggplot2::labeller(class = ggplot2::label_wrap_gen(width = 10))
      # ) +
      ggplot2::scale_x_continuous(limits = c(0, max_val ), expand = ggplot2::expansion(mult = c(0,0.1))) +
      ggplot2::scale_y_continuous(limits = c(0, max_val ), expand = ggplot2::expansion(mult = c(0,0.1))) +
      ggplot2::labs(
        x = x_lab
        , y = y_lab
        # , color = "image-field\ndiameter diff."
        # , subtitle = latex2exp::TeX("bulk volume ($m^3$) comparison")
        # , subtitle = paste("Predicted", title_lab,"versus Reference", title_lab)
      ) +
      ggplot2::theme_light() +
      ggplot2::theme(
        strip.text = ggplot2::element_text(size = 10, face = "bold", color = "black")
      ) 
  }
# get the plottttttttttttttttttt
dbh_plt_temp <-
  plt_fn_temp(dta_temp, which_col = "dbh_cm", title_lab = "DBH", x_lab = "reference DBH (cm)", y_lab = "predicted DBH (cm)")
# dbh_plt_temp

# get the plottttttttttttttttttt
ht_plt_temp <-
  plt_fn_temp(dta_temp, which_col = "tree_height_m", title_lab = "Height", x_lab = "reference Height (m)", y_lab = "predicted Height (m)")
ht_plt_temp
```

**DBH: predicted vs reference**

```{r}
dbh_plt_temp
```

yikes. see comments on DBH above

**height: predicted vs reference**

```{r}
ht_plt_temp
```

UAS-predicted height is highly accurate compared to field-measured values, a finding supported by many prior studies. 

```{r, warning=FALSE, message=FALSE, echo=FALSE, include=FALSE}
remove(list = ls()[grep("_temp",ls())])
gc()
```

full distribution of predicted and reference values irrespective of correct or incorrect prediction match

```{r}
ground_truth_prediction_match_ans %>%
  sf::st_drop_geometry() %>%
  dplyr::select(
    ref_tree_height_m, pred_tree_height_m
    , ref_dbh_cm, pred_dbh_cm
  ) %>% 
  tidyr::pivot_longer(
    cols = dplyr::everything()
    , names_to = "metric"
    , values_to = "value"
    , values_drop_na = T
  ) %>% 
  dplyr::mutate(
    which_data = dplyr::case_when(
        stringr::str_starts(metric,"field_") | stringr::str_starts(metric,"ref_") ~ "field"
        , stringr::str_starts(metric,"pred_") ~ "prediction"
        , T ~ "error"
      ) %>% 
      ordered()
    , tree_metric = metric %>% 
      stringr::str_remove("(_rmse|_rrmse|_mean|_mape)$") %>% 
      stringr::str_extract("(height|dbh)") %>% 
      factor(
        ordered = T
        , levels = c(
          "height"
          , "dbh"
        )
        , labels = c(
          "Height (m)"
          , "DBH (cm)"
        )
      )
  ) %>% 
  # dplyr::glimpse()
# plot dist
  ggplot2::ggplot(mapping = ggplot2::aes(x = value, color = which_data, fill = which_data)) +
  ggplot2::geom_density(mapping = ggplot2::aes(y=ggplot2::after_stat(scaled)), alpha = 0.7) +
  ggplot2::facet_grid(
    cols = dplyr::vars(tree_metric)
    , scales = "free_x", axes = "all_x"
    , switch = "y"
  ) +
  harrypotter::scale_color_hp_d(option = "hermionegranger") +
  harrypotter::scale_fill_hp_d(option = "hermionegranger") +
  ggplot2::scale_y_continuous(NULL,breaks=NULL) +
  ggplot2::scale_x_continuous(breaks=scales::breaks_extended(n=6)) +
  ggplot2::labs(
    color="",fill="",x=""
    # , subtitle = "comparison of height and DBH distributions"
  ) +
  ggplot2::theme_light() +
  ggplot2::theme(
    legend.position = "top"
    , strip.text.x = ggplot2::element_text(size = 9, color = "black", face = "bold")
    # , axis.text.x = ggplot2::element_text(size = 6)
    , panel.grid.major.y = ggplot2::element_blank()
    , panel.grid.minor.y = ggplot2::element_blank()
  )
```

let's shift our focus to relative error (MAPE) while providing absolute context (RMSE and ME) as text or secondary metrics to look into the penalty we face due to inaccuracies in DBH for small stems. for example, a 3 cm error on a 6 cm small tree is a 50% relative error while that same 3 cm error on a 30 cm merchantable tree is a minor 10% error.

let's look into the DBH prediction error over the range of tree sizes and use LOESS smoothing to highlight the trend in relative prediction error

```{r}
# Data trends (dashed) are visualized using a LOESS (locally weighted scatterplot smoothing) regression. 
# A span of 1.0 was utilized to include 100% of the observations in each local neighborhood, 
# and the local polynomial degree was restricted to 1 (local linear regression) to prevent overfitting and boundary inflation at the data extremes.

# The dashed line represents a locally estimated scatterplot smoothing (LOESS) trend line
# , calculated with a smoothing parameter (span) of 1.0 and a local linear polynomial fit (degree = 1).
ground_truth_prediction_match_ans %>%
  dplyr::mutate(
    # swap the error calc
    pct_diff_dbh_cm = (pred_dbh_cm-ref_dbh_cm)/ref_dbh_cm
  ) %>%
  sf::st_drop_geometry() %>%
  dplyr::filter(match_grp %in% c("true positive")) %>% 
  # dplyr::glimpse()
  ggplot2::ggplot(
    mapping = ggplot2::aes(y = pct_diff_dbh_cm, x = ref_dbh_cm)
  ) +
  ggplot2::geom_hline(yintercept = 0, color = "gray33", lwd = 1) +
  ggplot2::geom_point(color = "brown", alpha = 0.7, size = 0.7) +
  ggplot2::geom_smooth(
    method = "loess", se=F
    , span = 1 # 0.75 = default
    , method.args = list(degree = 1)
    , lwd = 1, color = "gray", linetype = "dashed"
  ) +
  ggplot2::annotate(
    "text"
    , x = Inf
    # , y = 0.045
    , y = Inf
    , label = "prediction > reference"
    , hjust = 1.1
    , vjust = 1.2
    , parse = F
    , size = 2.5
  ) +
  ggplot2::annotate(
    "text"
    , x = Inf
    # , y = -0.04
    , y = -Inf
    , label = "prediction < reference"
    , hjust = 1.1
    , vjust = -0.4
    , parse = F
    , size = 2.5
  ) +
  ggplot2::scale_x_continuous(breaks = scales::breaks_extended(n=8)) +
  ggplot2::scale_y_continuous(
    limits = c(-1,1)
    # breaks = seq(-1,1,by=0.2)
    , breaks = scales::breaks_extended(n = 8)
    , labels = scales::percent
  ) +
  ggplot2::labs(
    x = "field reference DBH (cm)", y = "relative DBH error"
    , subtitle = "relative DBH error (%) by reference tree DBH"
  ) +
  ggplot2::theme_light()
```

that's neat

do it for height too

```{r}
# Data trends (dashed) are visualized using a LOESS (locally weighted scatterplot smoothing) regression. 
# A span of 1.0 was utilized to include 100% of the observations in each local neighborhood, 
# and the local polynomial degree was restricted to 1 (local linear regression) to prevent overfitting and boundary inflation at the data extremes.

# The dashed line represents a locally estimated scatterplot smoothing (LOESS) trend line
# , calculated with a smoothing parameter (span) of 1.0 and a local linear polynomial fit (degree = 1).
ground_truth_prediction_match_ans %>%
  dplyr::mutate(
    # swap the error calc
    pct_diff_tree_height_m = (pred_tree_height_m-ref_tree_height_m)/ref_tree_height_m
  ) %>%
  sf::st_drop_geometry() %>%
  dplyr::filter(match_grp %in% c("true positive")) %>% 
  # dplyr::glimpse()
  ggplot2::ggplot(
    mapping = ggplot2::aes(y = pct_diff_tree_height_m, x = ref_tree_height_m)
  ) +
  ggplot2::geom_hline(yintercept = 0, color = "gray33", lwd = 1) +
  ggplot2::geom_point(color = "brown", alpha = 0.7, size = 0.7) +
  ggplot2::geom_smooth(
    method = "loess", se=F
    , span = 1 # 0.75 = default
    , method.args = list(degree = 1)
    , lwd = 1, color = "gray", linetype = "dashed"
  ) +
  ggplot2::annotate(
    "text"
    , x = Inf
    # , y = 0.045
    , y = Inf
    , label = "prediction > reference"
    , hjust = 1.1
    , vjust = 1.2
    , parse = F
    , size = 2.5
  ) +
  ggplot2::annotate(
    "text"
    , x = Inf
    # , y = -0.04
    , y = -Inf
    , label = "prediction < reference"
    , hjust = 1.1
    , vjust = -0.4
    , parse = F
    , size = 2.5
  ) +
  ggplot2::scale_x_continuous(breaks = scales::breaks_extended(n=8)) +
  ggplot2::scale_y_continuous(
    limits = c(-1,1)
    # breaks = seq(-1,1,by=0.2)
    , breaks = scales::breaks_extended(n = 8)
    , labels = scales::percent
  ) +
  ggplot2::labs(
    x = "field reference height (m)", y = "relative height error"
    , subtitle = "relative height error (%) by reference tree height"
  ) +
  ggplot2::theme_light()
```

that is also neat

## Stand-Level Accuracy Results{#stand_acc}

In addition to the tree-level accuracy metrics (e.g. height RMSE and MAPE), we can also calculate stand-level accuracy metrics by aggregating estimates from the reference (ground truth) and predicted trees irrespective of whether or not they were matched (i.e. true positive) during instance matching. 

*for this data, "stand-level" refers to the combined area covered by all field plots*

```{r}
pred_stand_temp <-
  treetops_sf %>% 
  sf::st_intersection(
    validation_plots %>% 
      sf::st_buffer(.$plot_radius_m) %>% 
      sf::st_union() %>% 
      sf::st_transform(terra::crs(treetops_sf)) %>% 
      sf::st_as_sf() %>% 
      dplyr::mutate(area_m2 = sf::st_area(.) %>% as.numeric())
  ) %>% 
  sf::st_drop_geometry() %>% 
  dplyr::group_by(area_m2) %>% 
  dplyr::summarise(
    n_trees = dplyr::n()
    , basal_area_m2 = sum(basal_area_m2)
  ) %>% 
  dplyr::ungroup() %>% 
  dplyr::mutate(
    area_ha = (area_m2/10000)
    , basal_area_m2_per_ha = basal_area_m2/area_ha
    , trees_per_ha = n_trees/area_ha
  ) %>% 
  dplyr::select(-area_m2)
# dplyr::glimpse(pred_stand_temp)
ref_stand_temp <-
  validation_trees %>% 
  sf::st_transform(terra::crs(treetops_sf)) %>% 
  sf::st_intersection(
    validation_plots %>% 
      sf::st_buffer(.$plot_radius_m) %>% 
      sf::st_union() %>% 
      sf::st_transform(terra::crs(treetops_sf)) %>% 
      sf::st_as_sf() %>% 
      dplyr::mutate(area_m2 = sf::st_area(.) %>% as.numeric())
  ) %>% 
  sf::st_drop_geometry() %>% 
  dplyr::mutate(basal_area_m2 = pi * dbh_cm^2 / (4*10000)) %>% 
  dplyr::group_by(area_m2) %>% 
  dplyr::summarise(
    n_trees = dplyr::n()
    , basal_area_m2 = sum(basal_area_m2)
  ) %>% 
  dplyr::ungroup() %>% 
  dplyr::mutate(
    area_ha = (area_m2/10000)
    , basal_area_m2_per_ha = basal_area_m2/area_ha
    , trees_per_ha = n_trees/area_ha
  ) %>% 
  dplyr::select(-area_m2)
# dplyr::glimpse(ref_stand_temp)
stand_temp <- 
  dplyr::inner_join(
    ref_stand_temp %>% dplyr::rename_with(.cols = -c(area_ha), ~paste0("ref_",.x,recycle0 = T))
    , pred_stand_temp %>% dplyr::rename_with(.cols = -c(area_ha), ~paste0("pred_",.x,recycle0 = T))
    , by = "area_ha"
  ) %>% 
  # comps not done in agg_ground_truth_match()
  dplyr::mutate(
    # 'diff_' columns are calculated as the predicted value minus the actual value
    diff_trees_per_ha = pred_trees_per_ha-ref_trees_per_ha
    , diff_basal_area_m2 = pred_basal_area_m2-ref_basal_area_m2
    , diff_basal_area_m2_per_ha = pred_basal_area_m2_per_ha-ref_basal_area_m2_per_ha
    # 'abs_diff_' columns are calculated as the predicted value minus the actual value
    , abs_diff_trees_per_ha = abs(diff_trees_per_ha)
    , abs_diff_basal_area_m2 = abs(diff_basal_area_m2)
    , abs_diff_basal_area_m2_per_ha = abs(diff_basal_area_m2_per_ha)
    # 'pct_diff_' columns are calculated as the predicted value minus the actual value divided by the actual value
    , pct_diff_trees_per_ha = (pred_trees_per_ha-ref_trees_per_ha)/ref_trees_per_ha
    , pct_diff_basal_area_m2 = (pred_basal_area_m2-ref_basal_area_m2)/ref_basal_area_m2
    , pct_diff_basal_area_m2_per_ha = (pred_basal_area_m2_per_ha-ref_basal_area_m2_per_ha)/ref_basal_area_m2_per_ha
  )
```

let's table the stand-level measurement error

```{r}
stand_temp %>% 
  dplyr::select(
    area_ha
    # detection
    , ref_n_trees
    , pred_n_trees
    # quantification
    , tidyselect::ends_with("_basal_area_m2_per_ha")
    , tidyselect::ends_with("_trees_per_ha")
  ) %>% 
  dplyr::select(!tidyselect::starts_with("abs_diff_")) %>% 
  dplyr::mutate(
    dplyr::across(
      .cols = c(
        tidyselect::starts_with("pct_diff_")
        , tidyselect::ends_with("_rate")
      )
      , .fn = ~ scales::percent(.x, accuracy = .1)
    )
    , dplyr::across(
      .cols = c(tidyselect::ends_with("_trees"))
      , .fn = ~ scales::comma(.x, accuracy = 1)
    )
    , dplyr::across(
      .cols = dplyr::where(is.numeric)
      , .fn = ~ scales::comma(.x, accuracy = 0.1)
    )
    # # one column for diff/pct
    , diff_basal_area_m2_per_ha = paste0(diff_basal_area_m2_per_ha, "<br>(", pct_diff_basal_area_m2_per_ha, ")")
    , diff_trees_per_ha = paste0(diff_trees_per_ha, "<br>(", pct_diff_trees_per_ha, ")")
  ) %>% 
  dplyr::select(
    !tidyselect::starts_with("pct_diff_")
  ) %>% 
  # dplyr::glimpse()
  
  kableExtra::kbl(
    caption = "Stand-Level Accuracy"
    , col.names = c(
      "field-measured area (ha)"
      , "reference", "predicted"
      , rep(c("reference","predicted","difference"), times = 2)
    )
    , escape = F
    # , digits = 2
  ) %>% 
  kableExtra::kable_styling(font_size = 11.5) %>% 
  kableExtra::add_header_above(
    c(
      " "=1
      , "# Trees" = 2
      # , "Area" = 3
      , "Basal Area<br />m<sup>2</sup> ha<sup>-1</sup>" = 3
      , "Trees<br />ha<sup>-1</sup>" = 3
    )
    , escape = F
  ) %>% 
  kableExtra::column_spec( c(1,seq(3,9,by=3) ), border_right = TRUE, include_thead = TRUE) %>% 
  kableExtra::collapse_rows(columns = c(1), valign = "top")
```

neat