# Tree Group Analysis

Evaluate the spatial configuration of the detected trees focusing on horizontal heterogeneity and an investigation of tree clumps and openings.

some relevant work includes:

* *Spatial Patterns of Pinyon–Juniper Woodland Expansion in Central Nevada* ([Weisberg et al. 2007](https://doi.org/10.2111/05-224R2.1))
* *Density-dependent tree mortality in pinyon-juniper woodlands* ([Greenwood and Weisberg 2008](https://doi.org/10.1016/j.foreco.2007.12.048))
* *Long-Term Effectiveness of Tree Removal to Re-Establish Sagebrush Steppe Vegetation and Associated Spatial Patterns in Surface Conditions and Soil Hydrologic Properties* ([Williams et al. 2020](https://doi.org/10.3390/w12082213))

```{r, warning=FALSE, message=FALSE, echo=FALSE, include=FALSE}
# when you get a chance i am wondering if we can do a little more evaluation of the PJ site data. Specifically i was thinking about some analysis of the clump structure so maybe something like a distribution of patch sizes, opening sizes  and maybe trees per patch distribution. also there was interest in some other spatial stats type  variables but these should be good for now.
remove(list = ls()[grep("_temp",ls())])
gc()
```

## Group Tree Crowns

While some research on tree spatial patterns use tree point locations buffered with a fixed radius (e.g. 3 m) to approximate the crown radius of a "typical" overstory tree
(e.g. [Churchill et al. (2016)](https://scholarworks.umt.edu/ico/3)), we'll use the irregular, more realistic crown shapes identified from the aerial data to determine tree clump groups.

Remember we are only looking at what could be considered "overstory" trees since we filtered the tree list to keep only trees that meet a certain minimum height threshold of `r min_tree_ht_m` m

```{r, echo=T}
treetops_sf %>% 
  sf::st_drop_geometry() %>% 
  dplyr::select(tree_height_m) %>% 
  summary() %>% 
  knitr::kable(format = "markdown")
```

Since a primary objective of this analysis is to evaluate how [different treatments](#silv_sum) influence forest structure, we'll treat each unit as a discrete area and exclude any trees located outside of the unit boundaries from the clump classification process. This approach assumes that trees within a unit do not form functional clusters with those in adjacent areas (even though some units share a common border and we have data extending beyond the unit boundaries). This assumption prevents data from untreated control areas or differently treated areas from influencing the results when summarized by treatment type. This intentional ignorance of edge effects means that the scale of focus of this analysis is strictly on the treatment unit.

define a function to make the tree clump groupings by treatment unit

```{r, eval = F, warning=FALSE, message=FALSE, echo=FALSE, include=FALSE}
c2t_output_dir <- "../data"
c2t_process_dir <- file.path(c2t_output_dir, "point_cloud_processing_delivery")
c2t_tracking_fnm <- file.path(c2t_process_dir, "processed_tracking_data.csv")
# fine-resolution CHM raster
chm_rast <- terra::rast( file.path(c2t_process_dir, "chm_0.1m.tif") )
stand_boundary <- 
   "c:/data/usfs/slash_pile_detecting/data/Dawson_Data/units/units.shp" %>% 
   # "d:/BLM_CO_SWDF_DawsonFuelsTreatment/units/units.shp" %>% 
   sf::st_read(quiet=T) %>% 
   dplyr::rename_with(tolower) %>% 
   dplyr::rename_with(stringr::str_squish) %>% 
   dplyr::rename_with(make.names) %>% 
   dplyr::rename_with(~stringr::str_replace_all(.x, "\\.{2,}", ".")) %>% 
   dplyr::rename_with(~stringr::str_remove(.x, "\\.$")) %>% 
   dplyr::rename_with(~stringr::str_replace_all(.x, "\\.", "_")) %>% 
   sf::st_set_geometry("geometry") %>% 
   sf::st_transform(terra::crs(chm_rast)) %>% 
   dplyr::mutate(
     unit_area_ha = as.numeric(sf::st_area(geometry))/10000
   )
# stand_boundary %>% dplyr::glimpse()
# where should we save the file?
treetops_fnm <- file.path(c2t_process_dir,"final_detected_tree_tops_attr.gpkg")
treetops_sf <- sf::st_read(dsn = treetops_fnm, quiet = TRUE)
# attach to treatment unit
treetops_sf <- treetops_sf %>% 
  dplyr::inner_join(
    treetops_sf %>% 
      sf::st_transform(sf::st_crs(stand_boundary)) %>% 
      sf::st_intersection(
        stand_boundary %>%
          dplyr::select(unit,block,tretmnt,unit_area_ha) %>% 
          dplyr::rename(type = tretmnt) %>% 
          dplyr::mutate(
            type =
              tolower(type) %>% 
              stringr::str_squish()
          ) %>% 
          dplyr::rename_with(.fn = ~paste0("trtmnt_", .x, recycle0 = T))
      ) %>% 
      sf::st_drop_geometry() %>% 
      dplyr::select(treeID, tidyselect::starts_with("trtmnt_")) %>% 
      dplyr::group_by(treeID) %>% 
      dplyr::slice(1) %>% 
      dplyr::ungroup()
    , by = "treeID"
  ) %>% 
  dplyr::mutate(
    is_in_trtmnt_unit = ifelse(is.na(trtmnt_type) | trtmnt_type=="control",F,T)
    , dplyr::across(c(trtmnt_type,trtmnt_block,trtmnt_unit),ordered)
  )
# sf::st_write(crowns_sf, dsn = "c:/users/georg/Downloads/crowns_sf.gpkg", append = F)
crowns_sf <- sf::st_read(dsn = "c:/users/georg/Downloads/crowns_sf.gpkg", quiet = TRUE) %>% 
  dplyr::select(-c(trtmnt_type,trtmnt_block,trtmnt_unit)) %>% 
  dplyr::inner_join(
    treetops_sf %>% 
      sf::st_drop_geometry() %>% 
      dplyr::select(treeID, trtmnt_type,trtmnt_block,trtmnt_unit)
    , by = "treeID"
  )
crowns_sf %>% dplyr::glimpse()
```

```{r, results=F}
get_clumps_temp <- function(tree_list, buffer = 0) {
  # buffer and dissolve (arcGIS logic)
  clump_polys <- tree_list %>%
    dplyr::ungroup() %>% 
    sf::st_buffer(buffer) %>% 
    sf::st_union() %>%
    sf::st_cast("POLYGON") %>%
    sf::st_as_sf() %>%
    dplyr::mutate(clump_id = dplyr::row_number())

  # attach to trees
  trees_with_clumps <- tree_list %>%
    # get rid of columns we'll create
    dplyr::select( -dplyr::any_of(c(
      "hey_xxxxxxxxxx"
      , "clump_id"
      , "clump_n_trees"
      , "clump_n_trees_grp"
      , "clump_dist_m"
    ))) %>%
    sf::st_join(clump_polys %>% dplyr::select(clump_id), join = sf::st_intersects, largest = T) %>%
    dplyr::group_by(clump_id) %>%
    dplyr::mutate(
      clump_n_trees = dplyr::n()
      , clump_n_trees_grp = dplyr::case_when(
        clump_n_trees == 1 ~ "Individual"
        , clump_n_trees >= 2 & clump_n_trees <= 4 ~ "2-4 trees"
        
        , clump_n_trees >= 5 & clump_n_trees <= 10 ~ "5-10 trees"
        , clump_n_trees > 10 ~ ">10 trees"
        # , clump_n_trees >= 5 & clump_n_trees <= 9 ~ "5-9 trees"
        # , clump_n_trees >= 10 & clump_n_trees <= 15 ~ "10-15 trees"
        # , clump_n_trees > 15 ~ ">15 trees"
        # # , T ~ "error"
      ) %>%
      factor(
        ordered = T
        , levels = c("Individual", "2-4 trees", "5-10 trees", ">10 trees")
        # , levels = c("Individual", "2-4 trees", "5-9 trees", "10-15 trees", ">15 trees")
      )
    ) %>%
    dplyr::ungroup()
  # add back to clumps
  clump_polys <- clump_polys %>% 
    dplyr::inner_join(
      trees_with_clumps %>% 
        sf::st_drop_geometry() %>% 
        dplyr::distinct(clump_id, clump_n_trees, clump_n_trees_grp)
      , by = "clump_id"
    )
  return(list(clump_polys = clump_polys, tree_list = trees_with_clumps))
}
# buffer for UAS-delineated crown polygons
my_buffer <- 1
```

make the clump groups using the UAS-delineated crown polygons. because we are using the irregular crown shapes that align with real-world crowns and want to account for slight misalignment between the UAS GPS and on-the-ground conditions we'll define clump group membership using a `r my_buffer/2` m buffer on the crowns. Under this condition, trees that do not have interlocking crowns based on UAS delineations but are within `r my_buffer` m of one another will be considered part of the same tree clump group. We'll use this same `r my_buffer` m distance threshold in defining treeless openings which will be required to comprise continuous space at least `r my_buffer` m away from the nearest tree.

```{r grp-them-crwns, results=F}
# crowns_sf %>% dplyr::glimpse()
# crowns_sf %>% sf::st_drop_geometry() %>% dplyr::count(trtmnt_type,trtmnt_block,trtmnt_unit)
# do it for all treatment units
# c2t_process_dir
crowns_clump_fnm <- file.path(c2t_process_dir, "crowns_clump.csv")
clump_polys_fnm <- file.path(c2t_process_dir, "crowns_clump_polys.gpkg")
if(
  !file.exists(crowns_clump_fnm) ||
  !file.exists(clump_polys_fnm)
){
  get_clumps_ans_temp <- 
    1:nrow(unq_units_df) %>% 
    purrr::map(function(x, buff = my_buffer/2){
      crowns_sf %>% 
        dplyr::inner_join(
          unq_units_df %>% dplyr::slice(x)
          , by = dplyr::join_by(trtmnt_type,trtmnt_block,trtmnt_unit)
        ) %>% 
        get_clumps_temp(buffer = buff)
    })
  # get trees
  crowns_sf <-
    get_clumps_ans_temp %>%
    purrr::map(
      \(x) purrr::pluck(x,"tree_list")
    ) %>%
    dplyr::bind_rows()
  # get clump polygons
  clump_polys <-
    1:nrow(unq_units_df) %>% 
    purrr::map(
      \(x)
      get_clumps_ans_temp[[x]] %>% 
        purrr::pluck("clump_polys") %>% 
        sf::st_set_geometry("geometry") %>% 
        dplyr::bind_cols(
          unq_units_df %>% dplyr::slice(x)
        )
    ) %>% 
    dplyr::bind_rows()
  
  # write
  crowns_sf %>% 
    sf::st_drop_geometry() %>% 
    dplyr::select(treeID, tidyselect::starts_with("clump_")) %>% 
    readr::write_csv(file = crowns_clump_fnm, append = F, progress = F)
  
  clump_polys %>% 
    sf::st_write(dsn = clump_polys_fnm, append = FALSE, quiet = TRUE)
}else{
  crowns_sf <- crowns_sf %>% 
    dplyr::select(-tidyselect::starts_with("clump_")) %>% 
    dplyr::inner_join(readr::read_csv(crowns_clump_fnm, progress = F), by = "treeID") %>% 
    dplyr::mutate(
      clump_n_trees_grp =  
        clump_n_trees_grp%>%
        factor(
          ordered = T
          , levels = c("Individual", "2-4 trees", "5-10 trees", ">10 trees")
          # , levels = c("Individual", "2-4 trees", "5-9 trees", "10-15 trees", ">15 trees")
        )
    )
  # clump_polys
  clump_polys <- sf::st_read(clump_polys_fnm, quiet = T) 
  clump_polys <- 
    clump_polys %>% 
    # convert to factors
    dplyr::mutate(
      dplyr::across(
        dplyr::all_of(
          names(clump_polys)[names(clump_polys) %in% names(unq_units_df)]
        )
        , ~ factor(
          .x
          , levels = levels(unq_units_df[[dplyr::cur_column()]])
          , ordered = T
        )
      )
      , dplyr::across(
        dplyr::all_of(c("clump_n_trees_grp"))
        , ~ factor(
          .x
          , levels = levels(crowns_sf[[dplyr::cur_column()]])
          , ordered = T
        )
      ) 
    )
}

```

let's look at what this clump definition did

```{r, include=F, eval = T}
get_huge_palette <- function(n_colors) {
  # 1. Establish the universally trusted, high-contrast Okabe-Ito base anchors
  # These are optimized for Protan, Deutan, and Tritan colorblindness
  okabe_ito <- c(
    "#0072B2", "#E69F00", "#009E73", "#56B4E9", 
    "#D55E00", "#CC79A7", "#F0E442", "#000000"
  )
  
  # Calculate remaining slots needed
  rem_colors <- base::max(0, n_colors - base::length(okabe_ito))
  colors_per_tier <- base::ceiling(rem_colors / 3)
  
  # 2. Build 3 hyper-contrasting HCL bands with maximized saturation
  # Band 1: Midnight/Deep tones (Low luminance, high chroma)
  tier_dark <- grDevices::hcl(
    h = base::seq(from = 0, to = 330, length.out = colors_per_tier), 
    c = 85, 
    l = 25
  )
  
  # Band 2: Electric/Vibrant tones (Medium luminance, max chroma)
  tier_medium <- grDevices::hcl(
    h = base::seq(from = 40, to = 370, length.out = colors_per_tier), 
    c = 95, 
    l = 55
  )
  
  # Band 3: Neon Pastel/Light tones (High luminance, moderate chroma)
  tier_light <- grDevices::hcl(
    h = base::seq(from = 20, to = 350, length.out = colors_per_tier), 
    c = 65, 
    l = 85
  )
  
  # 3. Interleave the three bands using tidyverse to enforce severe brightness jumps
  generated_colors <- tibble::tibble(
    dark   = tier_dark,
    medium = tier_medium,
    light  = tier_light
  ) %>%  
    tidyr::pivot_longer(
      cols = dplyr::everything(), 
      names_to = "tier", 
      values_to = "hex_color"
    ) %>%  
    dplyr::mutate(nr = dplyr::row_number()) %>% 
    # dplyr::slice_sample(n=n_colors*2) %>% 
    dplyr::slice_head(n=n_colors) %>%
    dplyr::arrange(nr) %>% 
    dplyr::pull(hex_color)
  
  # Combine anchors with generated colors
  raw_palette <- base::c(okabe_ito, generated_colors)[1:n_colors]
  
  # 4. Apply a wide mathematical dispersion stride
  # This breaks up the sequential sweeps entirely, packing contrasting tones side-by-side
  stride <- base::ceiling(base::sqrt(n_colors))
  indices <- base::seq(from = 1, to = n_colors, by = 1)
  dispersed_indices <- base::order((indices - 1) %% stride, (indices - 1) %/% stride)
  
  final_palette <- raw_palette[dispersed_indices]
  
  return(final_palette)
}

```

```{r}
dta_temp <- crowns_sf %>% 
  dplyr::inner_join(
    unq_units_df %>% dplyr::slice(3)
    , by = dplyr::join_by(trtmnt_type,trtmnt_block,trtmnt_unit)
  ) %>% 
  dplyr::mutate(clump_id = factor(clump_id))
# get_huge_palette(nrow(dta_temp)) %>% scales::show_col()
# get_huge_palette(8) %>% scales::show_col()
# get_huge_palette(22) %>% scales::show_col()
# plt
ggplot2::ggplot(data = dta_temp) + 
  # add clump outline
  ggplot2::geom_sf(
    data = 
      clump_polys %>% 
      dplyr::inner_join(
        dta_temp %>% sf::st_drop_geometry() %>% dplyr::distinct(trtmnt_type,trtmnt_block,trtmnt_unit)
        , by = dplyr::join_by(trtmnt_type,trtmnt_block,trtmnt_unit)
      )
    , fill = NA
    , color = "gray88"
  ) +
  # add crowns
  ggplot2::geom_sf(mapping = ggplot2::aes(fill = clump_id), color = "gray", alpha = 0.8) +
  ggplot2::scale_fill_manual(
    values = 
    get_huge_palette(nrow(dta_temp))
    # c(
    #   viridis::magma(n=nrow(dta_temp)/2, begin = 0, end = 0.85)
    #   , viridis::mako(n=nrow(dta_temp)/2, begin = 0, end = 0.85)
    #   , viridis::viridis(n=nrow(dta_temp)/2, begin = 0, end = 0.85)
    # ) %>% sample()
  ) +
  ggplot2::theme_void() +
  ggplot2::theme(legend.position = "none")
```

where each clump (light outline) is colored by a (nearly) distinct color

here it is with clumps (light outline) colored by clump group size

```{r}
hey_clump_pal <- 
  harrypotter::hp(n = length(levels(crowns_sf$clump_n_trees_grp)), option = "always") %>% 
  setNames(levels(crowns_sf$clump_n_trees_grp))
# scales::show_col(pal_clump_n_trees_grp)

ggplot2::ggplot(data = dta_temp) + 
  # add clump outline
  ggplot2::geom_sf(
    data = 
      clump_polys %>% 
      dplyr::inner_join(
        dta_temp %>% sf::st_drop_geometry() %>% dplyr::distinct(trtmnt_type,trtmnt_block,trtmnt_unit)
        , by = dplyr::join_by(trtmnt_type,trtmnt_block,trtmnt_unit)
      )
    , fill = NA
    , color = "gray88"
  ) +
  # add crowns
  ggplot2::geom_sf(mapping = ggplot2::aes(fill = clump_n_trees_grp), color = "gray") +
  ggplot2::scale_fill_manual(values = hey_clump_pal, name = "", drop = F) +
  ggplot2::labs(fill="") +
  ggplot2::theme_void() +
  ggplot2::theme(legend.position = "top", legend.direction = "horizontal") +
  ggplot2::guides(fill = ggplot2::guide_legend(override.aes = list(size = 5)))
```

and we have `clump_*` variables

```{r}
crowns_sf %>% 
  sf::st_drop_geometry() %>% 
  dplyr::select(treeID, tidyselect::starts_with("trtmnt_"), tidyselect::starts_with("clump_")) %>% 
  # dplyr::count(clump_n_trees_grp)
  dplyr::glimpse()
```

```{r, warning=FALSE, message=FALSE, echo=FALSE, include=FALSE}
# crowns_sf %>% 
#   sf::st_drop_geometry() %>% 
#   dplyr::count(treeID) %>% 
#   dplyr::filter(n>1) %>% 
#   dplyr::inner_join(
#     crowns_sf %>% 
#       sf::st_drop_geometry() %>% 
#       dplyr::select(treeID, tidyselect::starts_with("trtmnt_"), tidyselect::starts_with("clump_"))
#   ) %>% 
#   dplyr::arrange(treeID) %>% 
#   View()
remove(list = ls()[grep("_temp",ls())])
gc()
```

## Identify Openings

Let's identify the gaps within a treatment unit. We'll identify openings based on their distance from the nearest tree crown. We'll use this same `r my_buffer` m distance threshold in defining treeless openings, where each opening will be required to comprise continuous space at least `r my_buffer` m away from the nearest tree. This `r my_buffer` m buffer accounts for the zone of immediate canopy influence and ensures that areas categorized as open are functionally distinct from the canopy. This threshold is flexible and can be adjusted to meet specific management or analysis objectives (e.g. flame length; wildlife corridors). 

First, we'll generate a distance raster where each cell represents the distance to the nearest tree clump group (i.e. crown) within the treatment unit boundary. Areas with a distance value of `r my_buffer` m or greater are classified as openings and are defined by spatial connectedness, not a specific geometric shape. As such, they may be irregularly shaped as they are allowed to wind between clumps to follow a path where trees are distant. A single opening continues as long as the flow of open space is not entirely obstructed by surrounding tree groups. Once a gap is completely surrounded by canopy, the opening is closed off and is treated as a discrete spatial feature.

```{r}
# create function to pass a return from get_clump_summary() and get a distance raster
get_clump_dist_rast_temp <- function(x, clump_dist_m = 3){
  # get clump_dist_m
  # clump_dist_m <- min(get_clump_summary_ans$clump_dist_m, na.rm = T)
  # rasterize the clump polygons and then calculate distance between clumps as raster
  dist_rast <-
    terra::rasterize(
        x = 
          crowns_sf %>% 
          dplyr::inner_join(
            unq_units_df %>% dplyr::slice(x)
            , by = dplyr::join_by(trtmnt_type,trtmnt_block,trtmnt_unit)
          ) %>% 
          terra::vect()
        , y = 
          # crowns_sf %>% 
          # dplyr::inner_join(
          #   unq_units_df %>% dplyr::slice(x)
          #   , by = dplyr::join_by(trtmnt_type,trtmnt_block,trtmnt_unit)
          # ) %>% 
          # terra::vect() %>%
          stand_boundary %>% 
          dplyr::filter(
            unit == unq_units_df$trtmnt_unit[x]
            , block == unq_units_df$trtmnt_block[x]
            , stringr::str_squish(tolower(tretmnt)) == unq_units_df$trtmnt_type[x]
          ) %>% 
          terra::vect() %>% 
          terra::rast(res = 0.2)
      ) %>% 
      terra::distance() %>% 
      # crop it to stand extent
      terra::crop(
        stand_boundary %>% 
          dplyr::filter(
            unit == unq_units_df$trtmnt_unit[x]
            , block == unq_units_df$trtmnt_block[x]
            , stringr::str_squish(tolower(tretmnt)) == unq_units_df$trtmnt_type[x]
          ) %>% 
          terra::vect()
        , mask = T
      )
  ######### part 2
  # now create openings vector data
  openings_vect <- 
    dist_rast %>% 
      terra::classify(rcl = c(clump_dist_m,Inf), others = NA, include.lowest = T) %>% 
      terra::as.polygons(na.rm = T) %>% 
      sf::st_as_sf() %>% 
      sf::st_cast("POLYGON") %>% 
      dplyr::mutate(layer = dplyr::row_number()) %>% 
      dplyr::mutate(
        openining_area_m2 = sf::st_area(geometry) %>% as.numeric()
        , clump_dist_m = clump_dist_m
      )
  
  # return
  return(list(dist_rast = dist_rast, openings_vect = openings_vect))
}
```

let's get the distance raster and openings vector data

```{r}
# get it
get_clump_dist_rast_ans <- 
  1:nrow(unq_units_df) %>% 
  purrr::map(
    \(x)
    get_clump_dist_rast_temp(
      x, clump_dist_m = my_buffer
    )
  )
```

### Visualize Distance Raster

let's check out the distance raster for the same example unit we looked at above

```{r, results=F, fig.show='asis'}
# color the grps
plt_dist_rast_fn <- 
  function(x, add_clumps = F, max_dist = NA){
    plt_tit <- paste0(
      unq_units_df %>% dplyr::slice(x) %>% dplyr::pull(trtmnt_type)
      , " | block "
      , unq_units_df %>% dplyr::slice(x) %>% dplyr::pull(trtmnt_block)
      , " | "
      , unq_units_df %>% dplyr::slice(x) %>% dplyr::pull(trtmnt_unit)
    )
    
    # get around empty levels that causes problems in the legend of ggplot
    blank_df <- 
      dplyr::tibble(
        clump_n_trees_grp = levels(clump_polys$clump_n_trees_grp) %>% 
          ordered(levels = levels(clump_polys$clump_n_trees_grp))
        , geometry = rep("GEOMETRYCOLLECTION EMPTY", length(levels(clump_polys$clump_n_trees_grp))) %>% 
          sf::st_as_sfc(crs = sf::st_crs(clump_polys))
      ) %>% 
      sf::st_as_sf()
    
    # plt unit
    unit_temp <- stand_boundary %>% 
      dplyr::filter(
        unit == unq_units_df$trtmnt_unit[x]
        , block == unq_units_df$trtmnt_block[x]
        , stringr::str_squish(tolower(tretmnt)) == unq_units_df$trtmnt_type[x]
      ) %>% 
      sf::st_transform(terra::crs(chm_rast))
    
    plt_stand <- ggplot2::ggplot() +
      ggplot2::geom_sf(data = unit_temp, color = "black", fill = NA) +
      ggplot2::scale_x_continuous(expand = c(0, 0)) +
      ggplot2::scale_y_continuous(expand = c(0, 0)) +
      ggplot2::labs(
        x = ""
        , y = ""
        , subtitle = plt_tit
      ) +
      ggplot2::theme_void() +
      ggplot2::theme(
        legend.position = "top" # c(0.5,1)
        , legend.direction = "horizontal"
        , legend.box = "vertical"
        , legend.margin = ggplot2::margin(2.5,0,0,0) # trbl
        , legend.text = ggplot2::element_text(size = 7)
        , legend.title = ggplot2::element_text(size = 6)
        , legend.key = ggplot2::element_rect(fill = NA, color = NA)
        , legend.key.size = unit(0.5, "cm")
        # , plot.title = ggtext::element_markdown(size = 10, hjust = 0.5)
        , plot.title = ggplot2::element_text(size = 10, hjust = 0.5, face = "bold")
        , plot.subtitle = ggplot2::element_text(size = 9, hjust = 0.5, face = "bold")
      )
    # add raster
    plt_rast <- 
      plt_stand +
      ggplot2::geom_tile(
        data = get_clump_dist_rast_ans[[x]][["dist_rast"]] %>% 
          # terra::aggregate(fact = 2, fun = "max", cores = 4) %>% 
          # terra::crop(
          #   stand_boundaries %>% 
          #     dplyr::filter(poly_name==x)
          #     # dplyr::filter(poly_name=="west")
          #   , mask = T
          # ) %>% 
          terra::as.data.frame(xy = T) %>% 
          dplyr::rename(f=3)
        , mapping = ggplot2::aes(x=x, y=y, fill = f)
      )
    if(!is.na(max_dist)){
      plt_rast <- plt_rast +
        harrypotter::scale_fill_hp(
          "mischief"
          , na.value = "transparent"
          , name = "distance to\nnearest tree (m)"
          , limits = c(0,max_dist)
          , labels = scales::comma_format(accuracy = 1)
        )
    }else{
      plt_rast <- plt_rast +
        harrypotter::scale_fill_hp(
          "mischief"
          , na.value = "transparent"
          , name = "distance to\nnearest tree (m)"
          # , limits = c(0,max_dist)
          , labels = scales::comma_format(accuracy = 1)
        )
    }
    # add clumps
    if(add_clumps){
      plt_rast <- plt_rast +
        # # openings
        # ggplot2::geom_sf(
        #   data = get_clump_dist_rast_ans[[nm]]$openings_vect
        #   , mapping = ggplot2::aes(color = openining_area_m2), fill = NA
        # ) +
        # ggplot2::scale_color_gradient(
        #   low = "gray77", high = "gray11"
        #   , labels = scales::comma_format(accuracy = 1)
        #   , name = latex2exp::TeX("opening\narea ($\\m^2$)")
        #   , limits = c(0,openining_area_m2_max)
        # ) +
        # clumps
        ggnewscale::new_scale_fill() +
        ggplot2::geom_sf(
          data = clump_polys %>% 
            dplyr::inner_join(
              unq_units_df %>% dplyr::slice(x)
              , by = dplyr::join_by(trtmnt_type,trtmnt_block,trtmnt_unit)
            ) %>% 
            dplyr::select(clump_n_trees_grp)
          # , mapping = ggplot2::aes(fill = clump_n_trees_grp)
          , color = "gray33"
          , fill = NA
        ) +
        ggplot2::geom_sf(
          data = crowns_sf %>% 
            dplyr::inner_join(
              unq_units_df %>% dplyr::slice(x)
              , by = dplyr::join_by(trtmnt_type,trtmnt_block,trtmnt_unit)
            ) %>% 
            dplyr::select(clump_n_trees_grp) %>% 
            dplyr::bind_rows(blank_df)
          , mapping = ggplot2::aes(fill = clump_n_trees_grp)
          , color = NA
        ) +
        # harrypotter::scale_fill_hp_d("always", name = "clump size", drop = F) +
        ggplot2::scale_fill_manual(values = hey_clump_pal, name = "", drop = F) +
        # tree points
        ggplot2::geom_sf(
          data = 
            treetops_sf %>% 
            dplyr::inner_join(
              unq_units_df %>% dplyr::slice(x)
              , by = dplyr::join_by(trtmnt_type,trtmnt_block,trtmnt_unit)
            )
          , color = "gray88", shape = "."
        )
        # ggplot2::guides(fill = ggplot2::guide_legend(ncol = 2))
    }
    return(plt_rast)
  }
# one unit
plt_dist_rast_fn(3, add_clumps = F)
```

and here is the distance raster with the tree clump groups overlaid

```{r}
plt_dist_rast_fn(3, add_clumps = T)
```

```{r, include=F, eval=F}
###########!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!! terra::plot() version is fast but lacks easy customization (esp legend placement :\ )
# harrypotter::hp(n = length(levels(crowns_sf$clump_n_trees_grp)), option = "always") %>% scales::show_col()
plt_dist_rast_fn <- 
  function(x, add_crowns = F, add_clumps = F, add_treetops = F){
    plt <- paste0(
      unq_units_df %>% dplyr::slice(x) %>% dplyr::pull(trtmnt_type)
      , " | block "
      , unq_units_df %>% dplyr::slice(x) %>% dplyr::pull(trtmnt_block)
      , " | "
      , unq_units_df %>% dplyr::slice(x) %>% dplyr::pull(trtmnt_unit)
    )
    
    terra::plot(
      get_clump_dist_rast_ans[[x]][["dist_rast"]]
      , col = harrypotter::hp(n=100, option = "mischief")
      # , mar = c(b,l,t,r)
      , mar = c(0.2,0.2,2.2,3.5)
      , axes = F
      , plg = list(
        title = "distance to\nnearest tree (m)"
        , cex = 0.8, title.cex = 0.5
        , x = "right", shrink = 0.6
      )
      , main = plt
    ) 
    
    # add stand
    unit_temp <- stand_boundary %>% 
      dplyr::filter(
        unit == unq_units_df$trtmnt_unit[x]
        , block == unq_units_df$trtmnt_block[x]
        , stringr::str_squish(tolower(tretmnt)) == unq_units_df$trtmnt_type[x]
      ) %>% 
      sf::st_transform(terra::crs(get_clump_dist_rast_ans[[x]][["dist_rast"]])) %>% 
      terra::vect()
    terra::plot(
      unit_temp
      , border = "black", col = NA
       , add = T
    )
    
    if(add_crowns){
      terra::plot(
        crowns_sf %>% 
          dplyr::inner_join(
            unq_units_df %>% dplyr::slice(x)
            , by = dplyr::join_by(trtmnt_type,trtmnt_block,trtmnt_unit)
          ) %>% 
          terra::vect()
        , add = T, border = "cyan", col = NA, lwd = 0.7
      )
    }else if(add_clumps){
      # crowns_sf %>% 
      #   dplyr::inner_join(
      #     unq_units_df %>% dplyr::slice(x)
      #     , by = dplyr::join_by(trtmnt_type,trtmnt_block,trtmnt_unit)
      #   ) %>% 
      #   terra::vect() %>% 
      #   terra::plot(add = T, border = "cyan", col = NA, lwd = 0.7)
      clump_polys %>% 
        dplyr::inner_join(
          unq_units_df %>% dplyr::slice(x)
          , by = dplyr::join_by(trtmnt_type,trtmnt_block,trtmnt_unit)
        ) %>% 
        dplyr::select(clump_n_trees_grp) %>% 
        terra::vect() %>% 
        terra::plot(
          "clump_n_trees_grp"
          , col = rev(harrypotter::hp(n = length(levels(crowns_sf$clump_n_trees_grp)), option = "always"))
          , border = NA
          , axes = F
          , add = T
          , plg = list(
            x = "topright"
          )
        )
    }
    
    if(add_treetops){
      terra::plot(
        treetops_sf %>% 
          dplyr::inner_join(
            unq_units_df %>% dplyr::slice(x)
            , by = dplyr::join_by(trtmnt_type,trtmnt_block,trtmnt_unit)
          ) %>% 
          terra::vect()
        , add = T, col = "gray55", cex = 0.2
      )
    }
  }
# 1:nrow(unq_units_df) %>% 
#   .[3] %>% 
#   purrr::walk(plt_dist_rast_fn)
plt_dist_rast_fn(4)
```

let's zoom in on a single clump group and the surrounding area so we can see what is going on at the fine-scale level

```{r}
aoi_temp <- 
  clump_polys %>% 
  sf::st_intersection(
    clump_polys %>% 
      dplyr::inner_join(
        unq_units_df %>% dplyr::slice(3)
        , by = dplyr::join_by(trtmnt_type,trtmnt_block,trtmnt_unit)
      ) %>% 
      sf::st_union() %>% 
      sf::st_centroid() %>% 
      sf::st_buffer(sqrt(2500/4),endCapStyle = "SQUARE") ## numerator = desired plot size in m2) %>% 
  ) %>% 
  dplyr::arrange(
    desc(sf::st_area(.))
  ) %>% 
  dplyr::slice(2) %>% 
  sf::st_buffer(6.35) %>% 
  sf::st_bbox() %>% 
  sf::st_as_sfc()
# as.numeric(sf::st_area(aoi_temp))/10000
d_temp <- get_clump_dist_rast_ans[[3]][["dist_rast"]]
# dist rast
d_temp %>% 
   terra::crop(
    aoi_temp %>% 
      sf::st_transform(terra::crs(d_temp)) %>% 
      terra::vect()
    , mask=T
  ) %>% 
  terra::plot(
    col = harrypotter::hp(n=100,option="mischief")
    , axes = F
    , plg = list(
        title = "distance to\nnearest tree (m)"
        , cex = 0.8, title.cex = 0.5
        , x = "right", shrink = 0.6
      )
    , main = paste0("Example ", scales::comma(as.numeric(sf::st_area(aoi_temp))/10000, accuracy = 0.01), " ha area")
    # , loc.main = "topleft"
    , cex.main = 0.8
  )
# add clumps
clump_polys %>% 
  sf::st_intersection(aoi_temp) %>% 
  sf::st_transform(terra::crs(d_temp)) %>% 
  terra::vect() %>% 
  terra::plot(add = T, col = NA, border = "gray33", lwd = 1.3)
# add crowns
cr_temp <- 
  crowns_sf %>% 
  sf::st_intersection(aoi_temp)

cr_temp %>% 
  sf::st_transform(terra::crs(d_temp)) %>% 
  terra::vect() %>% 
  terra::plot(
    add = T, y = "clump_n_trees_grp"
    , col = hey_clump_pal[cr_temp$clump_n_trees_grp]
    , border = "gray88", lwd = 0.2
    , plg = list(
      x = "topright"
      , text.col = "gray88"
    )
  )
# add trees
treetops_sf %>% 
  sf::st_intersection(aoi_temp) %>% 
  sf::st_transform(terra::crs(d_temp)) %>% 
  terra::vect() %>% 
  terra::plot(add = T, col = "gray88", cex = 0.5)
```

we zoomed in on a `r scales::comma(as.numeric(sf::st_area(aoi_temp))/10000, accuracy = 0.01)` ha area and show the distance from the nearest tree crown with the tree crowns and tree top points overlaid and unique clump groups encircled. notice that even if the crowns are not directly touching, trees within `r my_buffer` m of one another are still grouped together in the same clump group.

now we'll overlay the tree clump groups on the distance raster for all units

```{r,eval=F,purl=FALSE, results=F, fig.show='asis', fig.height=9.5, fig.width=9.5}
# mapping
plt_blk_temp <- function(blk, add = F, save = F) {
  # which idx
  idx_temp <- 
    unq_units_df %>% 
    dplyr::ungroup() %>% 
    dplyr::mutate(rownum = dplyr::row_number()) %>% 
    dplyr::filter(trtmnt_block==blk) %>% 
    dplyr::pull(rownum)
  # max dist
  max_d <- 
    purrr::map(get_clump_dist_rast_ans[idx_temp], "dist_rast") %>% 
    purrr::map(
      \(r)
      terra::minmax(r) %>% 
        .[2] %>% 
        ceiling()  
    ) %>% 
    purrr::compact() %>% 
    unlist() %>% 
    max(na.rm = T)
  # plts
  plts_temp <- 
    idx_temp %>% 
    purrr::map(\(x) plt_dist_rast_fn(x, add_clumps = add, max_dist = max_d))
  # patchwork
  p_temp <- 
    patchwork::wrap_plots(
      plts_temp
      , ncol = 2
      , guides = "collect"
    ) &
    ggplot2::theme(legend.position = "bottom")
  
  # save 
  if(save){
    ggplot2::ggsave(
      filename = file.path("../data", paste0("plt_dist_block_", blk, ".png"))
      , plot = p_temp
      , width = 10
      , height = 10
      , dpi = 300
    )
  }else{
    p_temp
  }
}
# map over blks
unique(unq_units_df$trtmnt_block) %>% 
  # .[1:2] %>%
  purrr::map(\(x) plt_blk_temp(blk=x, add = T, save = F))
```

```{r, warning=FALSE, message=FALSE, echo=FALSE, include=FALSE}
remove(list = ls()[grep("_temp",ls())])
gc()
```

### Visualize Openings

let's check out the openings for the same example unit we looked at above

```{r, results=F, fig.show='asis'}
plt_openings_vect_fn <- 
  function(x, add_clumps = F, max_area = NA){
    plt_tit <- paste0(
      unq_units_df %>% dplyr::slice(x) %>% dplyr::pull(trtmnt_type)
      , " | block "
      , unq_units_df %>% dplyr::slice(x) %>% dplyr::pull(trtmnt_block)
      , " | "
      , unq_units_df %>% dplyr::slice(x) %>% dplyr::pull(trtmnt_unit)
    )
    
    # get around empty levels that causes problems in the legend of ggplot
    blank_df <- 
      dplyr::tibble(
        clump_n_trees_grp = levels(clump_polys$clump_n_trees_grp) %>% 
          ordered(levels = levels(clump_polys$clump_n_trees_grp))
        , geometry = rep("GEOMETRYCOLLECTION EMPTY", length(levels(clump_polys$clump_n_trees_grp))) %>% 
          sf::st_as_sfc(crs = sf::st_crs(clump_polys))
      ) %>% 
      sf::st_as_sf()

    
    # plt unit
    unit_temp <- stand_boundary %>% 
      dplyr::filter(
        unit == unq_units_df$trtmnt_unit[x]
        , block == unq_units_df$trtmnt_block[x]
        , stringr::str_squish(tolower(tretmnt)) == unq_units_df$trtmnt_type[x]
      ) %>% 
      sf::st_transform(terra::crs(chm_rast))
    
    plt_stand <- ggplot2::ggplot() +
      ggplot2::geom_sf(data = unit_temp, color = "black", fill = NA) +
      ggplot2::scale_x_continuous(expand = c(0, 0)) +
      ggplot2::scale_y_continuous(expand = c(0, 0)) +
      ggplot2::labs(
        x = ""
        , y = ""
        , subtitle = plt_tit
      ) +
      ggplot2::theme_void() +
      ggplot2::theme(
        legend.position = "top" # c(0.5,1)
        , legend.direction = "horizontal"
        , legend.box = "vertical"
        , legend.margin = ggplot2::margin(2.5,0,0,0) # trbl
        , legend.text = ggplot2::element_text(size = 7)
        , legend.title = ggplot2::element_text(size = 6)
        , legend.key = ggplot2::element_rect(fill = NA, color = NA)
        , legend.key.size = unit(0.5, "cm")
        # , plot.title = ggtext::element_markdown(size = 10, hjust = 0.5)
        , plot.title = ggplot2::element_text(size = 10, hjust = 0.5, face = "bold")
        , plot.subtitle = ggplot2::element_text(size = 9, hjust = 0.5, face = "bold")
      )
    # add open
    plt_open <- 
      plt_stand +
      # openings
      ggplot2::geom_sf(
        data = get_clump_dist_rast_ans[[x]][["openings_vect"]]
        , mapping = ggplot2::aes(fill = openining_area_m2), color = NA
      )
      # ggplot2::guides(
      #   fill = ggplot2::guide_colorbar(
      #     label.theme = ggplot2::element_text(
      #       angle = 45, hjust = 1.1, vjust = 1.4, size = 8
      #     )
      #   )
      # )
      
    if(!is.na(max_area)){
      plt_open <- plt_open +
        ggplot2::scale_fill_gradient(
          low = "gray77", high = "gray11"
          , labels = scales::comma_format(accuracy = 1)
          , name = "opening\narea (m²)" #latex2exp::TeX("opening\narea ($\\m^2$)")
          , limits = c(0,max_area)
          , guide = ggplot2::guide_colorbar(
            label.theme = ggplot2::element_text(
              angle = 55
              , hjust = 1
              , vjust = 1.3
            )
          )
        )
    }else{
      plt_open <- plt_open +
        ggplot2::scale_fill_gradient(
          low = "gray77", high = "gray11"
          , labels = scales::comma_format(accuracy = 1)
          , name = "opening\narea (m²)" #latex2exp::TeX("opening\narea ($\\m^2$)")
          # , limits = c(0,max_area)
          , guide = ggplot2::guide_colorbar(
            label.theme = ggplot2::element_text(
              angle = 55
              , hjust = 1
              , vjust = 1.3
            )
          )
        )
    }
    # add clumps
    if(add_clumps){
      plt_open <- plt_open +
        # clumps
        ggnewscale::new_scale_fill() +
        ggplot2::geom_sf(
          data = clump_polys %>% 
            dplyr::inner_join(
              unq_units_df %>% dplyr::slice(x)
              , by = dplyr::join_by(trtmnt_type,trtmnt_block,trtmnt_unit)
            ) %>% 
            dplyr::select(clump_n_trees_grp)
          # , mapping = ggplot2::aes(fill = clump_n_trees_grp)
          , color = "gray33"
          , fill = NA
        ) +
        ggplot2::geom_sf(
          data = crowns_sf %>% 
            dplyr::inner_join(
              unq_units_df %>% dplyr::slice(x)
              , by = dplyr::join_by(trtmnt_type,trtmnt_block,trtmnt_unit)
            ) %>% 
            dplyr::select(clump_n_trees_grp) %>% 
            dplyr::bind_rows(blank_df)
          , mapping = ggplot2::aes(fill = clump_n_trees_grp)
          , color = NA
        ) +
        # harrypotter::scale_fill_hp_d("always", name = "clump size", drop = F) +
        ggplot2::scale_fill_manual(values = hey_clump_pal, name = "", drop = F, limits = levels(crowns_sf$clump_n_trees_grp)) +
        # tree points
        ggplot2::geom_sf(
          data = 
            treetops_sf %>% 
            dplyr::inner_join(
              unq_units_df %>% dplyr::slice(x)
              , by = dplyr::join_by(trtmnt_type,trtmnt_block,trtmnt_unit)
            )
          , color = "gray88", shape = "."
        )
        # ggplot2::guides(fill = ggplot2::guide_legend(ncol = 2))
    }
    return(plt_open)
  }
# one unit
plt_openings_vect_fn(3, add_clumps = F)
```

and here are the openings with the tree clump groups overlaid

```{r}
plt_openings_vect_fn(3, add_clumps = T)
```

now we'll overlay the tree clump groups on the openings data for all units

```{r,eval=F,purl=FALSE, results=F, fig.show='asis', fig.height=9.5, fig.width=9.5}
# mapping
plt_blk_temp <- function(blk, add = F, save = F) {
  # which idx
  idx_temp <- 
    unq_units_df %>% 
    dplyr::ungroup() %>% 
    dplyr::mutate(rownum = dplyr::row_number()) %>% 
    dplyr::filter(trtmnt_block==blk) %>% 
    dplyr::pull(rownum)
  # max dist
  max_d <-
    purrr::map(get_clump_dist_rast_ans[idx_temp], "openings_vect") %>%
    dplyr::bind_rows() %>% 
    dplyr::pull(openining_area_m2) %>% 
    max(na.rm = T) %>% 
    ceiling()
  # plts
  plts_temp <- 
    idx_temp %>% 
    purrr::map(\(x) plt_openings_vect_fn(x, add_clumps = add, max_area = max_d))
  # patchwork
  p_temp <- 
    patchwork::wrap_plots(
      plts_temp
      , ncol = 2
      , guides = "collect"
    ) &
    ggplot2::theme(legend.position = "bottom")
  
  # save 
  if(save){
    ggplot2::ggsave(
      filename = file.path("../data", paste0("plt_open_block_", blk, ".png"))
      , plot = p_temp
      , width = 10
      , height = 10
      , dpi = 300
    )
  }else{
    p_temp
  }
}
# map over blks
unique(unq_units_df$trtmnt_block) %>% 
  # .[1:2] %>%
  purrr::map(\(x) plt_blk_temp(blk=x, add = T, save = F))
```

```{r, warning=FALSE, message=FALSE, echo=FALSE, include=FALSE}
remove(list = ls()[grep("_temp",ls())])
gc()
```

## Tree Clump Summary

let's get some summary stats of the tree clump groupings by treatment type

```{r}
# create a function to summarize by number of tree clump grouping
get_clump_n_trees_grp_summary_temp = function(x){
  # get area of harvest unit
  #...will use this area in the area calculations such that...
  #...TPA = trees in a certain group size across the whole stand area
  harvest_area_ha <- 
    stand_boundary %>% 
    dplyr::filter(
      unit == unq_units_df$trtmnt_unit[x]
      , block == unq_units_df$trtmnt_block[x]
      , stringr::str_squish(tolower(tretmnt)) == unq_units_df$trtmnt_type[x]
    ) %>% 
    sf::st_transform(terra::crs(chm_rast)) %>% 
    sf::st_union() %>% 
    sf::st_as_sf() %>% 
    dplyr::mutate(harvest_area_ha = sf::st_area(.) %>% as.numeric() %>% `/`(10000)) %>% 
    dplyr::pull(harvest_area_ha) %>% 
    .[1]
  # which dta
  get_tree_clumps_ans <- 
    crowns_sf %>% 
    sf::st_drop_geometry() %>% 
    dplyr::ungroup() %>% 
    dplyr::inner_join(
      unq_units_df %>% dplyr::slice(x)
      , by = dplyr::join_by(trtmnt_type,trtmnt_block,trtmnt_unit)
    ) %>% 
    dplyr::mutate(stand_area_ha = harvest_area_ha)
  # collapse and calculate silv metrics
  dta <- 
    get_tree_clumps_ans %>% 
    dplyr::group_by(trtmnt_type,trtmnt_block,trtmnt_unit,stand_area_ha,clump_n_trees_grp) %>%
    dplyr::summarise(
      # summary metrics
      n_trees = dplyr::n_distinct(treeID)
      , mean_dbh_cm = mean(dbh_cm, na.rm = T)
      , mean_tree_height_m = mean(tree_height_m, na.rm = T)
      , mean_crown_diameter_m = mean(crown_diameter_m, na.rm = T)
      # , loreys_height_m = sum(basal_area_m2*tree_height_m, na.rm = T) / sum(basal_area_m2, na.rm = T)
      # , basal_area_m2 = sum(basal_area_m2, na.rm = T)
      # , sum_dbh_cm_sq = sum(dbh_cm^2, na.rm = T)
      , .groups = "drop_last"
    ) %>%
    dplyr::ungroup() %>% 
    dplyr::mutate(
      trees_per_ha = (n_trees/stand_area_ha) # (n_trees/clump_area_ha) ... this was not right
      # , basal_area_m2_per_ha = (basal_area_m2/stand_area_ha) # (basal_area_m2/clump_area_ha) ... this was not right
      # , qmd_cm = sqrt(sum_dbh_cm_sq/n_trees)
      # stand calcs
      , stand_trees_per_ha = sum(n_trees)/stand_area_ha
      # , stand_basal_area_m2 = sum(basal_area_m2)
      # , stand_basal_area_m2_per_ha = sum(basal_area_m2)/stand_area_ha
      # , pct_stand_basal_area = basal_area_m2/stand_basal_area_m2
      , pct_stand_n_trees = n_trees/sum(n_trees)
      # , stand_qmd_cm = sqrt(sum(get_tree_clumps_ans$dbh_cm^2, na.rm = T)/sum(n_trees))
    ) %>%
    # dplyr::select(-c(sum_dbh_cm_sq)) %>% 
    # convert to imperial units
    calc_imperial_units_fn()
  # fill in missing
  ret_dta <- 
    dplyr::tibble(
      # get all possible labels from the factor levels
      clump_n_trees_grp = levels(crowns_sf$clump_n_trees_grp)
    ) %>% 
    dplyr::mutate(
      dplyr::across(
        dplyr::all_of(c("clump_n_trees_grp"))
        , ~ factor(
          .x
          , levels = levels(crowns_sf[[dplyr::cur_column()]])
          , ordered = T
        )
      )
    ) %>% 
    dplyr::bind_cols(unq_units_df %>% dplyr::slice(x)) %>% 
    dplyr::left_join(
      dta
      , by = dplyr::join_by(trtmnt_type,trtmnt_block,trtmnt_unit,clump_n_trees_grp)
    ) %>% 
    dplyr::mutate(
      dplyr::across(
        tidyselect::starts_with("stand_")
        , ~ max(.x, na.rm = T)
      )
      , dplyr::across(
        dplyr::where(is.numeric)
        , ~ dplyr::coalesce(.x, 0)
      )
    )
  # return
  return(ret_dta)
}
```

what does this do?

```{r}
get_clump_n_trees_grp_summary_temp(3) %>% dplyr::glimpse()
```

let's map over it for each treatment unit

```{r}
clump_n_trees_grp_summary <-
  1:nrow(unq_units_df) %>% 
  purrr::map(get_clump_n_trees_grp_summary_temp) %>% 
  dplyr::bind_rows()
```

and table it

```{r}
# table it
clump_n_trees_grp_summary %>% 
  dplyr::select(
    trtmnt_block,trtmnt_type,trtmnt_unit,clump_n_trees_grp
    , n_trees
    # , stand_trees_per_ha
    # , mean_dbh_cm
    # , qmd_cm
    , mean_tree_height_m
    , mean_crown_diameter_m
    # , loreys_height_m
    , trees_per_ha
    # , basal_area_ft2_per_ac, pct_stand_basal_area
    , pct_stand_n_trees
  ) %>% 
  dplyr::mutate(
    dplyr::across(
      .cols = c(n_trees)
      , .fns = ~ scales::comma(.x, accuracy = 1)
    )
    , dplyr::across(
      .cols = tidyselect::starts_with("pct_")
      , .fns = ~ scales::percent(.x, accuracy = 1)
    )
  ) %>% 
  # dplyr::glimpse()
  # dplyr::slice_head(n=8) %>% 
  kableExtra::kbl(
    digits = 1
    , escape = F
    , caption = paste0("tree clump group summary by treatment unit")
    , col.names = c(
       "block","type","unit"
      , "clump size", "trees"
      # , "mean<br>DBH (in)"
      # , "QMD (in)"
      , "mean<br>Ht. (m)"
      # , "Loreys<br>Ht. (m)"
      , "mean<br>Crown diam. (m)"
      # , "TPA"
      , "trees ha<sup>-1</sup>"
      # , "BA<br>ft<sup>2</sup> ac<sup>-1</sup>"
      # , "Pct<br>stand BA"
      , "Pct<br>stand trees"
      )
  ) %>% 
  kableExtra::kable_styling(font_size = 9.7) %>% 
  kableExtra::collapse_rows(columns = 1:3, valign = "top") %>% 
  kableExtra::scroll_box(height = "8.4in")
```

that is so much information but some people like tables and they might be able to copy-and-paste the data for additional big data deep dives using AI machine learning data mining statistical and graphical analysis

let's summarize the data by treatment type and plot it for those that prefer visualizations

```{r}
clump_n_trees_grp_summary %>% 
  dplyr::group_by(trtmnt_type, clump_n_trees_grp) %>% 
  dplyr::summarise(n = sum(n_trees)) %>% 
  dplyr::ungroup() %>% 
  # area by type
  dplyr::inner_join(
    unit_sum_df %>% 
      dplyr::group_by(trtmnt_type) %>% 
      dplyr::summarise(trtmnt_unit_area_ha=sum(trtmnt_unit_area_ha,na.rm=T), .groups = "drop_last")
    , by = "trtmnt_type"
  ) %>% 
  calc_imperial_units_fn() %>% # get acres
  dplyr::group_by(trtmnt_type) %>% 
  dplyr::mutate(
    pct = n/sum(n)
    # , pct_lab = scales::percent(pct,accuracy=0.1)
    , tpa = n/trtmnt_unit_area_ac
    , tph = n/trtmnt_unit_area_ha
    , lab = paste0(scales::comma(tph,accuracy=1), "\n",scales::percent(pct,accuracy=1) )
  ) %>% 
  dplyr::ungroup() %>% 
  # dplyr::glimpse()
  ggplot2::ggplot(
    mapping = ggplot2::aes(y = tph, x = clump_n_trees_grp)
  ) +
  ggplot2::geom_col(
    mapping = ggplot2::aes(fill = trtmnt_type)
    , width = 0.6
    , color = NA, alpha = 0.8
  ) +
  ggplot2::geom_text(
    mapping = ggplot2::aes(label = lab)
    , color = "black", size = 2.5, vjust = -0.2
  ) +
  ggplot2::scale_fill_brewer(palette = "Dark2") +
  ggplot2::scale_x_discrete(drop = F) + 
  ggplot2::scale_y_continuous(
    # breaks = seq(0,1,by=0.2)
    labels = scales::comma
    , expand = ggplot2::expansion(mult = c(0,0.1))
  ) +
  ggplot2::facet_grid(
    cols = dplyr::vars(trtmnt_type)
    # , rows = dplyr::vars(trtmnt_block)
    # , ncol = 2
    , axes = "all"
    # , switch = "y"
  ) +
  ggplot2::labs(
    x = "tree clump size", y = "trees ha⁻¹", fill = ""
    , subtitle = "Trees per hectare by tree clump size"
  ) +
  ggplot2::theme_light() +
  ggplot2::theme(
    legend.position = "none"
    , strip.text = ggplot2::element_text(size = 11, color = "black", face = "bold")
    , axis.text.x = ggplot2::element_text(size = 9, angle = 45, hjust = 1, vjust = 1)
    # , axis.text.x = ggplot2::element_text(size = 12)
    # , axis.text.y = ggplot2::element_blank()
    # , axis.ticks.y = ggplot2::element_blank()
  )
```

the plot above represents an average across all of the smaller treatment units by treatment type. averaging across the different treatment units results in a loss of the underlying distribution. let's look at the range of outcomes by treatment type

```{r}
clump_n_trees_grp_summary %>% 
  dplyr::ungroup() %>% 
  # dplyr::glimpse()
  ggplot2::ggplot(
    mapping = ggplot2::aes(y = trees_per_ha, x = clump_n_trees_grp)
  ) +
  ggplot2::geom_boxplot(
    mapping = ggplot2::aes(fill = trtmnt_type)
    , width = 0.6
    , alpha = 0.8
    , outliers = F
  ) +
  # ggplot2::geom_text(
  #   mapping = ggplot2::aes(label = lab)
  #   , color = "black", size = 2.5, vjust = -0.2
  # ) +
  ggplot2::scale_fill_brewer(palette = "Dark2") +
  ggplot2::scale_x_discrete(drop = F) + 
  ggplot2::scale_y_continuous(
    # breaks = seq(0,1,by=0.2)
    labels = scales::comma
    , expand = ggplot2::expansion(mult = c(0,0.1))
  ) +
  ggplot2::facet_grid(
    cols = dplyr::vars(trtmnt_type)
    # , rows = dplyr::vars(trtmnt_block)
    # , ncol = 2
    , axes = "all"
    # , switch = "y"
  ) +
  ggplot2::labs(
    x = "tree clump size", y = "trees ha⁻¹", fill = ""
    , subtitle = "Distribution of trees per hectare by tree clump size"
  ) +
  ggplot2::theme_light() +
  ggplot2::theme(
    legend.position = "none"
    , strip.text = ggplot2::element_text(size = 11, color = "black", face = "bold")
    , axis.text.x = ggplot2::element_text(size = 9, angle = 45, hjust = 1, vjust = 1)
    # , axis.text.x = ggplot2::element_text(size = 12)
    # , axis.text.y = ggplot2::element_blank()
    # , axis.ticks.y = ggplot2::element_blank()
  )
```

this is neat, let's check out the same type of plot but look at the proportion of unit-level trees within the different clump sizes

```{r}
clump_n_trees_grp_summary %>% 
  dplyr::ungroup() %>% 
  # dplyr::glimpse()
  ggplot2::ggplot(
    mapping = ggplot2::aes(y = pct_stand_n_trees, x = clump_n_trees_grp)
  ) +
  ggplot2::geom_boxplot(
    mapping = ggplot2::aes(fill = trtmnt_type)
    , width = 0.6
    , alpha = 0.8
    , outliers = F
  ) +
  # ggplot2::geom_text(
  #   mapping = ggplot2::aes(label = lab)
  #   , color = "black", size = 2.5, vjust = -0.2
  # ) +
  ggplot2::scale_fill_brewer(palette = "Dark2") +
  ggplot2::scale_x_discrete(drop = F) + 
  ggplot2::scale_y_continuous(
    limits = c(0,0.9)
    , breaks = seq(0,1,by=0.2)
    , labels = scales::percent
  ) +
  ggplot2::facet_grid(
    cols = dplyr::vars(trtmnt_type)
    # , rows = dplyr::vars(trtmnt_block)
    # , ncol = 2
    , axes = "all"
    # , switch = "y"
  ) +
  ggplot2::labs(
    x = "tree clump size", y = "proportion of unit trees", fill = ""
    , subtitle = "Distribution of unit-level tree proportions by tree clump size"
  ) +
  ggplot2::theme_light() +
  ggplot2::theme(
    legend.position = "none"
    , strip.text = ggplot2::element_text(size = 11, color = "black", face = "bold")
    , axis.text.x = ggplot2::element_text(size = 9, angle = 45, hjust = 1, vjust = 1)
    # , axis.text.x = ggplot2::element_text(size = 12)
    # , axis.text.y = ggplot2::element_blank()
    # , axis.ticks.y = ggplot2::element_blank()
  )
```

we can aggregate those proportions over all units in a given treatment type

```{r}
clump_n_trees_grp_summary %>% 
  dplyr::group_by(trtmnt_type, clump_n_trees_grp) %>% 
  dplyr::summarise(n = sum(n_trees)) %>% 
  dplyr::group_by(trtmnt_type) %>% 
  dplyr::mutate(
    pct = n/sum(n)
  ) %>% 
  dplyr::ungroup() %>% 
  # dplyr::glimpse()
  ggplot2::ggplot(
    mapping = ggplot2::aes(y = pct, x = clump_n_trees_grp)
  ) +
  ggplot2::geom_col(
    mapping = ggplot2::aes(fill = trtmnt_type)
    , width = 0.6
    , color = NA, alpha = 0.8
  ) +
  ggplot2::geom_text(
    mapping = ggplot2::aes(label = scales::percent(pct,accuracy = 0.1))
    , color = "black", size = 2.5, vjust = -0.2
  ) +
  ggplot2::scale_fill_brewer(palette = "Dark2") +
  ggplot2::scale_x_discrete(drop = F) + 
  ggplot2::scale_y_continuous(
    limits = c(0,0.9)
    , breaks = seq(0,1,by=0.2)
    , labels = scales::percent
  ) +
  ggplot2::facet_grid(
    cols = dplyr::vars(trtmnt_type)
    # , rows = dplyr::vars(trtmnt_block)
    # , ncol = 2
    , axes = "all"
    # , switch = "y"
  ) +
  ggplot2::labs(
    x = "tree clump size", y = "proportion of trees¹", fill = ""
    , subtitle = "Study-wide tree proportions by tree clump size"
  ) +
  ggplot2::theme_light() +
  ggplot2::theme(
    legend.position = "none"
    , strip.text = ggplot2::element_text(size = 11, color = "black", face = "bold")
    , axis.text.x = ggplot2::element_text(size = 9, angle = 45, hjust = 1, vjust = 1)
    # , axis.text.x = ggplot2::element_text(size = 12)
    # , axis.text.y = ggplot2::element_blank()
    # , axis.ticks.y = ggplot2::element_blank()
  )
```

we can look at that same data in a slightly different way

```{r}
clump_n_trees_grp_summary %>% 
  dplyr::group_by(trtmnt_type, clump_n_trees_grp) %>% 
  dplyr::summarise(n = sum(n_trees)) %>% 
  dplyr::group_by(trtmnt_type) %>% 
  dplyr::mutate(
    pct = n/sum(n)
  ) %>% 
  dplyr::ungroup() %>% 
  dplyr::select(trtmnt_type, clump_n_trees_grp, pct) %>% 
  dplyr::mutate(
    pct_lab = scales::percent(ifelse(pct>=0.08,pct,NA), accuracy = 1)
    , trtmnt_type = forcats::fct_rev(trtmnt_type)
    , clump_n_trees_grp = forcats::fct_rev(clump_n_trees_grp)
    , huh = ifelse(
      clump_n_trees_grp %in% c(">10 trees","5-10 trees")
      , "white"
      , "black"
    )
  ) %>% 
ggplot2::ggplot(
  mapping = ggplot2::aes(x = pct, y = trtmnt_type, fill=clump_n_trees_grp, group=clump_n_trees_grp)
) +
  ggplot2::geom_col(
    width = 0.7, alpha=0.95
  ) +
  ggplot2::geom_text(
    mapping = aes(
        label = pct_lab
        , color = huh
        , fontface = "bold"
      )
    , position = ggplot2::position_stack(vjust = 0.5)
    # , color = "black"
    , size = 4
  ) +
  ggplot2::scale_fill_manual(values = hey_clump_pal) +
  ggplot2::scale_color_identity() +
  ggplot2::scale_x_continuous(labels = scales::percent_format()) +
  ggplot2::labs(
    fill = "clump\nsize"
    , y = ""
    , x = "proportion of trees in different sized clumps"
    , subtitle = "Study-wide tree proportions by tree clump size"
  ) +
  ggplot2::theme_light() +
  ggplot2::theme(
    legend.position = "top"
    , legend.direction  = "horizontal"
    , legend.title = element_text(size=7)
    , axis.title.x = element_text(size=10, face = "bold")
    , axis.text.x = element_blank()
    , axis.ticks.x = element_blank()
    , axis.text.y = ggplot2::element_text(size = 9, face = "bold")
  ) +
  ggplot2::guides(
    fill = ggplot2::guide_legend(
      # nrow = 2, byrow = T, 
      reverse = T
      # , override.aes = list(alpha = 0.9)
    )
  )
```

```{r, eval=F, include=F}
# *DBH* 
crowns_sf %>% 
  sf::st_drop_geometry() %>% 
  dplyr::ungroup() %>%
  dplyr::select(trtmnt_type, clump_n_trees_grp, dbh_cm) %>% 
  calc_imperial_units_fn() %>% 
  # dplyr::glimpse()
  ggplot2::ggplot(
    mapping = ggplot2::aes(y = dbh_in, x = clump_n_trees_grp)
  ) +
  ggplot2::geom_boxplot(
    mapping = ggplot2::aes(fill = trtmnt_type)
    , width = 0.6
    , alpha = 0.8
    , outliers = F
  ) +
  # ggplot2::geom_text(
  #   mapping = ggplot2::aes(label = lab)
  #   , color = "black", size = 2.5, vjust = -0.2
  # ) +
  ggplot2::scale_fill_brewer(palette = "Dark2") +
  ggplot2::scale_x_discrete(drop = F) + 
  ggplot2::scale_y_continuous(
    # breaks = seq(0,1,by=0.2)
    labels = scales::comma
    , expand = ggplot2::expansion(mult = c(0,0.1))
  ) +
  ggplot2::facet_grid(
    cols = dplyr::vars(trtmnt_type)
    # , rows = dplyr::vars(trtmnt_block)
    # , ncol = 2
    , axes = "all"
    # , switch = "y"
  ) +
  ggplot2::labs(
    x = "tree clump size", y = "DBH (in)", fill = ""
    , subtitle = "Distribution of DBH by tree clump size"
  ) +
  ggplot2::theme_light() +
  ggplot2::theme(
    legend.position = "none"
    , strip.text = ggplot2::element_text(size = 11, color = "black", face = "bold")
    , axis.text.x = ggplot2::element_text(size = 9, angle = 45, hjust = 1, vjust = 1)
    # , axis.text.x = ggplot2::element_text(size = 12)
    # , axis.text.y = ggplot2::element_blank()
    # , axis.ticks.y = ggplot2::element_blank()
  )
```

finally, let's check out the distribution of tree sizes within the different clump size groups by treatment type

```{r}
crowns_sf %>% 
  sf::st_drop_geometry() %>% 
  dplyr::ungroup() %>%
  dplyr::select(trtmnt_type, clump_n_trees_grp, tree_height_m) %>% 
  # calc_imperial_units_fn() %>% 
  # dplyr::glimpse()
  ggplot2::ggplot(
    mapping = ggplot2::aes(y = tree_height_m, x = clump_n_trees_grp)
  ) +
  ggplot2::geom_boxplot(
    mapping = ggplot2::aes(fill = trtmnt_type)
    , width = 0.6
    , alpha = 0.8
    , outliers = F
  ) +
  # ggplot2::geom_text(
  #   mapping = ggplot2::aes(label = lab)
  #   , color = "black", size = 2.5, vjust = -0.2
  # ) +
  ggplot2::scale_fill_brewer(palette = "Dark2") +
  ggplot2::scale_x_discrete(drop = F) + 
  ggplot2::scale_y_continuous(
    # breaks = seq(0,1,by=0.2)
    labels = scales::comma
    , expand = ggplot2::expansion(mult = c(0,0.1))
  ) +
  ggplot2::facet_grid(
    cols = dplyr::vars(trtmnt_type)
    # , rows = dplyr::vars(trtmnt_block)
    # , ncol = 2
    , axes = "all"
    # , switch = "y"
  ) +
  ggplot2::labs(
    x = "tree clump size", y = "tree height (m)", fill = ""
    , subtitle = "Distribution of tree height by tree clump size"
  ) +
  ggplot2::theme_light() +
  ggplot2::theme(
    legend.position = "none"
    , strip.text = ggplot2::element_text(size = 11, color = "black", face = "bold")
    , axis.text.x = ggplot2::element_text(size = 9, angle = 45, hjust = 1, vjust = 1)
    # , axis.text.x = ggplot2::element_text(size = 12)
    # , axis.text.y = ggplot2::element_blank()
    # , axis.ticks.y = ggplot2::element_blank()
  )
```

for the most part, those are very wide distributions but interesting to compare across treatment types

## Openings Summary

lastly, we'll look at the openings information. 

we define openings as contiguous spatial areas that originate `r my_buffer` m away from the nearest tree crown. Although these openings may form highly sinuous (“wavy”) corridors winding through tree clump groups, they are closed-off wherever the flow of open space is entirely obstructed by surrounding tree groups. Once a gap is completely surrounded by canopy, the opening is closed off and is treated as a discrete spatial feature. We also require that an opening must have an area greater than 40.5 square meters (0.01 acres) to be considered an actual opening.

```{r}
my_min_openining_area_m2 <- 0.01*4047 # 1/100 of an acre where 4047=sq m in 1 acre

open_full_temp <- 
  get_clump_dist_rast_ans %>% 
  purrr::map(purrr::pluck("openings_vect")) %>%
  purrr::list_rbind(names_to = "row_num") %>% 
  sf::st_drop_geometry() %>% 
  dplyr::inner_join(unq_units_df %>% dplyr::mutate(row_num = dplyr::row_number()), by = "row_num") %>% 
  dplyr::filter(
    openining_area_m2 > my_min_openining_area_m2
  )
# dplyr::glimpse(open_full_temp)
# stand_boundary %>% nrow()
# stand_boundary %>% names()
# unit_sum_df %>% dplyr::glimpse()
cc_df_temp <- 
  1:nrow(unit_sum_df) %>% 
  purrr::map(function(x){
    # crown area
    area_m2_temp <- 
      crowns_sf %>% 
      dplyr::ungroup() %>% 
      dplyr::select(treeID) %>% 
      sf::st_transform(sf::st_crs(stand_boundary)) %>% 
      # crop crowns to treatment units
      sf::st_intersection(
        stand_boundary %>%
          dplyr::select(unit,block,tretmnt) %>% 
          dplyr::rename(type = tretmnt) %>% 
          dplyr::mutate(
            type =
              tolower(type) %>% 
              stringr::str_squish()
          ) %>% 
          dplyr::rename_with(.fn = ~paste0("trtmnt_", .x, recycle0 = T)) %>% 
          dplyr::inner_join(
            unit_sum_df %>% dplyr::slice(x) %>% 
              dplyr::select(trtmnt_type,trtmnt_block,trtmnt_unit)
            , by = dplyr::join_by(trtmnt_type,trtmnt_block,trtmnt_unit)
          )
      ) %>% 
      # union by unit to dissolve internal boundaries
      sf::st_union() %>%
      sf::st_area() %>% 
      as.numeric()
    
    ret_df <- 
      unit_sum_df %>% 
      dplyr::slice(x) %>% 
      dplyr::select(trtmnt_type,trtmnt_block,trtmnt_unit,trtmnt_unit_area_ha) %>% 
      dplyr::mutate(
        canopy_cov_area_m2 = dplyr::coalesce(area_m2_temp,0)
        , canopy_cov_area_ha = (canopy_cov_area_m2/10000)
        , canopy_cov_pct = canopy_cov_area_ha/trtmnt_unit_area_ha
      )
    return(ret_df)
  }) %>% 
  purrr::list_rbind()
# cc_df_temp %>% dplyr::glimpse()

# sum qualified openinings
open_temp <- 
  open_full_temp %>% 
  dplyr::group_by(trtmnt_type,trtmnt_block,trtmnt_unit) %>% 
  dplyr::summarise(
    mean_openining_area_m2 = mean(openining_area_m2, na.rm = T)
    , median_openining_area_m2 = median(openining_area_m2, na.rm = T)
    , sum_openining_area_m2 = sum(openining_area_m2, na.rm = T)
    , .groups = "drop_last"
  ) %>% 
  dplyr::ungroup() %>% 
  dplyr::inner_join(
    unit_sum_df %>% 
      dplyr::select(trtmnt_unit_area_ha,trtmnt_type,trtmnt_block,trtmnt_unit)
    , by = dplyr::join_by(trtmnt_type,trtmnt_block,trtmnt_unit)
  ) %>% 
  dplyr::mutate(
    sum_openining_area_ha = sum_openining_area_m2/10000
    , pct_open = sum_openining_area_ha/trtmnt_unit_area_ha
  ) %>% 
  # get canopy cover as well
  dplyr::left_join(
    cc_df_temp %>% dplyr::select(-c(trtmnt_unit_area_ha))
    , by = dplyr::join_by(trtmnt_type,trtmnt_block,trtmnt_unit)
  )
  
# dplyr::glimpse(open_temp)
```

the proportion of the unit covered by treeless openings for each treatment unit

```{r}
open_temp %>% 
  dplyr::mutate(
    my_fct = forcats::fct_cross(trtmnt_type,trtmnt_block,trtmnt_unit)
  ) %>% 
  dplyr::arrange(desc(trtmnt_type),pct_open) %>% 
  dplyr::mutate(my_fct = forcats::fct_inorder(my_fct)) %>% 
  ggplot2::ggplot(mapping = ggplot2::aes(x=pct_open,y=my_fct,fill=trtmnt_type,group = trtmnt_type)) +
  ggplot2::geom_col() +
  ggplot2::scale_fill_brewer(palette = "Dark2") +
  ggplot2::scale_x_continuous(labels = scales::percent) +
  ggplot2::theme_light() +
  ggplot2::labs(x = "% unit open", y = "") +
  ggplot2::theme(
    legend.position = "none"
    , axis.text.y = ggplot2::element_text(size = 7)
    , axis.text.x = ggplot2::element_text(size = 8)
  )
```

based on these unit-level data, we can look at the distribution of the proportion of the stand that is open

```{r}
open_temp %>% 
  ggplot2::ggplot(mapping = ggplot2::aes(x=pct_open)) +
  ggplot2::geom_violin(
    mapping = ggplot2::aes(y = 0, fill=trtmnt_type)
    , alpha = 0.8, color = NA
  ) +
  ggplot2::geom_boxplot(width = 0.1, fill = NA, outliers = F) +
  ggplot2::facet_grid(
    rows = dplyr::vars(trtmnt_type)
    , axes = "all"
    , switch = "y"
  ) +
  ggplot2::scale_fill_brewer(palette = "Dark2") +
  ggplot2::scale_x_continuous(limits=c(0,NA),labels = scales::percent, breaks = scales::breaks_extended(n=6)) +
  ggplot2::scale_y_continuous(NULL,breaks=NULL) +
  ggplot2::labs(
    fill="",x="% unit open",y=""
    , title = "distribution of treatment unit open area"
    , subtitle = "\nby treatment type"
  ) +
  ggplot2::theme_light() +
  ggplot2::theme(
    legend.position = "none"
    , strip.text = ggplot2::element_text(size = 10, color = "black", face = "bold")
    , axis.text.x = ggplot2::element_text(size = 9)
    , plot.subtitle = ggplot2::element_text(margin = ggplot2::margin(t = -20, unit = "pt"))
  )
```

this is what we expected based on the treatment type prescriptions.

let's also look at the distribution of canopy cover based on that UAS-delineated tree crowns in these different treatments

```{r}
open_temp %>% 
  ggplot2::ggplot(mapping = ggplot2::aes(x=canopy_cov_pct)) +
  ggplot2::geom_violin(
    mapping = ggplot2::aes(y = 0, fill=trtmnt_type)
    , alpha = 0.8, color = NA
  ) +
  ggplot2::geom_boxplot(width = 0.1, fill = NA, outliers = F) +
  ggplot2::facet_grid(
    rows = dplyr::vars(trtmnt_type)
    , axes = "all"
    , switch = "y"
  ) +
  ggplot2::scale_fill_brewer(palette = "Dark2") +
  ggplot2::scale_x_continuous(limits=c(0,NA),labels = scales::percent, breaks = scales::breaks_extended(n=6)) +
  ggplot2::scale_y_continuous(NULL,breaks=NULL) +
  ggplot2::labs(
    fill="",x="% canopy cover",y=""
    , title = "distribution of treatment canopy cover area"
    , subtitle = "\nby treatment type"
  ) +
  ggplot2::theme_light() +
  ggplot2::theme(
    legend.position = "none"
    , strip.text = ggplot2::element_text(size = 10, color = "black", face = "bold")
    , axis.text.x = ggplot2::element_text(size = 9)
    , plot.subtitle = ggplot2::element_text(margin = ggplot2::margin(t = -20, unit = "pt"))
  )
```

let's look at the proportion of canopy cover and open area together for each treatment unit. note, that the total of canopy cover and opening area will not comprise the entire unit area with the remaining space accounted for in the form of intermediate required spacing of `r my_buffer` m between tree crowns and open space as well as very small enclosed, openings with an area less than 40.5 square meters (0.01 acres) which was required for classification as an actual treeless opening

```{r}
open_temp %>% 
  ggplot2::ggplot(mapping = ggplot2::aes(x=pct_open,y=canopy_cov_pct,color=trtmnt_type)) +
  ggplot2::geom_function(
    fun = function(x) 1 - x
    , color = "gray"
    # , lwd = 1
  ) +
  ggplot2::geom_point(size = 4, alpha = 0.9) +
  ggplot2::scale_color_brewer(palette = "Dark2") +
  ggplot2::scale_x_continuous(limits = c(0, 1), breaks = seq(0,1,by=0.2), labels = scales::percent) +
  ggplot2::scale_y_continuous(limits = c(0, 1), breaks = seq(0,1,by=0.2), labels = scales::percent) +
  ggplot2::theme_light() +
  ggplot2::labs(
    x = "% unit open"
    , y = "% canopy cover"
    , color = ""
  ) +
  ggplot2::theme(
    legend.position = "top"
    , axis.text.y = ggplot2::element_text(size = 8)
    , axis.text.x = ggplot2::element_text(size = 8)
  )

```

the variability within the different treatment types is interesting

let's table this by unit (which is the underlying data used to generate the distribution immediately above)

```{r}
open_temp %>% 
  dplyr::select(
    trtmnt_type,trtmnt_block,trtmnt_unit
    , trtmnt_unit_area_ha
    , sum_openining_area_ha, pct_open
    , canopy_cov_area_ha, canopy_cov_pct
  ) %>% 
  dplyr::mutate(
    dplyr::across(tidyselect::ends_with("area_ha"), ~scales::comma(.x,accuracy=0.01))
    , dplyr::across(c(pct_open,canopy_cov_pct), ~scales::percent(.x,accuracy=0.1))
  ) %>% 
  dplyr::arrange(trtmnt_type, trtmnt_block, trtmnt_unit) %>% 
  kableExtra::kbl(
    caption = "openings by treatment unit"
    , col.names = c(
      "type","block","unit","ha"
      , "open ha", "% open"
      , "canopy cover ha", "% canopy cover"
    )
    , escape = F
    # , digits = 2
  ) %>% 
  kableExtra::kable_styling(font_size = 10.6) %>% 
  kableExtra::collapse_rows(columns = 1, valign = "top") %>% 
  kableExtra::scroll_box(height = "8.4in")
```

and table the summary of open space by treatment type

```{r}
# open_temp$pct_open %>% summary()
open_temp %>% 
  dplyr::select(
    trtmnt_type
    , trtmnt_unit_area_ha
    , pct_open
  ) %>% 
  dplyr::group_by(trtmnt_type) %>% 
  dplyr::summarise(
    dplyr::across(
      c(pct_open)
      , .fns = list(
        q25 = ~quantile(.x,na.rm=T,probs=0.25)
        , median = ~median(.x,na.rm=T)
        , mean = ~mean(.x,na.rm=T)
        , q75 = ~quantile(.x,na.rm=T,probs=0.75)
        # , q10 = ~quantile(.x,na.rm=T,probs=0.1)
        # , q50 = ~quantile(.x,na.rm=T,probs=0.5)
        # , q90 = ~quantile(.x,na.rm=T,probs=0.9)
        # , min = ~min(.x,na.rm=T)
        # , max = ~max(.x,na.rm=T)
        , range = ~paste0(
          scales::percent(min(.x,na.rm=T), accuracy = 0.1)
          ,"-"
          , scales::percent(max(.x,na.rm=T), accuracy = 0.1)
        )
      )
    )
    , n = dplyr::n()
    , trtmnt_unit_area_ha = mean(trtmnt_unit_area_ha, na.rm=T)
  ) %>% 
  dplyr::ungroup() %>% 
  dplyr::mutate(
    n = scales::comma(n,accuracy=1)
    , trtmnt_unit_area_ha = scales::comma(trtmnt_unit_area_ha,accuracy=.1)
    , dplyr::across(.cols = dplyr::where(is.numeric), ~scales::percent(.x,accuracy=0.1) )
  ) %>% 
  dplyr::arrange(trtmnt_type) %>% 
  dplyr::relocate(trtmnt_type,n,trtmnt_unit_area_ha) %>% 
  kableExtra::kbl(
    caption = "openings by treatment type"
    , col.names = c(
      "type","# units","mean unit area (ha)"
      , "1st Qu."
      , "median"
      , "mean"
      , "3rd Qu."
      , "range"
    )
    , escape = F
    # , digits = 2
  ) %>% 
  kableExtra::kable_styling(font_size = 10.6) %>% 
  kableExtra::add_header_above(c(
    " "=3
    # , "Detection" = 3
    # , "Area" = 3
    , "% unit open" = 5
  )) %>% 
  kableExtra::column_spec(c(3,8), border_right = TRUE, include_thead = TRUE) 
```

let's check out the raw size areas of the individual openings even though there is variation in the size of the treatment units which influences how large an opening can be

```{r}
open_full_temp %>% 
  ggplot2::ggplot(mapping = ggplot2::aes(x=openining_area_m2)) +
  ggplot2::geom_violin(
    mapping = ggplot2::aes(y = 0, fill=trtmnt_type)
    , alpha = 0.8, color = NA
  ) +
  ggplot2::geom_boxplot(width = 0.1, fill = NA, outliers = F) +
  ggplot2::facet_grid(
    rows = dplyr::vars(trtmnt_type)
    , axes = "all"
    , switch = "y"
  ) +
  ggplot2::scale_fill_brewer(palette = "Dark2") +
  ggplot2::scale_x_continuous(limits=c(0,NA),labels = scales::comma) +
  ggplot2::scale_y_continuous(NULL,breaks=NULL) +
  ggplot2::labs(
    fill="",x="opening\narea (m²)",y=""
    , title = "distribution of treatment unit open area"
    , subtitle = "\nby treatment type"
  ) +
  ggplot2::theme_light() +
  ggplot2::theme(
    legend.position = "none"
    , strip.text = ggplot2::element_text(size = 10, color = "black", face = "bold")
    , axis.text.x = ggplot2::element_text(size = 9)
    , plot.subtitle = ggplot2::element_text(margin = ggplot2::margin(t = -20, unit = "pt"))
  )
```

wow. that may not be very informative...let's use a log scale on the x-axis

```{r}
open_full_temp %>% 
  ggplot2::ggplot(mapping = ggplot2::aes(x=openining_area_m2)) +
  ggplot2::geom_violin(
    mapping = ggplot2::aes(y = 0, fill=trtmnt_type)
    , alpha = 0.8, color = NA
  ) +
  ggplot2::geom_boxplot(width = 0.1, fill = NA, outliers = F) +
  ggplot2::facet_grid(
    rows = dplyr::vars(trtmnt_type)
    , axes = "all"
    , switch = "y"
  ) +
  ggplot2::scale_fill_brewer(palette = "Dark2") +
  ggplot2::scale_x_log10(labels = scales::comma) +
  ggplot2::scale_y_continuous(NULL,breaks=NULL) +
  ggplot2::labs(
    fill="",x="opening\narea (m²)",y=""
    , title = "distribution of treeless openings area"
    , subtitle = "\nby treatment type"
    , caption = "* note the log scale of the x-axis"
  ) +
  ggplot2::theme_light() +
  ggplot2::theme(
    legend.position = "none"
    , strip.text = ggplot2::element_text(size = 10, color = "black", face = "bold")
    , axis.text.x = ggplot2::element_text(size = 9)
    , plot.subtitle = ggplot2::element_text(margin = ggplot2::margin(t = -20, unit = "pt"))
  )
```

table the summary statistics of the opening sizes

```{r}
# open_temp$pct_open %>% summary()
open_full_temp %>% 
  dplyr::select(
    trtmnt_type
    , openining_area_m2
  ) %>% 
  dplyr::group_by(trtmnt_type) %>% 
  dplyr::summarise(
    dplyr::across(
      c(openining_area_m2)
      , .fns = list(
        q25 = ~quantile(.x,na.rm=T,probs=0.25)
        , median = ~median(.x,na.rm=T)
        , mean = ~mean(.x,na.rm=T)
        , q75 = ~quantile(.x,na.rm=T,probs=0.75)
        # , q10 = ~quantile(.x,na.rm=T,probs=0.1)
        # , q50 = ~quantile(.x,na.rm=T,probs=0.5)
        # , q90 = ~quantile(.x,na.rm=T,probs=0.9)
        # , min = ~min(.x,na.rm=T)
        # , max = ~max(.x,na.rm=T)
        , range = ~paste0(
          scales::comma(min(.x,na.rm=T), accuracy = 0.1)
          ,"-"
          , scales::comma(max(.x,na.rm=T), accuracy = 0.1)
        )
      )
    )
    , n = dplyr::n()
    # , trtmnt_unit_area_ha = mean(trtmnt_unit_area_ha, na.rm=T)
  ) %>% 
  dplyr::ungroup() %>% 
  dplyr::mutate(
    n = scales::comma(n,accuracy=1)
    , dplyr::across(.cols = dplyr::where(is.numeric), ~scales::comma(.x,accuracy=0.1) )
  ) %>% 
  dplyr::arrange(trtmnt_type) %>% 
  dplyr::relocate(trtmnt_type,n) %>% 
  kableExtra::kbl(
    caption = "treeless opening sizes by treatment type"
    , col.names = c(
      "type","# openings"
      , "1st Qu."
      , "median"
      , "mean"
      , "3rd Qu."
      , "range"
    )
    , escape = F
    # , digits = 2
  ) %>% 
  kableExtra::kable_styling(font_size = 10.6) %>% 
  kableExtra::add_header_above(
    c(
      " "=2
      # , "Detection" = 3
      # , "Area" = 3
      , "opening area (m<sup>2</sup>)" = 5
    )
    , escape = F
  ) %>% 
  kableExtra::column_spec(c(2,7), border_right = TRUE, include_thead = TRUE) 
```


