# Tree Group Analysis

Evaluate the spatial configuration of the detected trees focusing on horizontal heterogeneity and an investigation of tree clumps and openings.

some relevant work includes:

* *Spatial Patterns of Pinyon–Juniper Woodland Expansion in Central Nevada* ([Weisberg et al. 2007](https://doi.org/10.2111/05-224R2.1))
* *Density-dependent tree mortality in pinyon-juniper woodlands* ([Greenwood and Weisberg 2008](https://doi.org/10.1016/j.foreco.2007.12.048))
* *Long-Term Effectiveness of Tree Removal to Re-Establish Sagebrush Steppe Vegetation and Associated Spatial Patterns in Surface Conditions and Soil Hydrologic Properties* ([Williams et al. 2020](https://doi.org/10.3390/w12082213))

```{r, warning=FALSE, message=FALSE, echo=FALSE, include=FALSE}
# when you get a chance i am wondering if we can do a little more evaluation of the PJ site data. Specifically i was thinking about some analysis of the clump structure so maybe something like a distribution of patch sizes, opening sizes  and maybe trees per patch distribution. also there was interest in some other spatial stats type  variables but these should be good for now.
remove(list = ls()[grep("_temp",ls())])
gc()
```

## Group Tree Crowns

While some research on tree spatial patterns use tree point locations buffered with a fixed radius (e.g. 3 m) to approximate the crown radius of a "typical" overstory tree
(e.g. [Churchill et al. (2016)](https://scholarworks.umt.edu/ico/3)), we'll use the irregular (more realistic?) crown shapes identified from the aerial data to determine tree clump groups.

Remember we are only looking at what could be considered "overstory" trees since we filtered the tree list to keep only trees that meet a certain minimum height

```{r}
crowns_sf %>% 
  sf::st_drop_geometry() %>% 
  dplyr::select(tree_height_m, dbh_cm) %>% 
  summary()
```

Since a primary objective of this analysis is to evaluate how [different treatments](#silv_sum) influence forest structure, we'll treat each unit as a discrete area and exclude any trees located outside of the unit boundaries from the clump classification process. This approach assumes that trees within a unit do not form functional clusters with those in adjacent areas (even though some units share a common border and we have data extending beyond the unit boundaries). This assumption prevents data from untreated control areas or differently treated areas from influencing the results when summarized by treatment type. This intentional ignorance of edge effects means that the scale of focus of this analysis is strictly on the treatment unit.

make the clump groups

```{r, eval = F, warning=FALSE, message=FALSE, echo=FALSE, include=FALSE}
c2t_output_dir <- "../data"
c2t_process_dir <- file.path(c2t_output_dir, "point_cloud_processing_delivery")
c2t_tracking_fnm <- file.path(c2t_process_dir, "processed_tracking_data.csv")
# fine-resolution CHM raster
chm_rast <- terra::rast( file.path(c2t_process_dir, "chm_0.1m.tif") )
stand_boundary <- 
   "c:/data/usfs/slash_pile_detecting/data/Dawson_Data/units/units.shp" %>% 
   # "d:/BLM_CO_SWDF_DawsonFuelsTreatment/units/units.shp" %>% 
   sf::st_read(quiet=T) %>% 
   dplyr::rename_with(tolower) %>% 
   dplyr::rename_with(stringr::str_squish) %>% 
   dplyr::rename_with(make.names) %>% 
   dplyr::rename_with(~stringr::str_replace_all(.x, "\\.{2,}", ".")) %>% 
   dplyr::rename_with(~stringr::str_remove(.x, "\\.$")) %>% 
   dplyr::rename_with(~stringr::str_replace_all(.x, "\\.", "_")) %>% 
   sf::st_set_geometry("geometry") %>% 
   sf::st_transform(terra::crs(chm_rast)) %>% 
   dplyr::mutate(
     unit_area_ha = as.numeric(sf::st_area(geometry))/10000
   )
# stand_boundary %>% dplyr::glimpse()
# where should we save the file?
treetops_fnm <- file.path(c2t_process_dir,"final_detected_tree_tops_attr.gpkg")
treetops_sf <- sf::st_read(dsn = treetops_fnm, quiet = TRUE)
# attach to treatment unit
treetops_sf <- treetops_sf %>% 
  dplyr::inner_join(
    treetops_sf %>% 
      sf::st_transform(sf::st_crs(stand_boundary)) %>% 
      sf::st_intersection(
        stand_boundary %>%
          dplyr::select(unit,block,tretmnt,unit_area_ha) %>% 
          dplyr::rename(type = tretmnt) %>% 
          dplyr::mutate(
            type =
              tolower(type) %>% 
              stringr::str_squish()
          ) %>% 
          dplyr::rename_with(.fn = ~paste0("trtmnt_", .x, recycle0 = T))
      ) %>% 
      sf::st_drop_geometry() %>% 
      dplyr::select(treeID, tidyselect::starts_with("trtmnt_")) %>% 
      dplyr::group_by(treeID) %>% 
      dplyr::slice(1) %>% 
      dplyr::ungroup()
    , by = "treeID"
  ) %>% 
  dplyr::mutate(
    is_in_trtmnt_unit = ifelse(is.na(trtmnt_type) | trtmnt_type=="control",F,T)
    , dplyr::across(c(trtmnt_type,trtmnt_block,trtmnt_unit),ordered)
  )
# sf::st_write(crowns_sf, dsn = "c:/users/georg/Downloads/crowns_sf.gpkg", append = F)
crowns_sf <- sf::st_read(dsn = "c:/users/georg/Downloads/crowns_sf.gpkg", quiet = TRUE) %>% 
  dplyr::select(-c(trtmnt_type,trtmnt_block,trtmnt_unit)) %>% 
  dplyr::inner_join(
    treetops_sf %>% 
      sf::st_drop_geometry() %>% 
      dplyr::select(treeID, trtmnt_type,trtmnt_block,trtmnt_unit)
    , by = "treeID"
  )
crowns_sf %>% dplyr::glimpse()
```

```{r, results=F}
get_clumps_temp <- function(tree_list) {
  # buffer and dissolve (arcGIS logic)
  clump_polys <- tree_list %>%
    dplyr::ungroup() %>% 
    sf::st_union() %>%
    sf::st_cast("POLYGON") %>%
    sf::st_as_sf() %>%
    dplyr::mutate(clump_id = dplyr::row_number())

  # attach to trees
  trees_with_clumps <- tree_list %>%
    # get rid of columns we'll create
    dplyr::select( -dplyr::any_of(c(
      "hey_xxxxxxxxxx"
      , "clump_id"
      , "clump_n_trees"
      , "clump_n_trees_grp"
      , "clump_dist_m"
    ))) %>%
    sf::st_join(clump_polys %>% dplyr::select(clump_id), join = sf::st_intersects, largest = T) %>%
    dplyr::group_by(clump_id) %>%
    dplyr::mutate(
      clump_n_trees = dplyr::n()
      , clump_n_trees_grp = dplyr::case_when(
        clump_n_trees == 1 ~ "Individual"
        , clump_n_trees >= 2 & clump_n_trees <= 4 ~ "2-4 trees"
        , clump_n_trees >= 5 & clump_n_trees <= 9 ~ "5-9 trees"
        , clump_n_trees >= 10 & clump_n_trees <= 15 ~ "10-15 trees"
        , clump_n_trees > 15 ~ ">15 trees"
        # , T ~ "error"
      ) %>%
      factor(
        ordered = T
        , levels = c("Individual", "2-4 trees", "5-9 trees", "10-15 trees", ">15 trees")
      )
    ) %>%
    dplyr::ungroup()
  # add back to clumps
  clump_polys <- clump_polys %>% 
    dplyr::inner_join(
      trees_with_clumps %>% 
        sf::st_drop_geometry() %>% 
        dplyr::distinct(clump_id, clump_n_trees, clump_n_trees_grp)
      , by = "clump_id"
    )
  return(list(clump_polys = clump_polys, tree_list = trees_with_clumps))
}
```


```{r grp-them-crwns, results=F}
# do it for all treatment units
crowns_clump_fnm <- file.path(c2t_process_dir, "crowns_clump.csv")
clump_polys_fnm <- file.path(c2t_process_dir, "crowns_clump_polys.gpkg")
if(
  !file.exists(crowns_clump_fnm) ||
  !file.exists(clump_polys_fnm)
){
  get_clumps_ans_temp <- 
    1:nrow(unq_units_df) %>% 
    purrr::map(
      \(x)
      crowns_sf %>% 
        dplyr::inner_join(
          unq_units_df %>% dplyr::slice(x)
          , by = dplyr::join_by(trtmnt_type,trtmnt_block,trtmnt_unit)
        ) %>% 
        get_clumps_temp()
    )
  # get trees
  crowns_sf <-
    get_clumps_ans_temp %>%
    purrr::map(
      \(x) purrr::pluck(x,"tree_list")
    ) %>%
    dplyr::bind_rows()
  # get clump polygons
  clump_polys <-
    1:nrow(unq_units_df) %>% 
    purrr::map(
      \(x)
      get_clumps_ans_temp[[x]] %>% 
        purrr::pluck("clump_polys") %>% 
        sf::st_set_geometry("geometry") %>% 
        dplyr::bind_cols(
          unq_units_df %>% dplyr::slice(x)
        )
    ) %>% 
    dplyr::bind_rows()
  
  # write
  crowns_sf %>% 
    sf::st_drop_geometry() %>% 
    dplyr::select(treeID, tidyselect::starts_with("clump_")) %>% 
    readr::write_csv(file = crowns_clump_fnm, append = F, progress = F)
  
  clump_polys %>% 
    sf::st_write(dsn = clump_polys_fnm, append = FALSE, quiet = TRUE)
}else{
  crowns_sf <- crowns_sf %>% 
    dplyr::select(-tidyselect::starts_with("clump_")) %>% 
    dplyr::inner_join(readr::read_csv(crowns_clump_fnm, progress = F), by = "treeID") %>% 
    dplyr::mutate(
      clump_n_trees_grp =  
        clump_n_trees_grp%>%
        factor(
          ordered = T
          , levels = c("Individual", "2-4 trees", "5-9 trees", "10-15 trees", ">15 trees")
        )
    )
  # clump_polys
  clump_polys <- sf::st_read(clump_polys_fnm, quiet = T) 
  clump_polys <- 
    clump_polys %>% 
    # convert to factors
    dplyr::mutate(
      dplyr::across(
        dplyr::all_of(
          names(clump_polys)[names(clump_polys) %in% names(unq_units_df)]
        )
        , ~ factor(
          .x
          , levels = levels(unq_units_df[[dplyr::cur_column()]])
          , ordered = T
        )
      )
      , dplyr::across(
        dplyr::all_of(c("clump_n_trees_grp"))
        , ~ factor(
          .x
          , levels = levels(crowns_sf[[dplyr::cur_column()]])
          , ordered = T
        )
      ) 
    )
}

```

let's look at what this clump definition did

```{r}
dta_temp <- crowns_sf %>% 
  dplyr::inner_join(
    unq_units_df %>% dplyr::slice(4)
    , by = dplyr::join_by(trtmnt_type,trtmnt_block,trtmnt_unit)
  ) %>% 
  dplyr::mutate(clump_id = factor(clump_id))
# plt
ggplot2::ggplot(data = dta_temp) + 
  ggplot2::geom_sf(mapping = ggplot2::aes(fill = clump_id), color = "gray", alpha = 0.8) +
  ggplot2::scale_fill_manual(
    values = c(viridis::turbo(n=nrow(dta_temp)/2), viridis::viridis(n=nrow(dta_temp)/2)) %>% sample()
  ) +
  ggplot2::theme_void() +
  ggplot2::theme(legend.position = "none")
```

where each clump is colored by a (nearly) distinct color

here it with clumps colored by clump group size

```{r}
ggplot2::ggplot(data = dta_temp) + 
  ggplot2::geom_sf(mapping = ggplot2::aes(fill = clump_n_trees_grp), color = "gray") +
  harrypotter::scale_fill_hp_d("always", drop = F) +
  ggplot2::labs(fill="") +
  ggplot2::theme_void() +
  ggplot2::theme(legend.position = "top", legend.direction = "horizontal") +
  ggplot2::guides(fill = ggplot2::guide_legend(override.aes = list(size = 5)))
```

and we have `clump_*` variables

```{r}
crowns_sf %>% 
  sf::st_drop_geometry() %>% 
  dplyr::select(treeID, tidyselect::starts_with("trtmnt_"), tidyselect::starts_with("clump_")) %>% 
  dplyr::glimpse()
```

```{r, warning=FALSE, message=FALSE, echo=FALSE, include=FALSE}
# crowns_sf %>% 
#   sf::st_drop_geometry() %>% 
#   dplyr::count(treeID) %>% 
#   dplyr::filter(n>1) %>% 
#   dplyr::inner_join(
#     crowns_sf %>% 
#       sf::st_drop_geometry() %>% 
#       dplyr::select(treeID, tidyselect::starts_with("trtmnt_"), tidyselect::starts_with("clump_"))
#   ) %>% 
#   dplyr::arrange(treeID) %>% 
#   View()
remove(list = ls()[grep("_temp",ls())])
gc()
```

## Identify Openings

Let's identify the gaps within a treatment unit. We'll identify openings based on their distance from the nearest tree crown where an opening is any area located at least 3 m from the nearest tree ([Churchill et al. 2016](https://scholarworks.umt.edu/ico/3)). This 3 m buffer accounts for the zone of immediate canopy influence and ensures that areas categorized as open are functionally distinct from the canopy. This threshold is flexible and can be adjusted to meet specific management or analysis objectives (e.g. flame length; wildlife corridors). 

First, we'll generate a distance raster where each cell represents the distance to the nearest tree clump group (i.e. crown) within the treatment unit boundary. Areas with a distance value of 3 m or greater are classified as openings and are defined by spatial connectedness, not a specific geometric shape. As such, they may be irregularly shaped as they are allowed to wind between clumps to follow a path where trees are distant. A single opening continues as long as the flow of open space is not entirely obstructed by surrounding tree groups. Once a gap is completely surrounded by canopy, the opening is closed off and is treated as a discrete spatial feature.


```{r}
# create function to pass a return from get_clump_summary() and get a distance raster
get_clump_dist_rast_temp <- function(x, clump_dist_m = 3){
  # get clump_dist_m
  # clump_dist_m <- min(get_clump_summary_ans$clump_dist_m, na.rm = T)
  # rasterize the clump polygons and then calculate distance between clumps as raster
  dist_rast <-
    terra::rasterize(
        x = 
          crowns_sf %>% 
          dplyr::inner_join(
            unq_units_df %>% dplyr::slice(x)
            , by = dplyr::join_by(trtmnt_type,trtmnt_block,trtmnt_unit)
          ) %>% 
          terra::vect()
        , y = 
          # crowns_sf %>% 
          # dplyr::inner_join(
          #   unq_units_df %>% dplyr::slice(x)
          #   , by = dplyr::join_by(trtmnt_type,trtmnt_block,trtmnt_unit)
          # ) %>% 
          # terra::vect() %>%
          stand_boundary %>% 
          dplyr::filter(
            unit == unq_units_df$trtmnt_unit[x]
            , block == unq_units_df$trtmnt_block[x]
            , stringr::str_squish(tolower(tretmnt)) == unq_units_df$trtmnt_type[x]
          ) %>% 
          terra::vect() %>% 
          terra::rast(res = 0.2)
      ) %>% 
      terra::distance() %>% 
      # crop it to stand extent
      terra::crop(
        stand_boundary %>% 
          dplyr::filter(
            unit == unq_units_df$trtmnt_unit[x]
            , block == unq_units_df$trtmnt_block[x]
            , stringr::str_squish(tolower(tretmnt)) == unq_units_df$trtmnt_type[x]
          ) %>% 
          terra::vect()
        , mask = T
      )
  ######### part 2
  # now create openings vector data
  openings_vect <- 
    dist_rast %>% 
      terra::classify(rcl = c(clump_dist_m,Inf), others = NA, include.lowest = T) %>% 
      terra::as.polygons(na.rm = T) %>% 
      sf::st_as_sf() %>% 
      sf::st_cast("POLYGON") %>% 
      dplyr::mutate(layer = dplyr::row_number()) %>% 
      dplyr::mutate(
        openining_area_m2 = sf::st_area(geometry) %>% as.numeric()
        , clump_dist_m = clump_dist_m
      )
  
  # return
  return(list(dist_rast = dist_rast, openings_vect = openings_vect))
}
```

let's get the distance raster and openinings vector data

```{r}
# get it
get_clump_dist_rast_ans <- 
  1:nrow(unq_units_df) %>% 
  purrr::map(
    \(x)
    get_clump_dist_rast_temp(
      x, clump_dist_m = 3
    )
  )
```

### Visualize Distance Raster

let's check out the distance raster for the same example unit we looked at above

```{r, results=F, fig.show='asis'}
# harrypotter::hp(n = length(levels(crowns_sf$clump_n_trees_grp)), option = "always") %>% scales::show_col()
# color the grps
hey_clump_pal <-
  harrypotter::hp(n = length(levels(crowns_sf$clump_n_trees_grp)), house = "always") %>%
  setNames(levels(crowns_sf$clump_n_trees_grp))
plt_dist_rast_fn <- 
  function(x, add_clumps = F, max_dist = NA){
    plt_tit <- paste0(
      unq_units_df %>% dplyr::slice(x) %>% dplyr::pull(trtmnt_type)
      , " | block "
      , unq_units_df %>% dplyr::slice(x) %>% dplyr::pull(trtmnt_block)
      , " | "
      , unq_units_df %>% dplyr::slice(x) %>% dplyr::pull(trtmnt_unit)
    )
    # plt unit
    unit_temp <- stand_boundary %>% 
      dplyr::filter(
        unit == unq_units_df$trtmnt_unit[x]
        , block == unq_units_df$trtmnt_block[x]
        , stringr::str_squish(tolower(tretmnt)) == unq_units_df$trtmnt_type[x]
      ) %>% 
      sf::st_transform(terra::crs(chm_rast))
    
    plt_stand <- ggplot2::ggplot() +
      ggplot2::geom_sf(data = unit_temp, color = "black", fill = NA) +
      ggplot2::scale_x_continuous(expand = c(0, 0)) +
      ggplot2::scale_y_continuous(expand = c(0, 0)) +
      ggplot2::labs(
        x = ""
        , y = ""
        , subtitle = plt_tit
      ) +
      ggplot2::theme_void() +
      ggplot2::theme(
        legend.position = "top" # c(0.5,1)
        , legend.direction = "horizontal"
        , legend.box = "vertical"
        , legend.margin = ggplot2::margin(2.5,0,0,0) # trbl
        , legend.text = ggplot2::element_text(size = 7)
        , legend.title = ggplot2::element_text(size = 6)
        , legend.key = ggplot2::element_rect(fill = NA, color = NA)
        , legend.key.size = unit(0.5, "cm")
        # , plot.title = ggtext::element_markdown(size = 10, hjust = 0.5)
        , plot.title = ggplot2::element_text(size = 10, hjust = 0.5, face = "bold")
        , plot.subtitle = ggplot2::element_text(size = 9, hjust = 0.5, face = "bold")
      )
    # add raster
    plt_rast <- 
      plt_stand +
      ggplot2::geom_tile(
        data = get_clump_dist_rast_ans[[x]][["dist_rast"]] %>% 
          # terra::aggregate(fact = 2, fun = "max", cores = 4) %>% 
          # terra::crop(
          #   stand_boundaries %>% 
          #     dplyr::filter(poly_name==x)
          #     # dplyr::filter(poly_name=="west")
          #   , mask = T
          # ) %>% 
          terra::as.data.frame(xy = T) %>% 
          dplyr::rename(f=3)
        , mapping = ggplot2::aes(x=x, y=y, fill = f)
      )
    if(!is.na(max_dist)){
      plt_rast <- plt_rast +
        harrypotter::scale_fill_hp(
          "mischief"
          , na.value = "transparent"
          , name = "distance to\nnearest tree (m)"
          , limits = c(0,max_dist)
          , labels = scales::comma_format(accuracy = 1)
        )
    }else{
      plt_rast <- plt_rast +
        harrypotter::scale_fill_hp(
          "mischief"
          , na.value = "transparent"
          , name = "distance to\nnearest tree (m)"
          # , limits = c(0,max_dist)
          , labels = scales::comma_format(accuracy = 1)
        )
    }
    # add clumps
    if(add_clumps){
      plt_rast <- plt_rast +
        # # openings
        # ggplot2::geom_sf(
        #   data = get_clump_dist_rast_ans[[nm]]$openings_vect
        #   , mapping = ggplot2::aes(color = openining_area_m2), fill = NA
        # ) +
        # ggplot2::scale_color_gradient(
        #   low = "gray77", high = "gray11"
        #   , labels = scales::comma_format(accuracy = 1)
        #   , name = latex2exp::TeX("opening\narea ($\\m^2$)")
        #   , limits = c(0,openining_area_m2_max)
        # ) +
        # clumps
        ggnewscale::new_scale_fill() +
        ggplot2::geom_sf(
          data = clump_polys %>% 
            dplyr::inner_join(
              unq_units_df %>% dplyr::slice(x)
              , by = dplyr::join_by(trtmnt_type,trtmnt_block,trtmnt_unit)
            ) %>% 
            dplyr::select(clump_n_trees_grp)
          , mapping = ggplot2::aes(fill = clump_n_trees_grp)
          , color = NA
        ) +
        # harrypotter::scale_fill_hp_d("always", name = "clump size", drop = F) +
        ggplot2::scale_fill_manual(values = hey_clump_pal, name = "", drop = F) +
        # tree points
        ggplot2::geom_sf(
          data = 
            treetops_sf %>% 
            dplyr::inner_join(
              unq_units_df %>% dplyr::slice(x)
              , by = dplyr::join_by(trtmnt_type,trtmnt_block,trtmnt_unit)
            )
          , color = "gray88", shape = "."
        )
        # ggplot2::guides(fill = ggplot2::guide_legend(ncol = 2))
    }
    return(plt_rast)
  }
# one unit
plt_dist_rast_fn(4, add_clumps = F)
```

```{r, include=F, eval=F}
###########!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!! terra::plot() version is fast but lacks easy customization (esp legend placement :\ )
# harrypotter::hp(n = length(levels(crowns_sf$clump_n_trees_grp)), option = "always") %>% scales::show_col()
plt_dist_rast_fn <- 
  function(x, add_crowns = F, add_clumps = F, add_treetops = F){
    plt <- paste0(
      unq_units_df %>% dplyr::slice(x) %>% dplyr::pull(trtmnt_type)
      , " | block "
      , unq_units_df %>% dplyr::slice(x) %>% dplyr::pull(trtmnt_block)
      , " | "
      , unq_units_df %>% dplyr::slice(x) %>% dplyr::pull(trtmnt_unit)
    )
    
    terra::plot(
      get_clump_dist_rast_ans[[x]][["dist_rast"]]
      , col = harrypotter::hp(n=100, option = "mischief")
      # , mar = c(b,l,t,r)
      , mar = c(0.2,0.2,2.2,3.5)
      , axes = F
      , plg = list(
        title = "distance to\nnearest tree (m)"
        , cex = 0.8, title.cex = 0.5
        , x = "right", shrink = 0.6
      )
      , main = plt
    ) 
    
    # add stand
    unit_temp <- stand_boundary %>% 
      dplyr::filter(
        unit == unq_units_df$trtmnt_unit[x]
        , block == unq_units_df$trtmnt_block[x]
        , stringr::str_squish(tolower(tretmnt)) == unq_units_df$trtmnt_type[x]
      ) %>% 
      sf::st_transform(terra::crs(get_clump_dist_rast_ans[[x]][["dist_rast"]])) %>% 
      terra::vect()
    terra::plot(
      unit_temp
      , border = "black", col = NA
       , add = T
    )
    
    if(add_crowns){
      terra::plot(
        crowns_sf %>% 
          dplyr::inner_join(
            unq_units_df %>% dplyr::slice(x)
            , by = dplyr::join_by(trtmnt_type,trtmnt_block,trtmnt_unit)
          ) %>% 
          terra::vect()
        , add = T, border = "cyan", col = NA, lwd = 0.7
      )
    }else if(add_clumps){
      # crowns_sf %>% 
      #   dplyr::inner_join(
      #     unq_units_df %>% dplyr::slice(x)
      #     , by = dplyr::join_by(trtmnt_type,trtmnt_block,trtmnt_unit)
      #   ) %>% 
      #   terra::vect() %>% 
      #   terra::plot(add = T, border = "cyan", col = NA, lwd = 0.7)
      clump_polys %>% 
        dplyr::inner_join(
          unq_units_df %>% dplyr::slice(x)
          , by = dplyr::join_by(trtmnt_type,trtmnt_block,trtmnt_unit)
        ) %>% 
        dplyr::select(clump_n_trees_grp) %>% 
        terra::vect() %>% 
        terra::plot(
          "clump_n_trees_grp"
          , col = rev(harrypotter::hp(n = length(levels(crowns_sf$clump_n_trees_grp)), option = "always"))
          , border = NA
          , axes = F
          , add = T
          , plg = list(
            x = "topright"
          )
        )
    }
    
    if(add_treetops){
      terra::plot(
        treetops_sf %>% 
          dplyr::inner_join(
            unq_units_df %>% dplyr::slice(x)
            , by = dplyr::join_by(trtmnt_type,trtmnt_block,trtmnt_unit)
          ) %>% 
          terra::vect()
        , add = T, col = "gray55", cex = 0.2
      )
    }
  }
# 1:nrow(unq_units_df) %>% 
#   .[3] %>% 
#   purrr::walk(plt_dist_rast_fn)
plt_dist_rast_fn(4)
```

now we'll overlay the tree clump groups on the distance raster for all units

```{r,eval=F,purl=FALSE, results=F, fig.show='asis', fig.height=9.5, fig.width=9.5}
# mapping
plt_blk_temp <- function(blk, add = F, save = F) {
  # which idx
  idx_temp <- 
    unq_units_df %>% 
    dplyr::ungroup() %>% 
    dplyr::mutate(rownum = dplyr::row_number()) %>% 
    dplyr::filter(trtmnt_block==blk) %>% 
    dplyr::pull(rownum)
  # max dist
  max_d <- 
    purrr::map(get_clump_dist_rast_ans[idx_temp], "dist_rast") %>% 
    purrr::map(
      \(r)
      terra::minmax(r) %>% 
        .[2] %>% 
        ceiling()  
    ) %>% 
    purrr::compact() %>% 
    unlist() %>% 
    max(na.rm = T)
  # plts
  plts_temp <- 
    idx_temp %>% 
    purrr::map(\(x) plt_dist_rast_fn(x, add_clumps = add, max_dist = max_d))
  # patchwork
  p_temp <- 
    patchwork::wrap_plots(
      plts_temp
      , ncol = 2
      , guides = "collect"
    ) &
    ggplot2::theme(legend.position = "bottom")
  
  # save 
  if(save){
    ggplot2::ggsave(
      filename = file.path("../data", paste0("plt_dist_block_", blk, ".png"))
      , plot = p_temp
      , width = 10
      , height = 10
      , dpi = 300
    )
  }else{
    p_temp
  }
}
# map over blks
unique(unq_units_df$trtmnt_block) %>% 
  # .[1:2] %>%
  purrr::map(\(x) plt_blk_temp(blk=x, add = T, save = F))
```

```{r, warning=FALSE, message=FALSE, echo=FALSE, include=FALSE}
remove(list = ls()[grep("_temp",ls())])
gc()
```

### Visualize Openings

let's check out the openings for the same example unit we looked at above

```{r, results=F, fig.show='asis'}
# harrypotter::hp(n = length(levels(crowns_sf$clump_n_trees_grp)), option = "always") %>% scales::show_col()
plt_openings_vect_fn <- 
  function(x, add_clumps = F, max_area = NA){
    plt_tit <- paste0(
      unq_units_df %>% dplyr::slice(x) %>% dplyr::pull(trtmnt_type)
      , " | block "
      , unq_units_df %>% dplyr::slice(x) %>% dplyr::pull(trtmnt_block)
      , " | "
      , unq_units_df %>% dplyr::slice(x) %>% dplyr::pull(trtmnt_unit)
    )
    # plt unit
    unit_temp <- stand_boundary %>% 
      dplyr::filter(
        unit == unq_units_df$trtmnt_unit[x]
        , block == unq_units_df$trtmnt_block[x]
        , stringr::str_squish(tolower(tretmnt)) == unq_units_df$trtmnt_type[x]
      ) %>% 
      sf::st_transform(terra::crs(chm_rast))
    
    plt_stand <- ggplot2::ggplot() +
      ggplot2::geom_sf(data = unit_temp, color = "black", fill = NA) +
      ggplot2::scale_x_continuous(expand = c(0, 0)) +
      ggplot2::scale_y_continuous(expand = c(0, 0)) +
      ggplot2::labs(
        x = ""
        , y = ""
        , subtitle = plt_tit
      ) +
      ggplot2::theme_void() +
      ggplot2::theme(
        legend.position = "top" # c(0.5,1)
        , legend.direction = "horizontal"
        , legend.box = "vertical"
        , legend.margin = ggplot2::margin(2.5,0,0,0) # trbl
        , legend.text = ggplot2::element_text(size = 7)
        , legend.title = ggplot2::element_text(size = 6)
        , legend.key = ggplot2::element_rect(fill = NA, color = NA)
        , legend.key.size = unit(0.5, "cm")
        # , plot.title = ggtext::element_markdown(size = 10, hjust = 0.5)
        , plot.title = ggplot2::element_text(size = 10, hjust = 0.5, face = "bold")
        , plot.subtitle = ggplot2::element_text(size = 9, hjust = 0.5, face = "bold")
      )
    # add open
    plt_open <- 
      plt_stand +
      # openings
      ggplot2::geom_sf(
        data = get_clump_dist_rast_ans[[x]][["openings_vect"]]
        , mapping = ggplot2::aes(fill = openining_area_m2), color = NA
      )
      # ggplot2::guides(
      #   fill = ggplot2::guide_colorbar(
      #     label.theme = ggplot2::element_text(
      #       angle = 45, hjust = 1.1, vjust = 1.4, size = 8
      #     )
      #   )
      # )
      
    if(!is.na(max_area)){
      plt_open <- plt_open +
        ggplot2::scale_fill_gradient(
          low = "gray77", high = "gray11"
          , labels = scales::comma_format(accuracy = 1)
          , name = "opening\narea (m²)" #latex2exp::TeX("opening\narea ($\\m^2$)")
          , limits = c(0,max_area)
          , guide = ggplot2::guide_colorbar(
            label.theme = ggplot2::element_text(
              angle = 55
              , hjust = 1
              , vjust = 1.3
            )
          )
        )
    }else{
      plt_open <- plt_open +
        ggplot2::scale_fill_gradient(
          low = "gray77", high = "gray11"
          , labels = scales::comma_format(accuracy = 1)
          , name = "opening\narea (m²)" #latex2exp::TeX("opening\narea ($\\m^2$)")
          # , limits = c(0,max_area)
          , guide = ggplot2::guide_colorbar(
            label.theme = ggplot2::element_text(
              angle = 55
              , hjust = 1
              , vjust = 1.3
            )
          )
        )
    }
    # add clumps
    if(add_clumps){
      plt_open <- plt_open +
        # clumps
        ggnewscale::new_scale_fill() +
        ggplot2::geom_sf(
          data = clump_polys %>% 
            dplyr::inner_join(
              unq_units_df %>% dplyr::slice(x)
              , by = dplyr::join_by(trtmnt_type,trtmnt_block,trtmnt_unit)
            ) %>% 
            dplyr::select(clump_n_trees_grp)
          , mapping = ggplot2::aes(fill = clump_n_trees_grp)
          , color = NA
        ) +
        # harrypotter::scale_fill_hp_d("always", name = "clump size", drop = F) +
        ggplot2::scale_fill_manual(values = hey_clump_pal, name = "", drop = F) +
        # tree points
        ggplot2::geom_sf(
          data = 
            treetops_sf %>% 
            dplyr::inner_join(
              unq_units_df %>% dplyr::slice(x)
              , by = dplyr::join_by(trtmnt_type,trtmnt_block,trtmnt_unit)
            )
          , color = "gray88", shape = "."
        )
        # ggplot2::guides(fill = ggplot2::guide_legend(ncol = 2))
    }
    return(plt_open)
  }
# one unit
plt_openings_vect_fn(6, add_clumps = F)
```

now we'll overlay the tree clump groups on the openings data for all units

```{r,eval=F,purl=FALSE, results=F, fig.show='asis', fig.height=9.5, fig.width=9.5}
# mapping
plt_blk_temp <- function(blk, add = F, save = F) {
  # which idx
  idx_temp <- 
    unq_units_df %>% 
    dplyr::ungroup() %>% 
    dplyr::mutate(rownum = dplyr::row_number()) %>% 
    dplyr::filter(trtmnt_block==blk) %>% 
    dplyr::pull(rownum)
  # max dist
  max_d <-
    purrr::map(get_clump_dist_rast_ans[idx_temp], "openings_vect") %>%
    dplyr::bind_rows() %>% 
    dplyr::pull(openining_area_m2) %>% 
    max(na.rm = T) %>% 
    ceiling()
  # plts
  plts_temp <- 
    idx_temp %>% 
    purrr::map(\(x) plt_openings_vect_fn(x, add_clumps = add, max_area = max_d))
  # patchwork
  p_temp <- 
    patchwork::wrap_plots(
      plts_temp
      , ncol = 2
      , guides = "collect"
    ) &
    ggplot2::theme(legend.position = "bottom")
  
  # save 
  if(save){
    ggplot2::ggsave(
      filename = file.path("../data", paste0("plt_open_block_", blk, ".png"))
      , plot = p_temp
      , width = 10
      , height = 10
      , dpi = 300
    )
  }else{
    p_temp
  }
}
# map over blks
unique(unq_units_df$trtmnt_block) %>% 
  # .[1:2] %>%
  purrr::map(\(x) plt_blk_temp(blk=x, add = T, save = T))
```

```{r, warning=FALSE, message=FALSE, echo=FALSE, include=FALSE}
remove(list = ls()[grep("_temp",ls())])
gc()
```

## Tree Clump Summary

let's get some summary stats of the tree clump groupings by treatment type

```{r}
# create function to pass a return from st_clump_points/get_tree_clumps and create clump polygons with summary stats
get_clump_summary <- function(get_tree_clumps_ans){
  # get clump_dist_m
  clump_dist_m <- min(get_tree_clumps_ans$clump_dist_m, na.rm = T)
  # create clump polys and summary
  clump_polys_temp <- 
    get_tree_clumps_ans %>% 
      dplyr::ungroup() %>% 
      # rename the active geometry column to "geometry"
      sf::st_set_geometry("geometry") %>% 
      dplyr::mutate(geometry = sf::st_buffer(geometry, dist = clump_dist_m/2)) %>% 
      dplyr::group_by(clump_id, clump_n_trees_grp, clump_n_trees_lb, clump_n_trees_ub) %>%
      dplyr::summarise(
        # union buffered tree points
        geometry = sf::st_union(geometry)
        # summary metrics
        , n_trees = dplyr::n_distinct(treeID)
        , mean_dbh_cm = mean(dbh_cm, na.rm = T)
        , min_dbh_cm = min(dbh_cm, na.rm = T)
        , max_dbh_cm = max(dbh_cm, na.rm = T)
        , mean_tree_height_m = mean(tree_height_m, na.rm = T)
        , min_tree_height_m = min(tree_height_m, na.rm = T)
        , max_tree_height_m = max(tree_height_m, na.rm = T)
        , loreys_height_m = sum(basal_area_m2*tree_height_m, na.rm = T) / sum(basal_area_m2, na.rm = T)
        , basal_area_m2 = sum(basal_area_m2, na.rm = T)
        , sum_dbh_cm_sq = sum(dbh_cm^2, na.rm = T)
        , .groups = "drop_last"
      ) %>%
      dplyr::ungroup() %>% 
      sf::st_make_valid() %>% 
      dplyr::mutate(
        clump_area_ha = sf::st_area(geometry) %>% as.numeric() %>% `/`(10000)
        , trees_per_ha = (n_trees/clump_area_ha)
        , basal_area_m2_per_ha = (basal_area_m2/clump_area_ha)
        , pct_stand_basal_area = basal_area_m2/sum(basal_area_m2)
        , pct_stand_n_trees = n_trees/sum(n_trees)
        , qmd_cm = sqrt(sum_dbh_cm_sq/n_trees)
      ) %>%
      dplyr::select(-c(sum_dbh_cm_sq)) %>% 
    # convert to imperial units
      calc_imperial_units_fn() %>% 
      dplyr::mutate(clump_dist_m = clump_dist_m)
  # calculate distance between clumps
  clump_polys_temp <- clump_polys_temp %>% 
    dplyr::mutate(
      nearest = sf::st_nearest_feature(clump_polys_temp)
      , distance_nearest_clump_m = sf::st_distance(
          clump_polys_temp
          , clump_polys_temp[nearest,]
          , by_element=TRUE
        ) %>% 
        as.numeric()
    ) %>% 
    dplyr::select(-c(nearest))
  # return
  return(clump_polys_temp)
}

# create a function to summarize by number of tree clump grouping
get_clump_n_trees_grp_summary = function(get_tree_clumps_ans, get_clump_summary_ans, stand_sf){
  # get area of harvest unit
  #...will use this area in the area calculations such that...
  #...TPA = trees in a certain group size across the whole stand area
  harvest_area_ha <- stand_sf %>%
    sf::st_union() %>% 
    sf::st_as_sf() %>% 
    dplyr::mutate(harvest_area_ha = sf::st_area(.) %>% as.numeric() %>% `/`(10000)) %>% 
    dplyr::pull(harvest_area_ha) %>% 
    .[1]
  # collapse and calculate silv metrics
  dta <- get_tree_clumps_ans %>% 
    sf::st_drop_geometry() %>% 
    dplyr::mutate(stand_area_ha = harvest_area_ha) %>% 
    dplyr::group_by(stand_area_ha,clump_n_trees_grp, clump_n_trees_lb, clump_n_trees_ub) %>%
    dplyr::summarise(
      # summary metrics
      n_trees = dplyr::n_distinct(treeID)
      , mean_dbh_cm = mean(dbh_cm, na.rm = T)
      , mean_tree_height_m = mean(tree_height_m, na.rm = T)
      , loreys_height_m = sum(basal_area_m2*tree_height_m, na.rm = T) / sum(basal_area_m2, na.rm = T)
      , basal_area_m2 = sum(basal_area_m2, na.rm = T)
      , sum_dbh_cm_sq = sum(dbh_cm^2, na.rm = T)
      , .groups = "drop_last"
    ) %>%
    dplyr::ungroup() %>% 
    # attach clump area
    dplyr::left_join(
      get_clump_summary_ans %>% 
        sf::st_drop_geometry() %>% 
        dplyr::group_by(clump_n_trees_grp) %>%
        dplyr::summarise(
          clump_area_ha = sum(clump_area_ha)
          , stand_n_clumps = dplyr::n()
          , .groups = "drop_last"
        ) %>% 
        dplyr::ungroup()
      , by = dplyr::join_by(clump_n_trees_grp)
    ) %>% 
    dplyr::mutate(
      trees_per_ha = (n_trees/stand_area_ha) # (n_trees/clump_area_ha) ... this was not right
      , basal_area_m2_per_ha = (basal_area_m2/stand_area_ha) # (basal_area_m2/clump_area_ha) ... this was not right
      , qmd_cm = sqrt(sum_dbh_cm_sq/n_trees)
      # stand calcs
      , stand_trees_per_ha = sum(n_trees)/stand_area_ha
      , stand_basal_area_m2 = sum(basal_area_m2)
      , stand_basal_area_m2_per_ha = sum(basal_area_m2)/stand_area_ha
      , pct_stand_basal_area = basal_area_m2/stand_basal_area_m2
      , pct_stand_n_trees = n_trees/sum(n_trees)
      , stand_qmd_cm = sqrt(sum(get_tree_clumps_ans$dbh_cm^2, na.rm = T)/sum(n_trees))
    ) %>%
    dplyr::select(-c(sum_dbh_cm_sq)) %>% 
    # convert to imperial units
    calc_imperial_units_fn()
  # return
  return(dta)
}


# call it
clump_n_trees_grp_summary2 <- 
  stand_boundaries$poly_name %>% 
  unique() %>% 
  purrr::set_names() %>% 
  purrr::map(
    \(x)
    get_clump_n_trees_grp_summary(
      get_tree_clumps_ans = treetops_clustered_new[[x]]
      , get_clump_summary_ans = clump_polys_new[[x]]
      , stand_sf = stand_boundaries %>% dplyr::filter(poly_name==x)
    ) 
  )
# table it
clump_n_trees_grp_summary2 %>% 
  purrr::imap(
   \(x,nm)  
    x %>% 
      dplyr::select(
        clump_n_trees_grp, n_trees
        , mean_dbh_in
        , qmd_in
        , mean_tree_height_ft
        , loreys_height_ft
        , trees_per_ac
        , basal_area_ft2_per_ac, pct_stand_basal_area, pct_stand_n_trees
      ) %>% 
      dplyr::mutate(
        dplyr::across(
          .cols = c(pct_stand_basal_area, pct_stand_n_trees) 
          , .fns = ~ scales::percent(.x, accuracy = 1)
        )
        , stand = nm
      )
  ) %>% 
  dplyr::bind_rows() %>% 
  dplyr::relocate(stand) %>% 
  kableExtra::kbl(
    digits = 1
    , escape = F
    , caption = paste0("Overstory tree clump summary: post-treatment")
    , col.names = c(
      "stand"
      , "", "trees"
      , "mean<br>DBH (in)"
      , "QMD (in)"
      , "mean<br>Ht. (ft)"
      , "Loreys<br>Ht. (ft)"
      , "TPA"
      , "BA<br>ft<sup>2</sup> ac<sup>-1</sup>"
      , "Pct<br>stand BA"
      , "Pct<br>stand trees"
      )
  ) %>% 
  kableExtra::kable_styling() %>% 
  kableExtra::collapse_rows(columns = 1, valign = "top")
```


## Openings Summary
